---
title: "GLOBIO Land-Use Reclassification to Broad Classes"
output: html_notebook
---
## Overview
This notebook takes **GLOBIO land-use GeoTIFF rasters in in zip files** and reclassifies their original land-use codes into a smaller set of **broad land-use classes**.


## Clean environment and load libraries
```{r}
rm(list = ls())
gc()

library(terra)
library(tidyverse)
```

## Define Terra options and temporary directories
Make sure to set up a local temporary directory for processing large raster files. It should
have enough space to handle the unzipped and intermediate raster files.

```{r}
local_temp <- "C:/temp/globio_reclass"
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk = TRUE,
  tempdir = local_temp,
  progress = 1
)
  
input_dat <- "S:\\Emmanuel_OcegueraConchas\\BES-SIM 2\\GLOBIO\\zip_original\\zip_original"
output <- "S:\\Emmanuel_OcegueraConchas\\BES-SIM 2\\GLOBIO\\aggregated_broadclasses"
dir.create(output, showWarnings = FALSE, recursive = TRUE)
```


## List all the zip files
```{r}
zip_files <- list.files(
  input_dat,
  pattern = "\\.zip$",
  full.names = TRUE,
  recursive = TRUE
)

length(zip_files)
cat(paste0(zip_files, collapse = "\n"), "\n" )

```

## Build raster metadata table
create a table (`tif_tbl`) that contains metadata for each raster file,
including the scenario, SSP, parameter, and year. This table will help us manage
the reclassification process by providing a structured overview of the rasters to be processed.
```{r}
tif_tbl <- tibble(
  path = tif,
  file = basename(tif)
) %>%
  mutate(
    scenario = case_when(
      str_detect(path, "NaC") ~ "NaC",
      str_detect(path, "NfN") ~ "NfN",
      str_detect(path, "NfS") ~ "NfS",
      TRUE ~ NA_character_
    ),
    ssp   = str_extract(path, "ssp\\d+"),
    param = str_remove_all(str_extract(path, "_\\d{2}_"), "_"),
    year  = as.integer(str_extract(file, "\\d{4}"))
  )
cat("Table of rasters to process:\n")
count(tif_tbl, scenario, ssp, param, year)
```

---

## Broad land-use reclassification table
Define reclassification table (`broad_lut`) that maps the original GLOBIO land-use codes
to broader land-use categories. This table will be used to reclassify the raster values.
This table has been provided by GLOBIO team.

```{r}
broad_lut <- tribble(
  ~value, ~broad_id, ~broad_name,
  1, 1, "Urban",
  2, 1, "Urban",
  190, 1, "Urban",

  10, 10, "Cropland",
  11, 10, "Cropland",
  12, 10, "Cropland",

  13, 13, "Potential secondary vegetation",
  14, 13, "Potential secondary vegetation",

  15, 15, "Agroforestry",

  21, 21, "Pasture",

  22, 22, "Rangeland",
  
  30, 30, "Forestry",

  40, 40, "Secondary vegetation",

  50, 50, "Unmanaged forest",
  60, 50, "Unmanaged forest",
  61, 50, "Unmanaged forest",
  62, 50, "Unmanaged forest",
  70, 50, "Unmanaged forest",
  71, 50, "Unmanaged forest",
  72, 50, "Unmanaged forest",
  80, 50, "Unmanaged forest",
  81, 50, "Unmanaged forest",
  82, 50, "Unmanaged forest",
  90, 50, "Unmanaged forest",
  100, 50, "Unmanaged forest",
  110, 50, "Unmanaged forest",

  160, 50, "Unmanaged forest",
  170, 50, "Unmanaged forest",
  180, 50, "Unmanaged forest",

  120, 120, "Shrubland/Grassland",
  121, 120, "Shrubland/Grassland",
  122, 120, "Shrubland/Grassland",
  130, 120, "Shrubland/Grassland",
  150, 150, "Shrubland/Grassland",
  151, 150, "Shrubland/Grassland",
  152, 150, "Shrubland/Grassland",
  153, 150, "Shrubland/Grassland",
 

  140, 140, "other",

  200, 200, "Bare_areas",
  201, 200, "Bare_areas",
  202, 200, "Bare_areas",

  210, 210, "Water_bodies",

  220, 220, "Permanent_snow_ice"
)
# Count the totla number of the unique original land-use values
cat("Number of unique original land-use values:", n_distinct(broad_lut$value), "\n")
cat("Broad land-use reclassification table:\n")
broad_lut
```


## Reclassification matrix
Build up the reclassification matrix (`rcl`) from the `broad_lut` table.

```{r}
rcl <- as.matrix(broad_lut[, c("value", "broad_id")])
```

## List all the GeoTIFF files inside the zip files and reclassify them
Process each zip file, extract the GeoTIFF files, and reclassify them using the
broad land-use classes defined in the reclassification matrix.

```{r}
for (z in zip_files) {

  # Make a unique temp folder for this zip
  z_base <- tools::file_path_sans_ext(basename(z))
  z_tmp  <- file.path(local_temp, paste0(z_base, "_", Sys.getpid()))
  dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)

  # Process only the z_base with 1.9 in the name
  if (!grepl("1.9", z_base)) {
    cat("Skipping ZIP (not _19_):", z, "\n")
    next
  }

  cat("\n=== Unzipping ZIP to local temp ===\n", z_tmp, "\n")

  # Unzip to local temp
  unzip(z, exdir = z_tmp)

  # List all tifs inside extracted contents
  tif <- list.files(
    z_tmp,
    pattern = "\\.tif$",
    full.names = TRUE,
    recursive = TRUE
  )

  # Process each tif file
  for (p in tif) {
    cat("Processing file:", p, "\n")
    p_norm <- gsub("\\\\", "/", p)

    scenario <- if (grepl("NaC", p_norm)) "NaC" else if (grepl("NfN", p_norm)) "NfN" else if (grepl("NfS", p_norm)) "NfS" else NA

    year <- as.integer(str_extract(basename(p_norm), "\\d{4}"))
    # ssp  <- str_extract(p_norm, "ssp\\d+")

    if (is.na(scenario) || is.na(year)) {
      warning("Skipping: ", p)
      next
    }

    cat("Reclassifying:", scenario, year, "\n")

    r  <- rast(p)
    r2 <- classify(r, rcl, others = NA)

    out_dir <- file.path(output, paste0(scenario, "_19"))
    dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

    out_file <- paste0(
      "Globio4_landuse_10sec_",
      year,
      "_World_crop_int_withPattern_broad.tif"
    )

    writeRaster(
      r2,
      file.path(out_dir, out_file),
      datatype = "INT1U",
      NAflag = 255,
      overwrite = TRUE,
      wopt = list(gdal = c("COMPRESS=DEFLATE"))
    )
  }
  # Clean up temp folder
  unlink(z_tmp)
}

```

## Reclassify historical 2020 raster separatelly 
```{r}
in_dir <- "S:\\Emmanuel_OcegueraConchas\\BES-SIM 2\\GLOBIO\\zip_original\\zip_original\\Historical_2000-2020.zip"
out_dir_hist <- "S:\\Emmanuel_OcegueraConchas\\BES-SIM 2\\GLOBIO\\aggregated_broadclasses\\Historical_2020"
dir.create(out_dir_hist, showWarnings = FALSE, recursive = TRUE)

# Create a temp folder for unzipping
z_base <- tools::file_path_sans_ext(basename(in_dir))
z_tmp  <- file.path(local_temp, paste0(z_base, "_", Sys.getpid()))
dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)

unzip(in_dir, exdir = z_tmp)

r_hist  <- rast(file.path(z_tmp, "Globio4_landuse_10sec_2020_World_crop_int.tif"))

# check the frequency of original land-use values
r2_hist <- classify(r_hist, rcl, others = NA)

writeRaster(
  r2_hist,
  file.path(out_dir_hist, "Globio4_landuse_10sec_2020_World_crop_int_broad.tif"),
  datatype = "INT1U",
  NAflag = 255,
  overwrite = TRUE,
  wopt = list(gdal = c("COMPRESS=DEFLATE"))
)
```

