---
title: "GLOBIO Land-Use - Reclassification and EBV Cube generation"
author: "Emmanuel Oceguera-Conchas, Andres Marmol Guijarro, Magali Weissgerber"
date: 23-02-2026
output: html_document
---

## 1. Introduction

This workflow processes high-resolution GLOBIO land-use data and converts it into an EBV Cube NetCDF file (dataset ID 98).

The GLOBIO data come at a resolution of 0.002777° (approx. 300 m at the equator) and contain numerous detailed land-use categories. For compatibility with the EBV Cube structure, these are first grouped into a smaller set of harmonized broad land-use classes.

The workflow comprises three core steps:

-   *Reclassification:* Original GLOBIO classes are mapped to harmonized broad land-use classes using a lookup table.

-   *Binarization:* Each broad land-use class is converted into a binary raster (1 for presence, 0 for absence).

-   *EBV Cube NetCDF file generation:* The binary rasters are structured into an EBV Cube, organizing the data by scenario, metric, time step, and entity (the broad land-use class).

In the notebook, these steps span five sections: preparing the working environment (loading `terra`, `ebvcube`, `tidyverse`, `future.apply` and setting paths), reclassifying the rasters, creating the binary entity rasters, building the EBV Cube structure, and populating it. The result is a high-resolution product that a subsequent workflow aggregates to 0.5° resolution.

### 1.1 GLOBIO scenarios and time steps

The workflow processes four GLOBIO scenarios across four years each. The 2020 historical year is shared across all scenarios, so differences in projections reflect only future changes.

|                    |                  |                        |
|--------------------|------------------|------------------------|
| **Scenario**       | **Abbreviation** | **Available Years**    |
| SSP2-Baseline      | Baseline         | 2020, 2030, 2050, 2070 |
| Nature as Culture  | NaC              | 2020, 2030, 2050, 2070 |
| Nature for Society | NfS              | 2020, 2030, 2050, 2070 |
| Nature for Nature  | NfN              | 2020, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   The GLOBIO ZIP archives and raster files, in the expected folder structure
-   `GLOBIO_categories.csv`, mapping original GLOBIO classes to broad classes
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   Enough temporary disk space for raster processing
-   A working `ebvcube` install, plus `terra`, `tidyverse`, and optionally `future.apply`

**Note**

One decision before knitting: binarization can run sequentially ([Section 9.2](#sec-9.2-binarize-rasters-without-parallel-processing)) or in parallel ([Section 9.3](#sec-9.3-binarize-rasters-with-parallel-processing)) - these are mutually exclusive. Parallel runs by default. For sequentially, set 9.2 to `eval = TRUE` and 9.3 to `eval = FALSE`.

```{r}
# note: Set cache = FALSE when re-running after changes to input data or paths
knitr::opts_chunk$set(cache = TRUE)
```

## 2. Load libraries

Here we load the packages used throughout the workflow: `terra` for raster processing (reading, reclassifying, binarizing, writing), `ebvcube` for creating and filling the EBV Cube NetCDF file, `tidyverse` for handling lookup tables and data preparation, and `future.apply` for optional parallel binarization later on.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")
# install.packages("tidyverse")
# install.packages("future.apply")

library(terra)
library(ebvcube)
library(tidyverse)
library(future.apply)
```

## 3. Configure temporary raster storage

Reclassification and binarization can write a lot of temporary data to disk, so point `terra` at a drive with enough free space.

```{r}

local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths

Input: the folder of GLOBIO ZIP archives.\
Output: the folder for reclassified broad-class rasters.

```{r}

input_root <- "../raw/10arcsecs"
output_root <- "../preprocessed/10arcsecs" 
dir.create(output_root, showWarnings = FALSE, recursive = TRUE)


cat("Input root:", input_root, "\n")
cat("Output root:", output_root, "\n")

```

## 5. Define the main workflow settings

Set the dataset ID, scenarios, and years to process.

```{r}

id <-  98
years  <- c(2020, 2030, 2050, 2070)
scenarios <- c("NfN", "NaC", "NfS", "Baseline")


cat("ID:", id, "\n")
cat("Years:", paste(years, collapse = ", "), "\n")
cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
```

## 6. Load the reclassification table

Load the lookup table that maps original GLOBIO land-use values to the broad harmonized classes used in binarization and EBV Cube creation.

```{r}

table_path <- "../GLOBIO_categories.csv"
globio_lut <- read.csv(table_path)

cat("Unique original codes:", n_distinct(globio_lut$values), "\n")
cat("Unique broad classes:", n_distinct(globio_lut$broad_id), "\n")
cat("Broad classes:\n", paste(unique(globio_lut$broader_category), collapse = ", "), "\n")
head(globio_lut)
```

### 6.1 Build a reclassification matrix

`terra::classify()` expects a two-column matrix: original values in the first column, new values in the second.

```{r}
rcl <- as.matrix(globio_lut[, c("values", "broad_id")])
head(rcl)
```

## 7. Identify the ZIP files to process

List all ZIP archives and keep only the ones needed. The filter excludes the historical archive (handled separately in [Section 8.2](#sec-8.2-reclassify-historical-2020-raster)) and keeps only the selected scenarios - matching the `1.9` version tag in the BES-SIM2 filenames (e.g. `GLOBIO4_landuse_10sec_2050_NfN_v1.9.zip`) plus the differently-named `Baseline.zip`.

```{r}
zip_files <- list.files(
  input_root,
  pattern = "\\.zip$",
  full.names = TRUE,
  recursive = TRUE
)

zip_files <- zip_files[!grepl("Historical_2000-2020", basename(zip_files))]
zip_files <- zip_files[
  grepl("1\\.9", basename(zip_files))  | grepl("Baseline\\.zip$", basename(zip_files))
]

cat("Total ZIP files to process:", length(zip_files), "\n")
head(zip_files, 10)
```

## 8. Reclassification

Convert the original GLOBIO land-use category values into broad land-use class IDs via the lookup table. The output store harmonized class IDs instead of the original detailed categories. This exercise is for harmonization purpose.

### 8.1 Reclassify scenario rasters

For each ZIP archive: extract it, find the `.tif` files, parse scenario and year from the path, keep only the selected years, reclassify, and write to disk.

Output settings\
-data type `INT1S`\
-NoData `-1`\
-compression `DEFLATE`

keep file sizes down and NoData handling consistent for binarization.

```{r}

# Check time
start_time <- Sys.time()

for (z in zip_files) {
  
  z_base <- tools::file_path_sans_ext(basename(z))
  
  z_tmp  <- file.path(
    local_tmp, 
    paste0(z_base, "_", Sys.getpid())
    )
  
  dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)

  cat("\n=== Unzipping ===\n", z, "\n-> ", z_tmp, "\n")
  unzip(z, exdir = z_tmp)

  tif_files <- list.files(
    z_tmp,
    pattern = "\\.tif$",
    full.names = TRUE,
    recursive = TRUE
  )

  for (p in tif_files) {

    yr <- as.integer(stringr::str_extract(basename(p), "\\d{4}"))
    
    if (is.na(yr) || !(yr %in% years)) next

    p_norm <- gsub("\\\\", "/", p)

    scenario <- NA_character_
    if (grepl("NaC", p_norm)) scenario  <- "NaC"
    if (grepl("NfN", p_norm)) scenario  <- "NfN"
    if (grepl("NfS", p_norm)) scenario  <- "NfS"
    if (grepl("Baseline", p_norm)) scenario  <- "Baseline"

    if (is.na(scenario)) {
      warning("Could not parse scenario, skipping: ", p)
      next
    }

    cat("Reclassifying:", scenario, "year", yr, "\n")

    r  <- rast(p)
    r2 <- classify(r, rcl, others = NA)

    out_dir <- file.path(
      output_root, 
      scenario
      )
    
    dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

    out_file <- file.path(
      out_dir, paste0("Globio4_landuse_10sec_", yr, 
      "_World_crop_int_withPattern_broad.tif")
      )

    writeRaster(
      r2,
      out_file,
      datatype  = "INT1S",
      NAflag    = -1,
      overwrite = TRUE,
      gdal = c("COMPRESS=DEFLATE")
    )

    rm(r, r2)
    gc()
    terra::tmpFiles(remove = TRUE)
  }

  unlink(z_tmp, recursive = TRUE)
}

total_time <- Sys.time() - start_time
cat("\nTotal processing time:", total_time, "\n")

```

### 8.2 Reclassify historical 2020 raster {#sec-8.2-reclassify-historical-2020-raster}

The 2020 layer lives in its own ZIP and serves as the shared baseline for all future scenarios, so it is processed separately.

```{r}

hist_zip <- file.path(
  input_root, 
  "Historical_2000-2020.zip"
  )

hist_out_dir <- file.path(
  output_root,
  "Historical_2020"
  )

dir.create(hist_out_dir, showWarnings = FALSE, recursive = TRUE)

z_base <- tools::file_path_sans_ext(basename(hist_zip))

z_tmp  <- file.path(
  local_tmp, 
  paste0(z_base, "_", Sys.getpid())
  )

dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)
unzip(hist_zip, exdir = z_tmp)

src <- list.files(
  z_tmp,
  pattern = "Globio4_landuse_10sec_2020_.*\\.tif$",
  full.names = TRUE,
  recursive = TRUE
)

stopifnot(length(src) == 1)

r_hist  <- rast(src)
r2_hist <- classify(r_hist, rcl, others = NA)

out_file <- file.path(
  hist_out_dir, 
  paste0("Globio4_landuse_10sec_2020_World_crop_int_broad.tif")
  )

writeRaster(
  r2_hist,
  out_file,
  datatype  = "INT1S",
  NAflag    = -1,
  overwrite = TRUE,
  gdal = c("COMPRESS=DEFLATE")
  )

unlink(z_tmp, recursive = TRUE)
rm(r_hist, r2_hist)
```

## 9. Binarization

Each reclassified cell now holds a harmonized broad land-use class ID. For the EBV Cube, each broad class is split into its own binary raster: `1` where the class is present, `0` elsewhere. Each broad class is one EBV entity, so each binary raster is one entity layer in the final cube.

### 9.1 Define paths for binarization

```{r}

# Define paths
tif_root <- output_root

out_data_root <- paste0(
  output_root,
  "/binary"
)

dir.create(out_data_root, showWarnings = FALSE, recursive = TRUE)

broad_ids <- unique(globio_lut$broad_id)

HIST_DIR <- file.path(tif_root, "Historical_2020")
BASE_DIR <- file.path(tif_root, "Baseline")

```

### 9.2 Binarize rasters without parallel processing {#sec-9.2-binarize-rasters-without-parallel-processing}

One binary raster per broad class, for each scenario and year. For each input raster, the workflow builds one binary layer per class and writes them as `entity_XX.tif`. For non-Baseline scenarios, 2020 is skipped here — the shared baseline is handled in [Section 9.4](#sec-9.4-binarize-historical-2020).

```{r eval = FALSE}
for (sc in scenarios) {
  for (yr in years) {
    
    if (yr == 2020 && sc != "Baseline") next
    
    out_root <- file.path(out_data_root, sc, yr)
    dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

    if (sc == "Baseline") {
      cand <- list.files(BASE_DIR, pattern="\\.tif$", full.names=TRUE)
    } else {
      SC_DIR <- file.path(tif_root, sc)
      cand <- list.files(SC_DIR, pattern="\\.tif$", full.names=TRUE)
    }

    pat_year <- paste0("_", yr, "_")
    src <- cand[grepl(pat_year, basename(cand))]

    if (length(src) == 0) stop("No raster found for ", sc, " ", yr)
    
    src <- src[1] 
    
    message("RUN: ", sc, " ", yr, " | ", basename(src))
    cat("Input raster:", src, "\n")
    r <- rast(src)
  
    for (i in seq_along(broad_ids)) {
      
      bin_r <- (r == broad_ids[i])  
      
      out <- file.path(out_root, sprintf("entity_%02d.tif", i))

      writeRaster(
        bin_r, out,
        overwrite = TRUE,
        datatype  = "INT1S",
        NAflag    = -1,
        gdal      = c("COMPRESS=DEFLATE")
      )
      rm(bin_r); gc()
    }
    rm(r); gc()
  }
}

```

### 9.3 Binarize rasters with parallel processing {#sec-9.3-binarize-rasters-with-parallel-processing}

The same operation, but processing scenario-year combinations in parallel. Each worker reads one raster and writes its full set of entity rasters. This is faster but uses more memory, so set the number of workers to match your available RAM and CPU.

```{r}

jobs <- expand.grid(
  sc = scenarios, 
  yr = years, 
  stringsAsFactors = FALSE
  )

# Exclude 2020 and Baseline
jobs <- jobs[!(jobs$yr == 2020 & jobs$sc != "Baseline"), ]

plan(multisession, workers = 8) # Check available CPU resources


future_lapply(seq_len(nrow(jobs)), function(k) {

  # each worker loads terra + sets temp
  library(terra)
  
  terraOptions(
    todisk  = TRUE,
    memfrac = 0.3,
    tempdir = local_tmp 
    )
  
  sc <- jobs$sc[k]
  yr <- jobs$yr[k]

  out_root <- file.path(out_data_root, sc, yr)
  dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

  # select input
  if (sc == "Baseline") {
    cand <- list.files(BASE_DIR, pattern="\\.tif$", full.names=TRUE)
  } else {
    SC_DIR <- file.path(tif_root, sc)
    cand <- list.files(SC_DIR, pattern="\\.tif$", full.names=TRUE)
  }

  pat_year <- paste0("_", yr, "_")
  src <- cand[grepl(pat_year, basename(cand))]

  if (length(src) == 0) stop("No raster found for ", sc, " ", yr)
  
  src <- src[1]

  r <- rast(src)

  for (i in seq_along(broad_ids)) {
    
    # binarization
    bin_r <- (r == broad_ids[i])
    out <- file.path(out_root, sprintf("entity_%02d.tif", i))

    writeRaster(
      bin_r, out,
      overwrite = TRUE,
      datatype  = "INT1S",
      NAflag    = -1,
      gdal      = c("COMPRESS=DEFLATE")
    )
    rm(bin_r); gc()
  }
  rm(r); gc()
})
```

### 9.4 Binarize historical 2020 {#sec-9.4-binarize-historical-2020}

The 2020 raster is processed separately because NfN, NaC, and NfS all share it as their starting point — a common 2020 state so that differences between them reflect only scenario assumptions. Baseline uses its own 2020 raster instead, matching its business-as-usual trajectory.

```{r}

sc <- "Historical"
yr <- 2020

out_root <- file.path(
  out_data_root, 
  sc, 
  yr
  )

dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

cand <- list.files(HIST_DIR, pattern="\\.tif$", full.names=TRUE)
pat_year <- paste0("_", yr, "_")
src <- cand[grepl(pat_year, basename(cand))]


if (length(src) == 0) stop("No Historical raster found for 2020 (pattern ", pat_year, ")")

r <- rast(src)

start.time <- Sys.time()
for (i in seq_along(broad_ids)) {
  bin_r <- (r == broad_ids[i])
  out <- file.path(out_root, sprintf("entity_%02d.tif", i))

  writeRaster(
    bin_r, out,
    overwrite = TRUE,
    datatype  = "INT1S",
    NAflag    = -1,
    gdal      = c("COMPRESS=DEFLATE")
  )
  rm(bin_r); gc()
}
rm(r); gc()

# Calculate time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

## 10. Create EBV Cube NetCDF file

### 10.1 Check paths and available binary rasters

Define the input/output paths for this step — the metadata JSON and the target NetCDF file — and confirm the binary rasters from Section 9 are present in the expected folders. This check catches missing inputs before the cube is built.

```{r}

# Define date and check id 
creation_date <- 20260223
cat("Created date:", format(as.Date(as.character(creation_date), format = "%Y%m%d")), "\n")
cat("ID: ", id, "\n")

# Define paths
root <- paste0('../json')
js <- paste0(root, "/metadata_", id, ".json")

# Note: 
# Please follow the naming convention 
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0('../cubes', '/alkemade_ecostr_id',id,'_',creation_date,'.nc')


cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Binary root folder: ", out_data_root, "\n")

# List all tiffs
tif_files <- list.files(
  out_data_root, 
  pattern='*\\.tif$', 
  recursive = TRUE, 
  full.names = TRUE
  )

cat("Total binary GeoTIFF files found:", length(tif_files), "\n")
```

### 10.2 Define EBV Cube structure

Define the cube's spatial, temporal, and thematic components: entities, extent and resolution, CRS, time steps, fill value, and data precision. Entities here are the harmonized broad land-use classes.

```{r}

cat("Entity variable\n")
cat("=======================\n")
cat("Entity Values: ", broad_ids, "\n")

# Entity names
entities <- unique(globio_lut$broader_category)

# Remove water bodies and permanent snow and ice
entities <- entities[!entities %in% c("Water bodies", "Permanent snow and ice")]

cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")

# Spatio-temporal setting
r0 <- rast(tif_files[1])
ext <- as.vector(ext(r0))
res <- res(r0)
epsg <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = years, extent = ext, res = res, epsg = epsg)

# Fill value and datatype
fillvalue <- -1
prec <- "byte"
```

### 10.3 Create empty NetCDF file

Create the empty NetCDF file from the properties defined above. It holds the cube structure but no raster data yet. that comes in [Section 11](#sec-11.-populate-the-ebv-cube).

```{r}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = ext,
  resolution = res,
  timesteps  = as.character(years),
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)
```

At this stage, the file contains the EBV cube structure but not the data, which will be incorporate in the next step.

## 11. Populate the EBV Cube {#sec-11.-populate-the-ebv-cube}

Insert the binary rasters into the cube one by one. For each scenario and time step, the workflow picks the matching folder — in 2020, Baseline uses `Baseline/2020` while NfN, NaC, and NfS use the shared `Historical/2020`; later years use each scenario's own folder — and places every `entity_XX.tif` at its correct scenario / time step / entity position.

```{r}

# Get the properties and entities
props <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

# Define metric
metric_no <-  1
timesteps <- years

cat("Entities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (t_no in seq_along(timesteps)) {

    yr <- timesteps[t_no]
    
    # Historical 2020 shared by all scenarios
    if (yr == 2020) {
      if (sc == "Baseline") {
        in_dir <- file.path(out_data_root, "Baseline", "2020")
      } else {
        in_dir <- file.path(out_data_root, "Historical", "2020")
      }
    } else {
      in_dir <- file.path(out_data_root, sc, as.character(yr))
    }

    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")
    cat("Input dir:", in_dir, "\n")

    if (!dir.exists(in_dir)) {
      stop("Input directory does not exist: ", in_dir)
    }
    
    for (e_no in 1:n_entities) {

      tif_in <- file.path(
        in_dir, 
        sprintf("entity_%02d.tif", e_no)
        )
      
      if (!file.exists(tif_in)) {
        stop("Missing entity file: ", tif_in)
      }
      
      ebv_add_data(
        filepath_nc = nc,
        data        = tif_in,
        scenario    = s_no,
        metric      = metric_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = TRUE
      )

      gc(verbose = FALSE)
      cat("  ✔ inserted entity", e_no, "\n")
    }
  }
}

# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken


cat("\n🎉 DONE. All binary entities inserted into:\n", nc, "\n")
```

## 12. Validation and visualization

A quick check that the cube is readable and that the entities and time steps
line up.

```{r}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@ebv_cube$type
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 1,
        timestep = 1, classes = 2, verbose = FALSE)
```


## 13. Clean up temporary binary rasters

With everything inserted, the temporary binary folder can be removed to free disk space.

```{r}
cat("Removing temporary binary raster folder:", out_data_root, "\n")
unlink(out_data_root, recursive = TRUE)

if (!dir.exists(out_data_root)) {
  cat("✔ Temporary binary folder successfully removed.\n")
} else {
  warning("Binary folder could not be removed: ", out_data_root)
}
```
