---
title: "GLOBIO Land-Use - Aggregation to 0.5° and EBV Cube generation"
author: "Emmanuel Oceguera-Conchas, Andres Marmol Guijarro"
date: 11-02-2026
output: html_document
---

## 1. Introduction

This workflow processes GLOBIO land-use data and converts it into an EBV Cube NetCDF file.

Unlike the high-resolution binary workflow, it aggregates the preprocessed layers from 0.002777° (approx. 300 m at the equator) to a coarser resolution of 0.5°, storing each class's proportional cover per grid cell.

The workflow comprises two core steps:

-   *Aggregation to 0.5°:* The preprocessed GLOBIO rasters (≈0.002777°, about 300 m) are aggregated to 0.5°. Aggregation runs per land-use class, computing each class's proportional cover within every 0.5° cell — so the outputs are continuous composition layers, not categorical maps.

-   *EBV Cube generation:* The aggregated rasters are inserted into an EBV Cube NetCDF file, organized by scenario, metric, time step, and entity (the harmonized land-use class), ready for publication on the EBV Portal.

In the notebook, these steps span four sections: preparing the working environment (loading `terra`, `ebvcube`, `stringr`, `dplyr` and setting paths), aggregating the rasters, building the EBV Cube structure, and populating it.

### 1.1 GLOBIO scenarios and time steps

The workflow processes four GLOBIO scenarios across four years each. The 2020 historical year is shared across all scenarios, so differences in projections reflect only future changes.

| Scenario           | Abbreviation | Available Years        |
|--------------------|--------------|------------------------|
| SSP2-Baseline      | Baseline     | 2020, 2030, 2050, 2070 |
| Nature as Culture  | NaC          | 2020, 2030, 2050, 2070 |
| Nature for Society | NfS          | 2020, 2030, 2050, 2070 |
| Nature for Nature  | NfN          | 2020, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   The preprocessed GLOBIO binary rasters, in the expected folder structure
-   `GLOBIO_categories.csv`, mapping original GLOBIO classes to broad classes
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   Enough temporary disk space for raster processing
-   A working `ebvcube` install, plus `terra`, `stringr`, and `dplyr`

```{r}
# note: Set cache = FALSE when re-running after changes to input data or paths
knitr::opts_chunk$set(cache = TRUE)

```

## 2. Load libraries

Here we load the packages used throughout the workflow: `terra` for raster processing (reading, aggregating, writing), `ebvcube` for creating and filling the EBV Cube NetCDF file, and `stringr` and `dplyr` for string handling and data preparation.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")
# install.packages("stringr")
# install.packages("dplyr")

library(terra)
library(ebvcube)
library(stringr)
library(dplyr)

```

## 3. Configure temporary raster storage

Aggregation can write a lot of temporary data to disk, so point `terra` at a drive with enough free space.

```{r}

local_temp <- "D:/temp" # Define local disk
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp 
)

cat("Local temporary folder: ", local_temp)
```

## 4. Define input and output paths

Input: the folder of preprocessed binary rasters. Output: the folder for the aggregated 0.5° rasters.

```{r}

in_root  <- "../preprocessed/10arcsecs/binary"
out_root <- "../preprocessed/10arcsecs/0.5degree "
dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

cat("Input root:", in_root, "\n")
cat("Output root:", out_root, "\n")
```

## 5. Define the main workflow settings

Set the dataset ID, scenarios, years, and aggregation settings.

```{r}

id <- 103
years  <- c(2020, 2030, 2050, 2070)
scenarios <- c("NfN", "NaC", "NfS", "Baseline")

# Aggregation settings
target_deg <- 0.5   # target resolution in degrees
nodata_val <- -1   # NoData value used during aggregation

cat("ID:", id, "\n")
cat("Years:", paste(years, collapse = ", "), "\n")
cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
```

## 6. Load the reclassification table

Load the lookup table that maps original GLOBIO land-use values to the broad harmonized classes used in the subsequent steps.

```{r}

table_path <- "../GLOBIO_categories.csv" 
globio_lut <- read.csv(table_path)  

cat("Unique original codes:", n_distinct(globio_lut$values), "\n")
cat("Unique broad classes:", n_distinct(globio_lut$broad_id), "\n") 
cat("Broad classes:\n", paste(unique(globio_lut$broader_category), collapse = ", "), "\n") 
head(globio_lut)
```

## 7. Aggregation to 0.5° resolution {#sec-7.-aggregation-to-0.5°-resolution-}

Aggregate the high-resolution rasters to 0.5°. Each output cell holds the proportional cover of one land-use class within that coarse grid cell. Baseline uses its own input for every year; NfN, NaC, and NfS use the shared Historical 2020 raster for 2020 and their own rasters for 2030, 2050, and 2070.

### 7.1 Aggregate scenario rasters

```{r}

for (sc in scenarios){
  for(yr in years){
    
    if (sc != "Baseline" && yr == 2020) {
      # All non-baseline scenarios use Historical 2020
      in_dir <- file.path(in_root, "Historical", "2020")
    } else {
      # Baseline 2020 + all other scenario years
      in_dir <- file.path(in_root, sc, as.character(yr))
    }
    
    cat("\n==========", sc, yr, "==========\n")
    # Produce the out path
    
    out_dir <- file.path(
      out_root, 
      sc, 
      as.character(yr)
      )
    
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    tifs <- list.files(
      in_dir, 
      pattern="\\.tif$", 
      full.names=TRUE
      )
    
    for (tif in tifs){
      
      cat("Processing:", basename(tif), "\n")
      
      r <- rast(tif)
      NAflag(r) <- nodata_val
      fact <- target_deg / res(r)[1]
      
      # Aggregate - Upscale
      r_d <- aggregate(
        r, 
        fact = fact, 
        fun = mean, 
        na.rm = TRUE
        )
      
      out_file <- file.path(
        out_dir, 
        sub("\\.tif$", "_p05.tif",basename(tif))
        )
      
      # Write raster
      writeRaster(
        r_d, 
        out_file, 
        datatype  = "FLT4S", 
        overwrite=TRUE,
        gdal=c("COMPRESS=DEFLATE"))
      
      rm(r, r_d)
      gc()
    }
  }
}
```

### 7.2 Aggregate historical 2020

The 2020 raster is aggregated separately because NfN, NaC, and NfS all share it as their reference year, so differences between these scenarios in later years reflect only scenario assumptions. Baseline uses its own 2020 raster instead, matching its business-as-usual trajectory.

```{r}

# Define input and output paths 
in_dir_hist  <- file.path(in_root, "Historical", "2020")
out_dir_hist <- file.path(out_root, "Historical", "2020")
dir.create(out_dir_hist, recursive = TRUE, showWarnings = FALSE)

tifs_hist <- list.files(
  in_dir_hist, 
  pattern="\\.tif$", 
  full.names=TRUE
  )

for (tif in tifs_hist) {

  cat("Processing Historical:", basename(tif), "\n")
  r <- rast(tif)
  NAflag(r) <- nodata_val
  
  fact <- target_deg / res(r)[1]
  
  # Aggregate - Upscale
  r_d <- aggregate(
    r, 
    fact = fact,
    fun = mean,
    na.rm = TRUE
    )

  out_file <- file.path(
    out_dir_hist, 
    sub("\\.tif$", "_p05.tif", basename(tif))
    )

  writeRaster(
    r_d,
    out_file,
    datatype  = "FLT4S",
    overwrite = TRUE,
    wopt      = list(gdal = c("COMPRESS=DEFLATE"))
    )
  rm(r, r_d)
  gc()
}
```

## 8. Create EBV Cube NetCDF file

### 8.1 Check paths and available binary rasters

Define the input/output paths for this step — the metadata JSON and the target NetCDF file — and confirm the aggregated rasters from [Section 7](#sec-7.-aggregation-to-0.5°-resolution-) are present in the expected folders. This check catches missing inputs before the cube is built.s.

```{r}

# Define date and check id 
creation_date <- 20260216
cat("Created date:", format(as.Date(as.character(creation_date), format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Define paths
root <- paste0('../json')
js <- paste0(root, "/metadata_", id, ".json")

# Note: 
# Please follow the naming convention 
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0('../cubes', '/alkemade_ecostr_id',id,'_',creation_date,'.nc')

cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Binary root folder: ", out_root, "\n")

# List all tiffs
tif_files <- list.files(
  out_root, 
  pattern='*\\.tif$', 
  recursive = TRUE, 
  full.names = TRUE
  )

cat("Total binary GeoTIFF files found:", length(tif_files), "\n")
```

### 8.2 Define EBV Cube structure

Define the cube's spatial, temporal, and thematic components: entities, extent and resolution, CRS, time steps, fill value, and data precision. Entities here are the harmonized broad land-use classes.

```{r}

cat("Entity variable\n")
cat("=======================\n")
cat("Entity Values: ", unique(globio_lut$broad_id), "\n")

# Entity names
entities <- unique(globio_lut$broader_category)

# Remove water bodies and permanent snow and ice
entities <- entities[!entities %in% c("Water bodies", "Permanent snow and ice")]

cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")


# Spatio-temporal setting
r0 <- rast(tif_files[1])         
ext <- as.vector(ext(r0))
res <- res(r0)
epsg <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = years, extent = ext, res = res, epsg = epsg)

# Fill value and datatype
fillvalue <- nodata_val
prec <- 'float'
```

### 8.2 Create empty NetCDF file

After defining the main properties of the EBV Cube, an empty NetCDF file is created. This file establishes the dataset’s structure, including its spatial extent, resolution, coordinate reference system, time steps, entities, and technical settings such as fill value and data precision.

```{r}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = ext,
  resolution = res,
  timesteps  = as.character(years),
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)
```

At this stage, the file contains the EBV cube structure but not the data, which will be incorporate in the next step.

## 9. Populate the EBV Cube

Insert the aggregated rasters into the cube one by one. For each scenario and time step, the workflow picks the matching folder — in 2020, Baseline uses `Baseline/2020` while NfN, NaC, and NfS use the shared `Historical/2020`; later years use each scenario's own folder — and places every `entity_XX_p05.tif` at its correct scenario / time step / entity position..

```{r}
# Get the properties and entities
props <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

# Define metric
metric_no <- 1
timesteps <- years

cat("\nEntities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {
  
  sc <- scenarios[s_no]
  
  for (t_no in seq_along(timesteps)) {
    
    yr <- timesteps[t_no]
    
    # Historical 2020 shared by all scenarios
    if (yr == 2020) {
        if (sc == "Baseline"){
            in_dir <- file.path(tif_root, "Baseline", "2020")
        } else {
            in_dir <- file.path(tif_root, "Historical", "2020")
        }
    } else {
        in_dir <- file.path(tif_root, sc, as.character(yr))
    }

    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")
    cat("Input dir:", in_dir, "\n")

    if (!dir.exists(in_dir)) {
      stop("Input directory does not exist: ", in_dir)
    }
  
    for (e_no in 1:n_entities) {

      tif_in <- file.path(
        in_dir, 
        sprintf("entity_%02d_p05.tif", e_no)
        )

      if (!file.exists(tif_in)) {
        stop("Missing entity file: ", tif_in)
      }

      ebv_add_data(
        filepath_nc = nc,
        data        = tif_in,
        scenario    = s_no,
        metric      = metric_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = TRUE
      )
      
      gc(verbose = FALSE)
      cat("  ✔ inserted entity", e_no, "\n")
    }
  }
}


# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

cat("\n🎉 DONE. All binary entities inserted into:\n", nc, "\n")
```

## 11. Validation and visualization

A quick check that the cube is readable and that the entities and time steps line up.

```{r}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@ebv_cube$type
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 1,
        timestep = 1, verbose = FALSE)
```

## 12. Clean up temporary aggregated rasters

With everything inserted, the temporary aggregated folder can be removed to free disk space.

```{r}
cat("Removing temporary aggregated raster folder:", out_root, "\n")
unlink(out_root, recursive = TRUE)

if (!dir.exists(out_root)) {
  cat("✔ Temporary aggregated folder successfully removed.\n")
} else {
  warning("Aggregated folder could not be removed: ", out_root)
}
```
