---
title: "GLOBIO Land-Use Reclassification and EBV Cube Generation Workflow"
author: "Emmanuel Oceguera Conchas"
date: 23-02-2026
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Introduction

This workflow processes high-resolution GLOBIO land-use data and
converts it into an EBV Cube NetCDF file.

The original GLOBIO data is at a resolution of 0.002777° (approx. 300 m)
and contains numerous specific land-use categories. For compatibility
with the EBV Cube structure, these are first grouped into a smaller set
of harmonized broad land-use classes.

The workflow comprises three core steps:

1.  *Reclassification:* Original GLOBIO classes are mapped to harmonized
    broad land-use classes using a lookup table.

2.  *Binarization:* Each broad land-use class is converted into a binary
    raster (1 for presence, 0 for absence).

3.  *EBV Cube NetCDF file generation:* The binary rasters are structured
    into an EBV Cube, organizing the data by scenario, metric, time
    step, and entity (the broad land-use class).

This high-resolution product is the foundation for a subsequent workflow
that aggregates the binary rasters to a 0.5° resolution.

### 1.1 Overview of the workflow

This notebook follows five core processing stages:

1.  *Prepare the working environment:* Load necessary R packages
    (`terra`, `ebvcube`, `tidyverse, future.apply`), define temporary
    storage, and set input/output paths.

2.  *Reclassify GLOBIO rasters*: Convert the original land-use
    categories into harmonized, broader classes.

3.  *Create binary entity rasters*: Generate a separate binary raster
    for each broad land-use class, scenario, and year.

4.  *Build the EBV Cube NetCDF file*: Define the structure of the EBV
    Cube, including the spatial extent, metric, time steps, and
    entities, and initialize the file.

5.  *Populate the EBV Cube NetCDF file:* Insert processed raster data
    into the correct positions within the NetCDF file, corresponding to
    the specific scenario, entity, and time step.

### 1.2 GLOBIO scenarios and time steps

The workflow processes GLOBIO land-use data for four key scenarios and
four specific years. The historical year 2020 is consistently used
across all scenarios to ensure that differences in projections reflect
only future changes.

|                        |                  |                        |
|------------------------|------------------|------------------------|
| **Scenario**           | **Abbreviation** | **Available Years**    |
| **SSP2-Baseline**      | Baseline         | 2020, 2030, 2050, 2070 |
| **Nature as Culture**  | NaC              | 2020, 2030, 2050, 2070 |
| **Nature for Society** | NfS              | 2020, 2030, 2050, 2070 |
| **Nature for Nature**  | NfN              | 2020, 2030, 2050, 2070 |

### 1.3 Before you start

To ensure the workflow runs smoothly, the following conditions must be
met:

-   The original GLOBIO ZIP archives and raster files are available in
    the expected folder structure
-   The reclassification table (`GLOBIO_categories.csv`) is available
    and contains the mapping from original GLOBIO land-use classes to
    harmonized broad classes
-   The `metadata.json` file for the EBV Cube dataset already exists
-   Sufficient temporary disk space is available for raster processing.
-   The `ebvcube` package is installed and working correctly
-   The required R packages, especially `terra`, `tidyverse`, and
    optionally `future.apply`, are installed

Before running the notebook, one additional decision is required:
Sections 9.2 and 9.3 provide two alternative binarization strategies —
sequential and parallel — and are mutually exclusive.

By default, Section 9.2 is set to `eval = FALSE` and the parallel
version in Section 9.3 will run. If you prefer sequential processing
(e.g. on a machine with limited CPU resources), set Section 9.2 to
`eval = TRUE` and Section 9.3 to `eval = FALSE` before knitting.

```{r}
# note: Set cache = FALSE when re-running after changes to input data or paths
knitr::opts_chunk$set(cache = TRUE)
```

## 2. Load libraries

In this section, we load the R packages required for raster processing,
data handling, and EBV Cube creation.

-   `terra` is used for reading, reclassifying, binarizing, and writing
    raster data

-   `ebvcube` is used to create and fill the EBV Cube NetCDF file

-   `tidyverse` is used for handling lookup tables, inspecting tabular
    data, and supporting data preparation steps

-   `future.apply` is used later in the workflow for optional parallel
    processing during binarization

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean user-defined objects, not knitr internals
rm(list = setdiff(ls(), c("knitr", "params")))
gc()

# install.packages("terra")
# install.packages("ebvcube")
# install.packages("tidyverse")
# install.packages("future.apply")

library(terra)
library(ebvcube)
library(tidyverse)
library(future.apply)
```

## 3. Configure temporary raster storage

Processing large raster files during reclassification and binarization
can generate substantial temporary data on disk. For this reason, it is
recommended to define a temporary folder on a drive with sufficient free
storage space.

```{r}

local_temp <- "D:/temp" # Define local disk
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp 
)

cat("Local temporary folder: ", local_temp)
```

## 4. Define input and output paths

Here we define:

-   the folder containing the original ZIP archives with GLOBIO rasters
-   the output folder for the reclassified broad-class rasters

```{r}

input_root <- "../raw/10arcsecs"
output_root <- "../preprocessed/10arcsecs" 
dir.create(output_root, showWarnings = FALSE, recursive = TRUE)


cat("Input root:", input_root, "\n")
cat("Output root:", output_root, "\n")

```

## 5. Define the main workflow settings

In this step, we define the dataset ID, scenarios, and years to be
processed.

```{r}

id <-  98
years  <- c(2020, 2030, 2050, 2070)
scenarios <- c("NfN", "NaC", "NfS", "Baseline")


cat("ID:", id, "\n")
cat("Years:", paste(years, collapse = ", "), "\n")
cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
```

## 6. Load the reclassification table

In this step, we load the lookup table used to map the original GLOBIO
land-use category values to the broader land-use classes applied in this
workflow.

This table defines how the original GLOBIO land-use classes are mapped
to the broader harmonized classes used in the subsequent binarization
and EBV Cube creation steps.

```{r}

table_path <- "../GLOBIO_categories.csv"
globio_lut <- read.csv(table_path)

cat("Unique original codes:", n_distinct(globio_lut$values), "\n")
cat("Unique broad classes:", n_distinct(globio_lut$broad_id), "\n")
cat("Broad classes:\n", paste(unique(globio_lut$broader_category), collapse = ", "), "\n")
head(globio_lut)
```

### 6.1 Build a reclassification matrix

`terra::classify()` expects a two-column matrix in which the first
column contains the original values and the second column contains the
new reclassified values.

```{r}
rcl <- as.matrix(globio_lut[, c("values", "broad_id")])
head(rcl)
```

## 7. Identify the ZIP files to process

The original GLOBIO data are stored in ZIP archives. In this step, we
list all ZIP files and keep only those needed for this workflow.

The filtering logic does two things:

-   it excludes the historical ZIP archive, because that file is handled
    separately later

-   it keeps only the ZIP files corresponding to the selected scenarios,
    using the version identifier `1.9` that appears in the filenames of
    the BES-SIM2 scenario archives (e.g.
    `GLOBIO4_landuse_10sec_2050_NfN_v1.9.zip`), and a separate match for
    the `Baseline.zip` archive which follows a different naming
    convention

```{r}
zip_files <- list.files(
  input_root,
  pattern = "\\.zip$",
  full.names = TRUE,
  recursive = TRUE
)

zip_files <- zip_files[!grepl("Historical_2000-2020", basename(zip_files))]
zip_files <- zip_files[
  grepl("1\\.9", basename(zip_files))  | grepl("Baseline\\.zip$", basename(zip_files))
]

cat("Total ZIP files to process:", length(zip_files), "\n")
head(zip_files, 10)
```

## 8. Reclassification

This step converts the original GLOBIO land-use category values into the
broader land-use classes defined for this workflow.

The reclassification is based on a lookup table that links each original
GLOBIO value to a corresponding broad-class ID. As a result, the output
raster no longer stores the original detailed category values, but the
harmonized class IDs used in the subsequent steps.

### 8.1 Reclassify scenario rasters

For each ZIP archive, the workflow:

-   extracts the files into a temporary directory

-   searches for all `.tif` files contained in the archive

-   identifies the scenario and year from the file path or filename

-   keeps only the selected years

-   applies the reclassification table to the raster values

-   writes the reclassified raster to disk

The output rasters are written with the following settings:

-   data type: `INT1S`

-   NoData value: `-1`

-   compression: `DEFLATE`

This keeps file sizes manageable and ensures consistent NoData handling
in the subsequent binarization step.

```{r}

# Check time
start_time <- Sys.time()

for (z in zip_files) {
  
  z_base <- tools::file_path_sans_ext(basename(z))
  
  z_tmp  <- file.path(
    local_temp, 
    paste0(z_base, "_", Sys.getpid())
    )
  
  dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)

  cat("\n=== Unzipping ===\n", z, "\n-> ", z_tmp, "\n")
  unzip(z, exdir = z_tmp)

  tif_files <- list.files(
    z_tmp,
    pattern = "\\.tif$",
    full.names = TRUE,
    recursive = TRUE
  )

  for (p in tif_files) {

    yr <- as.integer(stringr::str_extract(basename(p), "\\d{4}"))
    
    if (is.na(yr) || !(yr %in% years)) next

    p_norm <- gsub("\\\\", "/", p)

    scenario <- NA_character_
    if (grepl("NaC", p_norm)) scenario  <- "NaC"
    if (grepl("NfN", p_norm)) scenario  <- "NfN"
    if (grepl("NfS", p_norm)) scenario  <- "NfS"
    if (grepl("Baseline", p_norm)) scenario  <- "Baseline"

    if (is.na(scenario)) {
      warning("Could not parse scenario, skipping: ", p)
      next
    }

    cat("Reclassifying:", scenario, "year", yr, "\n")

    r  <- rast(p)
    r2 <- classify(r, rcl, others = NA)

    out_dir <- file.path(
      output_root, 
      scenario
      )
    
    dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

    out_file <- file.path(
      out_dir, paste0("Globio4_landuse_10sec_", yr, 
      "_World_crop_int_withPattern_broad.tif")
      )

    writeRaster(
      r2,
      out_file,
      datatype  = "INT1S",
      NAflag    = -1,
      overwrite = TRUE,
      gdal = c("COMPRESS=DEFLATE")
    )

    rm(r, r2)
    gc()
    terra::tmpFiles(remove = TRUE)
  }

  unlink(z_tmp, recursive = TRUE)
}

total_time <- Sys.time() - start_time
cat("\nTotal processing time:", total_time, "\n")

```

### 8.2 Reclassify historical 2020 raster

The historical 2020 layer is processed separately because it is stored
in its own ZIP archive and serves as the shared baseline for the future
scenarios.

This step ensures that all scenarios can later use the same 2020
reference.

```{r}

hist_zip <- file.path(
  input_root, 
  "Historical_2000-2020.zip"
  )

hist_out_dir <- file.path(
  output_root,
  "Historical_2020"
  )

dir.create(hist_out_dir, showWarnings = FALSE, recursive = TRUE)

z_base <- tools::file_path_sans_ext(basename(hist_zip))

z_tmp  <- file.path(
  local_temp, 
  paste0(z_base, "_", Sys.getpid())
  )

dir.create(z_tmp, showWarnings = FALSE, recursive = TRUE)
unzip(hist_zip, exdir = z_tmp)

src <- list.files(
  z_tmp,
  pattern = "Globio4_landuse_10sec_2020_.*\\.tif$",
  full.names = TRUE,
  recursive = TRUE
)

stopifnot(length(src) == 1)

r_hist  <- rast(src)
r2_hist <- classify(r_hist, rcl, others = NA)

out_file <- file.path(
  hist_out_dir, 
  paste0("Globio4_landuse_10sec_2020_World_crop_int_broad.tif")
  )

writeRaster(
  r2_hist,
  out_file,
  datatype  = "INT1S",
  NAflag    = -1,
  overwrite = TRUE,
  gdal = c("COMPRESS=DEFLATE")
  )

unlink(z_tmp, recursive = TRUE)
rm(r_hist, r2_hist)
```

## 9. Binarization

After reclassification, each raster cell stores the ID of a broad
land-use class. To prepare these data for the EBV cube NetCDF file, each
broad class is converted into a separate binary raster, where cells
belonging to that class are assigned `1` and all other cells are
assigned `0`.

In this dataset, each harmonized broad land-use class is treated as one
EBV entity, and each binary raster therefore represents a single entity
layer in the final cube.

### 9.1 Define paths for binarization

```{r}

# Define paths
tif_root <- output_root

out_data_root <- paste0(
  output_root,
  "/binary"
)

dir.create(out_data_root, showWarnings = FALSE, recursive = TRUE)

broad_ids <- unique(globio_lut$broad_id)

HIST_DIR <- file.path(tif_root, "Historical_2020")
BASE_DIR <- file.path(tif_root, "Baseline")

```

### 9.2 Binarize rasters without parallel processing

This step creates one binary raster per broad land-use class for each
scenario and year. For every input raster, the workflow selects the
corresponding reclassified file, creates a binary layer for each class,
and writes the outputs to disk as `entity_XX.tif`.

For non-Baseline scenarios, the year 2020 is not processed here because
the shared historical baseline is handled separately.

```{r eval = FALSE}
for (sc in scenarios) {
  for (yr in years) {
    
    if (yr == 2020 && sc != "Baseline") next
    
    out_root <- file.path(out_data_root, sc, yr)
    dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

    if (sc == "Baseline") {
      cand <- list.files(BASE_DIR, pattern="\\.tif$", full.names=TRUE)
    } else {
      SC_DIR <- file.path(tif_root, sc)
      cand <- list.files(SC_DIR, pattern="\\.tif$", full.names=TRUE)
    }

    pat_year <- paste0("_", yr, "_")
    src <- cand[grepl(pat_year, basename(cand))]

    if (length(src) == 0) stop("No raster found for ", sc, " ", yr)
    
    src <- src[1] 
    
    message("RUN: ", sc, " ", yr, " | ", basename(src))
    cat("Input raster:", src, "\n")
    r <- rast(src)
  
    for (i in seq_along(broad_ids)) {
      
      bin_r <- (r == broad_ids[i])  
      
      out <- file.path(out_root, sprintf("entity_%02d.tif", i))

      writeRaster(
        bin_r, out,
        overwrite = TRUE,
        datatype  = "INT1S",
        NAflag    = -1,
        gdal      = c("COMPRESS=DEFLATE")
      )
      rm(bin_r); gc()
    }
    rm(r); gc()
  }
}

```

### 9.3 Binarize rasters with parallel processing

For larger datasets, binarization can be accelerated by processing
different scenario-year combinations in parallel. Each worker reads one
reclassified raster and creates the full set of binary entity rasters
for that input.

This can reduce processing time, but it also increases memory use. The
number of workers should therefore be chosen according to the available
RAM and CPU resources.

```{r}

jobs <- expand.grid(
  sc = scenarios, 
  yr = years, 
  stringsAsFactors = FALSE
  )

# Exclude 2020 and Baseline
jobs <- jobs[!(jobs$yr == 2020 & jobs$sc != "Baseline"), ]

plan(multisession, workers = 8) # Check available CPU resources


future_lapply(seq_len(nrow(jobs)), function(k) {

  # each worker loads terra + sets temp
  library(terra)
  
  terraOptions(
    todisk  = TRUE,
    memfrac = 0.3,
    tempdir = local_temp 
    )
  
  sc <- jobs$sc[k]
  yr <- jobs$yr[k]

  out_root <- file.path(out_data_root, sc, yr)
  dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

  # select input
  if (sc == "Baseline") {
    cand <- list.files(BASE_DIR, pattern="\\.tif$", full.names=TRUE)
  } else {
    SC_DIR <- file.path(tif_root, sc)
    cand <- list.files(SC_DIR, pattern="\\.tif$", full.names=TRUE)
  }

  pat_year <- paste0("_", yr, "_")
  src <- cand[grepl(pat_year, basename(cand))]

  if (length(src) == 0) stop("No raster found for ", sc, " ", yr)
  
  src <- src[1]

  r <- rast(src)

  for (i in seq_along(broad_ids)) {
    
    # binarization
    bin_r <- (r == broad_ids[i])
    out <- file.path(out_root, sprintf("entity_%02d.tif", i))

    writeRaster(
      bin_r, out,
      overwrite = TRUE,
      datatype  = "INT1S",
      NAflag    = -1,
      gdal      = c("COMPRESS=DEFLATE")
    )
    rm(bin_r); gc()
  }
  rm(r); gc()
})
```

### 9.4 Binarize historical 2020

The historical 2020 raster is processed separately because it serves as
the shared starting point for the Nature for Nature (NfN), Nature as
Culture (NaC), and Nature for Society (NfS) scenario projections.

This common starting state ensures a fair comparison of differences
arising only from scenario assumptions. Conversely, the
"business-as-usual" Baseline scenario uses its own distinct 2020 raster,
tailored to its continuation trajectory. This methodological separation
maintains the integrity and comparability of the four modeled outcomes.

```{r}

sc <- "Historical"
yr <- 2020

out_root <- file.path(
  out_data_root, 
  sc, 
  yr
  )

dir.create(out_root, recursive = TRUE, showWarnings = FALSE)

cand <- list.files(HIST_DIR, pattern="\\.tif$", full.names=TRUE)
pat_year <- paste0("_", yr, "_")
src <- cand[grepl(pat_year, basename(cand))]


if (length(src) == 0) stop("No Historical raster found for 2020 (pattern ", pat_year, ")")

r <- rast(src)

start.time <- Sys.time()
for (i in seq_along(broad_ids)) {
  bin_r <- (r == broad_ids[i])
  out <- file.path(out_root, sprintf("entity_%02d.tif", i))

  writeRaster(
    bin_r, out,
    overwrite = TRUE,
    datatype  = "INT1S",
    NAflag    = -1,
    gdal      = c("COMPRESS=DEFLATE")
  )
  rm(bin_r); gc()
}
rm(r); gc()

# Calculate time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken
```

## 10. Create EBV Cube NetCDF file

### 10.1 Check paths and available binary rasters

Before creating the EBV Cube NetCDF file, we define the main input and
output paths for this step, including the metadata JSON file and the
location where the final NetCDF file will be written.

In addition, we check that the binary rasters created in the previous
step are available in the expected folder structure. These rasters
provide the binary entity layers that will later be inserted into the
EBV Cube.

Verifying their availability at this stage helps ensure that the NetCDF
creation step is based on the correct input files and that the workflow
can continue without missing-data errors.

```{r}

# Define date and check id 
creation_date <- 20260223
cat("Created date:", format(as.Date(as.character(creation_date), format = "%Y%m%d")), "\n")
cat("ID: ", id, "\n")

# Define paths
root <- paste0('../json')
js <- paste0(root, "/metadata_", id, ".json")

# Note: 
# Please follow the naming convention 
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0('../cubes', '/alkemade_ecostr_id',id,'_',creation_date,'.nc')


cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Binary root folder: ", out_data_root, "\n")

# List all tiffs
tif_files <- list.files(
  out_data_root, 
  pattern='*\\.tif$', 
  recursive = TRUE, 
  full.names = TRUE
  )

cat("Total binary GeoTIFF files found:", length(tif_files), "\n")
```

### 10.2 Define EBV Cube structure

Before creating the NetCDF file, we first define its *core spatial*,
*temporal*, and *thematic components*. These include the list of
entities, the spatial extent and resolution, the coordinate reference
system (CRS), the time steps, and the fill value and data precision to
be used in the file.

In this dataset, the entities correspond to the harmonized broad
land-use classes derived from the original GLOBIO categories.

```{r}

cat("Entity variable\n")
cat("=======================\n")
cat("Entity Values: ", broad_ids, "\n")

# Entity names
entities <- unique(globio_lut$broader_category)

# Remove water bodies and permanent snow and ice
entities <- entities[!entities %in% c("Water bodies", "Permanent snow and ice")]

cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")

# Spatio-temporal setting
r0 <- rast(tif_files[1])
ext <- as.vector(ext(r0))
res <- res(r0)
epsg <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = years, extent = ext, res = res, epsg = epsg)

# Fill value and datatype
fillvalue <- -1
prec <- "byte"
```

### 10.3 Create empty NetCDF file

After defining the main properties of the EBV Cube, an empty NetCDF file
is created. This file establishes the dataset’s structure, including its
spatial extent, resolution, coordinate reference system, time steps,
entities, and technical settings such as fill value and data precision.

```{r}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = ext,
  resolution = res,
  timesteps  = as.character(years),
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)
```

At this stage, the file contains the EBV cube structure but not the
binary raster data, which will be added in the next step.

## 11. Populate the EBV Cube

***Insert binary rasters into the NetCDF file***

Once the empty EBV NetCDF file has been created, the binary rasters are
inserted into the EBV Cube one by one.

For each scenario and time step, the workflow selects the corresponding
folder of binary entity rasters. In 2020, `Baseline` uses its own
`Baseline/2020` folder, while `NfN`, `NaC`, and `NfS` use the shared
`Historical/2020` folder. For later years, each scenario uses its own
folder.

The rasters stored as `entity_XX.tif` are then inserted into the correct
position in the EBV Cube according to scenario, time step, and entity.

```{r}

# Get the properties and entities
props <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

# Define metric
metric_no <-  1
timesteps <- years

cat("Entities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (t_no in seq_along(timesteps)) {

    yr <- timesteps[t_no]
    
    # Historical 2020 shared by all scenarios
    if (yr == 2020) {
      if (sc == "Baseline") {
        in_dir <- file.path(out_data_root, "Baseline", "2020")
      } else {
        in_dir <- file.path(out_data_root, "Historical", "2020")
      }
    } else {
      in_dir <- file.path(out_data_root, sc, as.character(yr))
    }

    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")
    cat("Input dir:", in_dir, "\n")

    if (!dir.exists(in_dir)) {
      stop("Input directory does not exist: ", in_dir)
    }
    
    for (e_no in 1:n_entities) {

      tif_in <- file.path(
        in_dir, 
        sprintf("entity_%02d.tif", e_no)
        )
      
      if (!file.exists(tif_in)) {
        stop("Missing entity file: ", tif_in)
      }
      
      ebv_add_data(
        filepath_nc = nc,
        data        = tif_in,
        scenario    = s_no,
        metric      = metric_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = TRUE
      )

      gc(verbose = FALSE)
      cat("  ✔ inserted entity", e_no, "\n")
    }
  }
}

# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken


cat("\n🎉 DONE. All binary entities inserted into:\n", nc, "\n")
```

## 12. Clean up temporary binary rasters

Once all binary rasters have been incorporated into the NetCDF file, the
temporary binary folder is no longer needed and can be removed to free
disk space.

```{r}
cat("Removing temporary binary raster folder:", out_data_root, "\n")
unlink(out_data_root, recursive = TRUE)

if (!dir.exists(out_data_root)) {
  cat("✔ Temporary binary folder successfully removed.\n")
} else {
  warning("Binary folder could not be removed: ", out_data_root)
}
```
