---
title: "GLOBIO Downsacaling process - EBV Dataset"
author: "Andres Camilo Marmol Guijarro, Emmanuel Oceguera-Conchas"
date: 2026-02-11
output: html_documen
editor_options: 
  markdown: 
    wrap: 72
---

## 1. Introduction

This workflow processes GLOBIO land-use rasters and converts them into
an EBV Cube NetCDF file. In contrast to the high-resolution binary
workflow, this notebook focuses on aggregating preprocessed land-use
layers to a coarser spatial resolution of 0.5° and storing the resulting
proportional cover values in an EBV-compliant data cube.\
\
The workflow comprises two core steps:\
\
1. *Aggregate the original raster data to 0.5°*\
The original GLOBIO land-use rasters have a spatial resolution of
approximately 0.002777° (about 300 m at the equator). In this step, they
are aggregated to a coarser resolution of 0.5°.\
\
2. *Build and populate the EBV Cube NetCDF file\
*Once the aggregated rasters have been created, they are inserted into
an EBV Cube NetCDF file that follows the standard EBV data structure and
can be prepared for publication in the EBV Portal.\
\
The aggregation is performed separately for each land-use class. This
approach calculates the proportional cover of each class within every
0.5° grid cell. As a result, the outputs are no longer simple
categorical maps, but continuous layers describing the land-use
composition of each coarse grid cell.\
\
The final NetCDF file follows the standard EBV Cube structure and is
organized by scenario, metric, time step, and entity. In this dataset,
the entities correspond to the different harmonized land-use classes.

### 1.1 Overview of the workflow

This notebook follows four core processing stages:

1.  *Prepare the working environment* Load the required R packages,
    define temporary raster storage, and set the main input and output
    paths.

2.  *Aggregate the original GLOBIO rasters to 0.5°*\
    Process the high-resolution rasters for all selected scenarios and
    years and write the aggregated outputs to disk.

3.  *Create the EBV Cube NetCDF file*\
    Define the spatial, temporal, and thematic structure of the EBV
    Cube, including resolution, time steps, entities, and technical
    settings, and initialize the empty NetCDF file.

4.  *Populate the EBV Cube NetCDF file* Insert the aggregated rasters
    into the correct position in the EBV Cube according to scenario,
    entity, and time step.

### 1.2 GLOBIO scenarios and time steps

The workflow processes GLOBIO land-use data for four key scenarios and
four selected years. The historical year 2020 is used as a common
reference point so that differences among future projections reflect
scenario assumptions rather than different starting conditions.

|                        |                  |                        |
|------------------------|------------------|------------------------|
| **Scenario**           | **Abbreviation** | **Available Years**    |
| **SSP2-Baseline**      | Baseline         | 2020, 2030, 2050, 2070 |
| **Nature as Culture**  | NaC              | 2020, 2030, 2050, 2070 |
| **Nature for Society** | NfS              | 2020, 2030, 2050, 2070 |
| **Nature for Nature**  | NfN              | 2020, 2030, 2050, 2070 |

### 1.3 Before you start

To ensure the workflow runs smoothly, the following conditions should be
met:

-   The original GLOBIO raster files are available in the expected
    folder structure.
-   The `metadata.json` file for the EBV dataset already exists.
-   The `GLOBIO_categories.csv` file is available and contains the
    land-use class names.
-   Sufficient temporary disk space is available for raster processing.
-   The `ebvcube` package is installed and working correctly.

This workflow assumes that the processed 0.5° rasters will be used to
create a valid EBV Cube NetCDF file compatible with the EBV Portal
structure.

```{r}
# note: Set cache = FALSE when re-running after changes to input data or paths
knitr::opts_chunk$set(cache = TRUE)

```

## 2. Load libraries

In this section, we load the R packages required for raster processing
and EBV Cube creation.

-   `terra` is used for reading, aggregating, and writing raster data.
-   `ebvcube` is used to create and populate the EBV Cube NetCDF file.
-   `stringr` and `dplyr` are used for string handling and basic data
    preparation.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")
# install.packages("stringr")
# install.packages("dplyr")

library(terra)
library(ebvcube)
library(stringr)
library(dplyr)

```

## 3. Configure temporary raster storage

R aster aggregation can create large temporary files. For this reason,
it is recommended to define a temporary folder on a drive with
sufficient free storage space.

```{r}

local_temp <- "D:/temp" # Define local disk
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp 
)

cat("Local temporary folder: ", local_temp)
```

## 4. Define input and output paths

Here we define:

-   the folder containing the input rasters

-   the folder where the aggregated 0.5° rasters will be stored

```{r}

in_root  <- "../preprocessed/10arcsecs/binary"
out_root <- "../preprocessed/10arcsecs/0.5degree "
dir.create(out_root, showWarnings = FALSE, recursive = TRUE)

cat("Input root:", in_root, "\n")
cat("Output root:", out_root, "\n")
```

## 5. Define the main workflow settings

In this step, we define the dataset ID, scenarios, and years to be
processed.

```{r}

id <- 103
years  <- c(2020, 2030, 2050, 2070)
scenarios <- c("NfN", "NaC", "NfS", "Baseline")

cat("ID:", id, "\n")
cat("Years:", paste(years, collapse = ", "), "\n")
cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
```

## 6. Load the land-use classes table

In this step, we load the lookup table used to map broader land-use
classes applied in this workflow.

This table defines how the original GLOBIO land-use classes are mapped
to the broader harmonized classes used in the subsequent EBV Cube
creation steps.

```{r}

table_path <- "../GLOBIO_categories.csv" 
globio_lut <- read.csv(table_path)  

cat("Unique original codes:", n_distinct(globio_lut$values), "\n")
cat("Unique broad classes:", n_distinct(globio_lut$broad_id), "\n") 
cat("Broad classes:\n", paste(unique(globio_lut$broader_category), collapse = ", "), "\n") 
head(globio_lut)
```

## 7. Aggregation to 0.5° resolution

In this section, the high-resolution GLOBIO rasters are aggregated to
0.5° resolution. Each output raster represents the proportional cover of
a single land-use class within each coarse grid cell.

Processing logic:

-   *Baseline:* aggregate all available years (2020, 2030, 2050, 2070)
    using the corresponding Baseline input rasters.

-   *NaC, NfN, and NfS:* use the shared Historical 2020 rasters for the
    year 2020, and the scenario-specific rasters for 2030, 2050, and

    2070. 

This ensures that all non-Baseline scenarios share the same historical
reference year, while future projections remain scenario-specific.

### 7.1 Process scenario-specific rasters

In this step, the GLOBIO rasters are aggregated by scenario and year.
For the Baseline scenario, each year is processed using its
corresponding input raster. For the NaC, NfN, and NfS scenarios, the
shared Historical 2020 raster is used for the year 2020, while the
future years 2030, 2050, and 2070 use the scenario-specific raster
inputs.

This ensures that all non-Baseline scenarios start from the same
historical reference year, while future projections remain
scenario-specific.

```{r}

for (sc in scenarios){
  for(yr in years){
    
    if (sc != "Baseline" && yr == 2020) {
      # All non-baseline scenarios use Historical 2020
      in_dir <- file.path(in_root, "Historical", "2020")
    } else {
      # Baseline 2020 + all other scenario years
      in_dir <- file.path(in_root, sc, as.character(yr))
    }
    
    cat("\n==========", sc, yr, "==========\n")
    # Produce the out path
    
    out_dir <- file.path(
      out_root, 
      sc, 
      as.character(yr)
      )
    
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
    
    tifs <- list.files(
      in_dir, 
      pattern="\\.tif$", 
      full.names=TRUE
      )
    
    for (tif in tifs){
      
      cat("Processing:", basename(tif), "\n")
      
      r <- rast(tif)
      NAflag(r) <- nodata_val
      fact <- target_deg / res(r)[1]
      
      # Aggregate - Upscale
      r_d <- aggregate(
        r, 
        fact = fact, 
        fun = mean, 
        na.rm = TRUE
        )
      
      out_file <- file.path(
        out_dir, 
        sub("\\.tif$", "_p05.tif",basename(tif))
        )
      
      # Write raster
      writeRaster(
        r_d, 
        out_file, 
        datatype  = "FLT4S", 
        overwrite=TRUE,
        gdal=c("COMPRESS=DEFLATE"))
      
      rm(r, r_d)
      gc()
    }
  }
}
```

### 7.1 Proccess historical 2020

In this step, the historical 2020 rasters are aggregated separately
because they provide the shared reference year for the *Nature for
Nature* (NfN), *Nature as Culture* (NaC), and *Nature for Society* (NfS)
scenarios.

Using a common 2020 baseline ensures that differences among these
scenarios in later years reflect the scenario assumptions rather than
different initial conditions.

```{r}

# Define input and output paths 
in_dir_hist  <- file.path(in_root, "Historical", "2020")
out_dir_hist <- file.path(out_root, "Historical", "2020")
dir.create(out_dir_hist, recursive = TRUE, showWarnings = FALSE)

tifs_hist <- list.files(
  in_dir_hist, 
  pattern="\\.tif$", 
  full.names=TRUE
  )

for (tif in tifs_hist) {

  cat("Processing Historical:", basename(tif), "\n")
  r <- rast(tif)
  NAflag(r) <- nodata_val
  
  fact <- target_deg / res(r)[1]
  
  # Aggregate - Upscale
  r_d <- aggregate(
    r, 
    fact = fact,
    fun = mean,
    na.rm = TRUE
    )

  out_file <- file.path(
    out_dir_hist, 
    sub("\\.tif$", "_p05.tif", basename(tif))
    )

  writeRaster(
    r_d,
    out_file,
    datatype  = "FLT4S",
    overwrite = TRUE,
    wopt      = list(gdal = c("COMPRESS=DEFLATE"))
    )
  rm(r, r_d)
  gc()
}
```

## 8. Create EBV Cube NetCDF file

### 8.1 Check paths and available binary rasters

Before creating the EBV Cube NetCDF file, we define the main input and
output paths for this step, including the metadata JSON file and the
location where the final NetCDF file will be written.

In addition, we check that the binary rasters created in the previous
step are available in the expected folder structure. These rasters
provide the binary entity layers that will later be inserted into the
EBV Cube.

Verifying their availability at this stage helps ensure that the NetCDF
creation step is based on the correct input files and that the workflow
can continue without missing-data errors.

```{r}

# Define date and check id 
creation_date <- 20260216
cat("Created date:", format(as.Date(as.character(creation_date), format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Define paths
root <- paste0('../json')
js <- paste0(root, "/metadata_", id, ".json")

# Note: 
# Please follow the naming convention 
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0('../cubes', '/alkemade_ecostr_id',id,'_',creation_date,'.nc')

cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Binary root folder: ", out_root, "\n")

# List all tiffs
tif_files <- list.files(
  out_root, 
  pattern='*\\.tif$', 
  recursive = TRUE, 
  full.names = TRUE
  )

cat("Total binary GeoTIFF files found:", length(tif_files), "\n")
```

### 8.2 Define EBV Cube structure

Before creating the NetCDF file, we first define its *core spatial*,
*temporal*, and *thematic components*. These include the list of
entities, the spatial extent and resolution, the coordinate reference
system (CRS), the time steps, and the fill value and data precision to
be used in the file.

In this dataset, the entities correspond to the harmonized broad
land-use classes derived from the original GLOBIO categories.

```{r}

cat("Entity variable\n")
cat("=======================\n")
cat("Entity Values: ", unique(globio_lut$broad_id), "\n")

# Entity names
entities <- unique(globio_lut$broader_category)

# Remove water bodies and permanent snow and ice
entities <- entities[!entities %in% c("Water bodies", "Permanent snow and ice")]

cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")


# Spatio-temporal setting
r0 <- rast(tif_files[1])         
ext <- as.vector(ext(r0))
res <- res(r0)
epsg <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = years, extent = ext, res = res, epsg = epsg)

# NoData handling 
nodata_val <- 255

# Fill value and datatype
fillvalue <- 255
prec <- 'float'
```

### 8.2 Create empty NetCDF file

After defining the main properties of the EBV Cube, an empty NetCDF file
is created. This file establishes the dataset’s structure, including its
spatial extent, resolution, coordinate reference system, time steps,
entities, and technical settings such as fill value and data precision.

```{r}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath = js,
  outputpath = nc,
  entities = entities,
  epsg = epsg,
  extent = extent,
  resolution = res_xy,
  timesteps = as.character(years),
  fillvalue = nodata_val,
  prec = prec,
  force_4D = TRUE,
  overwrite = TRUE,
  verbose = TRUE
)
```

At this stage, the file contains the EBV cube structure but not the
binary raster data, which will be added in the next step.

## 9. Populate the EBV Cube

Once the empty EBV NetCDF file has been created, the aggregated rasters
are inserted into the EBV Cube one by one.

For each scenario and time step, the workflow selects the correct folder
of aggregated entity rasters:

-   *Baseline 2020* uses the `Baseline/2020` folder

-   *all other scenarios in 2020* use the shared `Historical/2020`\
    folder

-   *future years* use the scenario-specific folders

Each `entity_XX_p05.tif` file is then inserted into the correct position
in the EBV Cube.

```{r}
# Get the properties and entities
props <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

# Define metric
metric_no <- 1
timesteps <- years

cat("\nEntities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {
  
  sc <- scenarios[s_no]
  
  for (t_no in seq_along(timesteps)) {
    
    yr <- timesteps[t_no]
    
    # Historical 2020 shared by all scenarios
    if (yr == 2020) {
        if (sc == "Baseline"){
            in_dir <- file.path(tif_root, "Baseline", "2020")
        } else {
            in_dir <- file.path(tif_root, "Historical", "2020")
        }
    } else {
        in_dir <- file.path(tif_root, sc, as.character(yr))
    }

    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")
    cat("Input dir:", in_dir, "\n")

    if (!dir.exists(in_dir)) {
      stop("Input directory does not exist: ", in_dir)
    }
  
    for (e_no in 1:n_entities) {

      tif_in <- file.path(
        in_dir, 
        sprintf("entity_%02d_p05.tif", e_no)
        )

      if (!file.exists(tif_in)) {
        stop("Missing entity file: ", tif_in)
      }

      ebv_add_data(
        filepath_nc = nc,
        data        = tif_in,
        scenario    = s_no,
        metric      = metric_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = TRUE
      )
      
      gc(verbose = FALSE)
      cat("  ✔ inserted entity", e_no, "\n")
    }
  }
}


# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

cat("\n🎉 DONE. All binary entities inserted into:\n", nc, "\n")
```
