---
title: "Working with EBVcube Data: A Practical Workflow Guide for BES-SIM 2"
author: "Andrés Mármol-Guijarro, Emmanuel Oceguera"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: 3
    highlight: tango
    number_sections: true
    latex_engine: xelatex   # handles unicode characters
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = TRUE,
                      message = FALSE,
                      warning = FALSE,
                      fig.width  = 8,
                      fig.height = 6,
                      dpi        = 150,
                      fig.align  = "center"   # centers all figures
)
```

------------------------------------------------------------------------

## Introduction

This guide walks you through the core workflow for accessing **EBVcube netCDF files** using the `ebvcube` R package. EBVcube is a standardised format for **Essential Biodiversity Variable (EBV)** data — gridded, multi-dimensional datasets tracking biodiversity change across space and time. A single `.nc` file can contain multiple data cubes organised hierarchically by scenario and metric, each with four dimensions: longitude, latitude, time, and entity (e.g., a species, land-use class, or ecological group). The datasets used throughout this guide are the **EBV cubes produced under BES-SIM 2**. For more information, visit the [ebvcube repository](https://github.com/EBVcube/ebvcube).

------------------------------------------------------------------------

## Required Packages

Restarting and running on a clean environment

```{r}
rm(list = ls())          
```

```{r libraries}
library(ebvcube)
library(terra)
library(geodata)
```

If BioConductor dependencies (`rhdf5`, `DelayedArray`, `HDF5Array`) are missing, install them via `BiocManager` before loading `ebvcube`.

```{r install, eval=FALSE}
#install.packages(c("ebvcube", "terra"))

#install.packages("BiocManager")
#BiocManager::install(c("rhdf5", "DelayedArray", "HDF5Array"))
```

------------------------------------------------------------------------

# 1. Downloading Cube Files from the NextCloud Repository

The BES-SIM 2 EBVcube `.nc` files are hosted on a shared NextCloud folder. Two folders have been created: `Data_geo` for downloading all input datasets and `BES_results` to upload results, reports and figures. Links and passwords have been circulated with the modelling teams. In case you have lost your password, please contact: a.marmol\@idiv.de.

------------------------------------------------------------------------

# 2. Setting Up Relative Paths

The notebook is designed to work with a mirrored folder structure. Anchoring the working directory to the script's location means all paths resolve correctly on any machine replicating the same layout — no absolute paths needed.

**Input data**

```         
└─ Data_geo\
   ├─ climate_data\
   │  ├─ ISIMIP3b\
   │  └─ WorldClim\
   ├─ land_use\
   │  ├─ GLOBIO\
   │  │  ├─ cubes\          # GLOBIO BES-SIM 2 EBVcube .nc files
   │  │  ├─ raw\
   │  │  └─ code\
   │  └─ PLUM\
   │     ├─ cubes\          # PLUM BES-SIM 2 EBVcube .nc files
   │     ├─ raw\
   │     └─ code\
   └─ protected_areas\
      ├─ GLOBIO-based_PAs\
      └─ PLUM-based_PAs\
```

**Outputs**

```         
└─ BES_results\
   ├─ EBVcube_ensamble_Tutorial\   # outputs from this notebook
   ├─ results_biodiversity\
   │   ├─ Model1\
   │   │   ├─ EBVcubes\
   │   │   ├─ figures\
   │   │   └─ reports\
   │   └─ Model2\
   │       ├─ EBVcubes\
   │       ├─ figures\
   │       └─ reports\
   └─ results_ecosystems\
       ├─ Model3\
       │   ├─ EBVcubes\
       │   ├─ figures\
       │   └─ reports\
       └─ Model4\
           ├─ EBVcubes\
           ├─ figures\
           └─ reports\
```

Setting relative paths to read input files

```{r paths}
# Anchor to script location (RStudio only)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
# Alternative for non-RStudio environments:
# setwd(getSrcDirectory(function() {})[1])

# EBVcube inputs
root_globio <- "../../Data_geo/land_use/GLOBIO/cubes/"   
root_plum   <- "../../Data_geo/land_use/PLUM/cubes/"     

# Protected areas
root_globiopa <- "../../Data_geo/protected_areas/GLOBIO-based_PAs/"
root_plumpa   <- "../../Data_geo/protected_areas/PLUM-based_PAs/"  
```

Setting relative paths for outputs (can be deleted if not relevant for this tutorial)

```{r paths outputs}
# Outputs — tutorial
root_tutorial <- "../../BES_results/EBVcube_ensamble_Tutorial/"

# Outputs — model folders (template: repeat for each model)
# Replace "results_biodiversity" / "results_ecosystems" and "Model1" as needed
root_m1 <- "../../BES_results/results_biodiversity/Model1/"
root_m1_cubes   <- file.path(root_m1, "EBVcubes/")
root_m1_figures <- file.path(root_m1, "figures/")
root_m1_reports <- file.path(root_m1, "reports/")
```

Once the paths are set, it is now possible to list all the land use cubes for GLOBIO or PLUM.

```{r discover-files}
nc_globio <- list.files(root_globio, pattern = "\\.nc$", full.names = TRUE, ignore.case = TRUE)
nc_plum   <- list.files(root_plum,   pattern = "\\.nc$", full.names = TRUE, ignore.case = TRUE)

nc_plumpa <- list.files(root_plumpa,   pattern = "\\.nc$", full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

nc_globiopa <- list.files(root_globiopa,   pattern = "\\.nc$", full.names = TRUE, recursive = TRUE, ignore.case = TRUE)

cat("GLOBIO cubes:\n"); print(nc_globio)
cat("PLUM cubes:\n");   print(nc_plum)
cat("GLOBIO protected areas cubes:\n"); print(nc_globiopa)
cat("PLUM protected areas cubes:\n");   print(nc_plumpa)
```

Files within each vector are referenced by index throughout (`nc_globio[1]`, `nc_plum[2]`, etc.). Adjust indices to match your local file inventory. The following table provides an index to all the input data cubes, which are indexed in the code box below.

| Filepath | Data cube name | Short name |
|----------------------------|-------------------------|-------------------|
| `alkemade_ecostr_id103_20260216.nc` | Upscaled 0.5 Land Use projections (BES-SIM 2 GLOBIO) | `globio_lu_05` |
| `alkemade_ecostr_id98_20260223.nc"` | Land Use projection (BES-SIM 2 GLOBIO) | `globio_lu` |
| `alexander_ecostr_id99_20260217.nc"` | Land Use projection (BES-SIM 2 LandSymm-PLUM) | `plum_lu` |
| `alexander_ecostr_id100_20260204.nc"` | Forest management (BES-SIM 2 LandSyMM-PLUM) | `plum_frstmng` |
| `alexander_croscut_id101_20260202.nc"` | Crop areas and intensities (BES-SIM 2 LandSyMM-PLUM) | `plum_crop_int` |
| `alexander_ecostr_id102_20260309.nc"` | Native 0.5 Land Use projection (BES-SIM 2 LandSymm-PLUM) | `plum_lu_05` |
| `alkemade_ecostr_idTBD_20260309.nc` | GLOBIO Protected area layers | `globio_pas` |
| `alexander_ecostr_idTBD_20260306.nc` | LandSyMM-PLUM Protected area layers | `plum_pas` |

```{r indexing-inpun-data-cubes}
dc <- list(
  globio_lu_05  = "../../Data_geo/land_use/GLOBIO/cubes//alkemade_ecostr_id103_20260216.nc",
  globio_lu     = "../../Data_geo/land_use/GLOBIO/cubes//alkemade_ecostr_id98_20260223.nc",
  plum_lu       = "../../Data_geo/land_use/PLUM/cubes//alexander_ecostr_id99_20260217.nc",
  plum_frstmng  = "../../Data_geo/land_use/PLUM/cubes//alexander_ecostr_id100_20260204.nc",
  plum_crop_int = "../../Data_geo/land_use/PLUM/cubes//alexander_croscut_id101_20260202.nc",
  plum_lu_05    = "../../Data_geo/land_use/PLUM/cubes//alexander_ecostr_id102_20260309.nc",
  globio_pas    = "../../Data_geo/protected_areas/GLOBIO-based_PAs//cubes/alkemade_ecostr_idTBD_20260309.nc",
  plum_pas      = "../../Data_geo/protected_areas/PLUM-based_PAs//cubes/alexander_ecostr_idTBD_20260306.nc"
)
```

Then call any cube using `dc$data_cube_name` (e.g. `dc$plum_lu`)

------------------------------------------------------------------------

# 3. Accessing the Data Cubes

`ebv_properties()` returns an S4 object with all file-level metadata. The key slots for exploration are `general`, `spatial`, and `temporal`.

Start by loading the properties of one file and inspecting the available slots:

```{r file-properties}
globio05_dc <- ebv_properties(dc$globio_lu_05[1], verbose = FALSE)
slotNames(globio05_dc)
```

The `general` slot contains descriptive metadata — the title, a description of the dataset, and the list of entity names. Entity names are important because their index position maps directly to the `entity` argument in `ebv_read*` functions:

```{r general-metadata}
globio05_dc@general$title
globio05_dc@general$description
globio05_dc@general$entity_names   # entity index maps directly to the entity argument in ebv_read*
```

The `spatial` slot reports the coordinate reference system, cell resolution, and bounding extent of the grid:

```{r spatial-metadata}
cat("EPSG:      ", globio05_dc@spatial$epsg,       "\n")
cat("Resolution:", globio05_dc@spatial$resolution, "\n")
cat("Extent:    ", globio05_dc@spatial$extent,     "\n")
```

The `temporal` slot lists all available time steps as dates. This tells you the temporal coverage of the cube and helps you identify which `timestep` index to pass when reading a layer:

```{r temporal-metadata}
cat(as.character(globio05_dc@temporal$dates), sep = "\n")
```

**Note:** Not all EBV cubes have time steps. The protected area cubes for GLOBIO and LandSyMM-PLUM, for instance, contain only one mask layer per scenario with no associated year. This is an example of the latter.

```{r PA-temporal metadata}
# getting LandSyMM-PLUM propertioes 
plum_pas <- ebv_properties(dc$plum_pas[1], verbose = FALSE)

# checking is temporal metadata
cat(as.character(plum_pas@temporal$dates), sep = "\n")
```

The time step within it shows `2000-01-01`, which has been set as default as a placeholder for the temporal slot by the `ebvcube` R-package.

------------------------------------------------------------------------

# 4. Accessing Layers Within the Cube

The internal hierarchy is **Scenario → Metric → Entity** (scenarios are optional). `ebv_datacubepaths()` resolves this and returns a data frame of all valid paths, with descriptions of the metric and scenario each path represents.

```{r datacube-paths}
globio05_paths <- ebv_datacubepaths(dc$globio_lu_05, verbose = FALSE)
print(globio05_paths)
```

In this file there are four scenarios — *Nature for Nature*, *Nature as Culture*, *Nature for Society*, and *SSP2-Baseline* — each containing the same metric (Land cover). The `datacubepaths` column gives the internal HDF5 path used to locate a specific cube within the file. This path is passed to subsequent functions to specify which cube to work with.

In the example, `datacubepath = globio05_paths[1, 1]` selects the first scenario — **"Nature for Nature"** — which is then used to retrieve the cube's properties via `ebv_properties()`. Alternatively, **"Nature as Culture"** could be selected using `datacubepath = globio05_paths[2, 1]`.

The `@metric` and `@ebv_cube$type` slots provide additional metadata about the selected datacube, describing the metric ("Land cover") and the data type used to store the values, respectively.

```{r select-path}
dc_path   <- globio05_paths[1, 1]
globio05_dc <- ebv_properties(dc$globio_lu_05[1], dc_path, verbose = FALSE)

# description of the metric
globio05_dc@metric

# description of the data type (i.e. float, interger, boolean)
globio05_dc@ebv_cube$type
```

## 4.1 Plotting layers within a EBVcube

`ebv_read()` loads a specific entity and timestep as a `SpatRaster` (`type = "r"`) or array (`type = "a"`). At this point it is advisable to recall all the entities within each cube using `globio@general$entity_names`.

```{r recalling-entities}
globio05_dc@general$entity_names
```

The output indicates the numbers with their corresponding names within `globio_lu_05` ebvcube. With this information it is now simple to choose a desired entity to plot. In this example, `entity = 9` ("Unmanaged forest") and `timestep = 1` ( corresponding to the first available time step in the dataset that in this case is "2020") have been selected. This cube will display projections made for "Nature for Nature" scenarios as specified by `datacubepath = dc_path` (see above). These parameters could be changed to query other entities and timesteps accordingly.

```{r read-layer, eval=FALSE}
start_time <- Sys.time()

data_layer <- ebv_read(
  filepath     = dc$globio_lu_05,
  datacubepath = dc_path,              # this is selecting Nature for Nature projections.
  entity       = 5,
  timestep     = 1,
  ignore_RAM   = TRUE
)

cat("Read time:", round(Sys.time() - start_time, 2), "seconds\n")
plot(data_layer)
```

Alternatively, all the available time steps can be plotted using `timestep = 1:4` .

```{r read-layers, eval = FALSE}
start_time <- Sys.time()

data_layer <- ebv_read(
  filepath     = dc$globio_lu_05[1],
  datacubepath = dc_path,
  entity       = 5,
  timestep     = 1:4,
  ignore_RAM   = TRUE
)

cat("Read time:", round(Sys.time() - start_time, 2), "seconds\n")
plot(data_layer)
```

In this case, `lyr.1`, `lyr.2`, `lyr.3` and `lyr.4` correspond to 2020, 2030, 2050 and 2070, respectively.

**NOTE:** For global 10 arc-second files, always use `ignore_RAM = TRUE` together with a spatial subset (Section 5) — loading the full extent is not feasible on standard workstations.

------------------------------------------------------------------------

# 5. Subsetting High-Resolution Cubes (10 arc-seconds) by Region

At 10 arc-second resolution, a global layer contains approximately 1.3 billion cells, which can exceed the available RAM. The package provides two spatial subsetting functions that extract only the requested region directly from the file on disk, avoiding the need to load the full global extent into memory.

| Function | Input | Notes |
|------------------------|------------------------|------------------------|
| `ebv_read_bb()` | Bounding box `c(xmin, xmax, ymin, ymax)` | Fast rectangular clip |
| `ebv_read_shp()` | Path to a `.shp` file | Clips to boundary, masks exterior cells to `NA` |

## 5.1 Bounding Box — EU Extent

For the following exercise, we now switch to read `dc$globio_lu`. The bounding box below covers mainland EU in EPSG 4326.

For this is necessary to redefine the datacube paths by reading `dc$globio_lu`

```{r reading datacube paths for globio_lu}

globio_paths <- ebv_datacubepaths(dc$globio_lu, verbose = FALSE)
print(globio_paths)
```

Selecting "Nature for Society" by using `globio_paths[3, 1]` to plot.

```{r}
dc_path   <- globio_paths[3, 1]
globio_dc <- ebv_properties(dc$globio_lu, dc_path, verbose = FALSE)
```

Revealing entities within the cube.

```{r}
globio_dc@general$entity_names
```

`entity = 8` "Forestry" for all time steps is plotted

```{r read-bb-eu, eval=FALSE}
br_bb <- c(-74, -28, -34, 6)   # xmin, xmax, ymin, ymax

br_subset <- ebv_read_bb(
  filepath     = dc$globio_lu,
  datacubepath = dc_path,
  entity       = 8,
  timestep     = 1:4,
  bb           = br_bb,
  ignore_RAM   = TRUE,
  verbose      = TRUE
)

plot(br_subset)
```

## 5.2 Shapefile Subset — EU Administrative Boundary

Subsetting the dataset using shapefiles, for instance of countries, is also possible. For this example, the dataset will be subset to be plotted within Brazilian boundaries. Shapefiles are downloaded using `geodata` package.

```{r r download-shp-eu, eval=FALSE}

#calling geodata package
library(geodata)

#setting tempdir within environment
temp_dir  = '../../Temp'
dir.create(temp_dir, showWarnings = FALSE, recursive = TRUE)

# Download Brazil level 0 boundary
brazil_shp <- geodata::gadm(country = "BRA", level = 0, path = temp_dir)

# Write to shapefile in temp_dir (required by ebv_read_shp)
brazil_shp_path <- file.path(temp_dir, "brazil.shp")
terra::writeVector(brazil_shp, brazil_shp_path, overwrite = TRUE)
```

Once donwloaded, Brazil shapefile is then indicated in `shp` parameter of the `ebv_read_shp()` function.

```{r read-shp-eu, eval=FALSE}
globio_br <- ebv_read_shp(
  filepath     = dc$globio_lu,
  datacubepath = dc_path,
  entity       = 8,                        # Forestry
  timestep     = 1:4,
  shp          = brazil_shp_path,
  ignore_RAM   = TRUE,
  verbose      = TRUE
)

plot(globio_br)
```

------------------------------------------------------------------------

# 6. Saving GeoTIFF Files

## 6.1 From a loaded SpatRaster

The `ebv_write()` function exports a subset or processed EBV object to disk as a GeoTIFF file. It accepts the in-memory data object and a specified output path. The resulting tiff will be written in the folder that was recently created for saving Brazil shapefile.

**Note:** the type of data selected to save the dataset matters for the resulting size of the final tif. Selection will depend on the metric that needs to be written as output. For instance, globio_lu has only binary land cover type, meaning that 16-bit INT2S would work. FLT8S (64-bit) is recommended for biodiversity indices and modelled outputs requiring maximum numerical precision.

The table below summarizes the available GeoTIFF data types supported by the `type` argument:

| Code | Type | Bits | Range | Use Case |
|---------------|---------------|---------------|---------------|---------------|
| INT1S | Signed integer | 8-bit | -128 to 127 | Compact categorical data |
| INT2S | Signed integer | 16-bit | -32,768 to 32,767 | Elevation, temperature anomalies |
| INT2U | Unsigned integer | 16-bit | 0 to 65,535 | Land cover classes, counts |
| INT4S | Signed integer | 32-bit | -2.1B to 2.1B | Large integer values |
| INT4U | Unsigned integer | 32-bit | 0 to 4.3B | Large positive integer values |
| FLT4S | Signed float | 32-bit | \~±3.4×10^38^ | Continuous data, moderate precision |
| **FLT8S** | **Signed float** | **64-bit** | **\~±1.8×10^308^** | **High precision continuous data** |

```{r write-ebvwrite, eval=FALSE}
ebv_write(
  data       = br_subset,
  outputpath = file.path(temp_dir, "br_entity9_t1.tif"),  # fixed filename
  overwrite  = TRUE,
  verbose    = TRUE,
  type       = "INT2S"
)
```

## 6.2 Direct write during read (preferred for large subsets)

Passing `outputpath` to `ebv_read_bb()` or `ebv_read_shp()` writes directly to disk and returns the path rather than a `SpatRaster`, avoiding loading into RAM. However, the downside of writing directly through the `ebv_read_shp()` is that writing data `type` cannot be parametrised, which may render larger tif files.

```{r write-direct, eval=FALSE}
ebv_read_shp(
  filepath     = dc$globio_lu,
  datacubepath = globio_paths[1,1], #this is Nature for Nature scenario
  entity       = 9,
  timestep     = 1:4,
  shp          = brazil_shp_path,
  outputpath   = file.path(temp_dir, "br_entity9_t1.tif"),
  overwrite    = TRUE,
  ignore_RAM   = TRUE
)
```

## 6.3 Via `terra::writeRaster()` for format control

Use this when specific datatype, compression, or tiling is required. Match `datatype` to `globio_dc@ebv_cube$type` to avoid precision loss or unnecessary file size.

```{r write-terra, eval=FALSE}
terra::writeRaster(
  x         = br_subset,
  filename  = file.path(temp_dir, "br_entity9_t1.tif"),
  filetype  = "GTiff",
  datatype  = "INT2S",   # or INT2S / INT4U depending on the cube's data type
  overwrite = TRUE,
  gdal      = c("COMPRESS=DEFLATE")
  
)
```

------------------------------------------------------------------------

# Quick Reference

| Step | Function | Notes |
|------------------------|------------------------|------------------------|
| File metadata | `ebv_properties(filepath)` | Returns S4 object; key slots: `general`, `spatial`, `temporal` |
| List cubes | `ebv_datacubepaths(filepath)` | One row per scenario/metric combination |
| Cube metadata | `ebv_properties(filepath, datacubepath)` | Metric description and data type |
| Read full layer | `ebv_read()` | Use only for coarser resolutions |
| Bounding box subset | `ebv_read_bb()` | Essential for 10 arc-second data |
| Shapefile subset | `ebv_read_shp()` | Masks exterior cells to `NA` |
| Save (ebvcube) | `ebv_write()` | Straightforward write from SpatRaster |
| Save (terra) | `terra::writeRaster()` | Full format and compression control |

------------------------------------------------------------------------

## Further Resources

-   [ebvcube GitHub](https://github.com/EBVcube/ebvcube) — source code and issue tracker
-   [ebvcube CRAN](https://cran.r-project.org/package=ebvcube) — stable release and function reference
-   [EBV Data Portal](https://portal.geobon.org) — GEO BON dataset catalogue
-   [EBVCube format specification](https://github.com/EBVcube/EBVCube-format)
-   [terra documentation](https://rspatial.github.io/terra/)
