---
title: "PLUM Land-Use – Rasterization to 0.5° and EBV Cube Generation"
author: "Emmanuel Oceguera-Conchas"
date: 2026-02-05
output: html_document
---

## 1. Introduction

This workflow processes the PLUM 0.5° land-use data and converts it into an EBV Cube NetCDF file (dataset ID 102).

Each PLUM table lists, per 0.5° cell, the area of every land-use class split by protection status (`protected` / `unprotected`). For each class the workflow computes three percentage metrics — protected, unprotected, and total — relative to the cell's combined land area, stacks the years into one raster per class, and stores the result as an EBV Cube.

The workflow comprises two core steps:

-   *Rasterization:* Each class area is divided by the cell's combined land area (all classes, protected + unprotected) and multiplied by 100, giving the protected, unprotected, and total percentage; each metric is written into the 0.5° grid as one stacked GeoTIFF per class (one band per year).

-   *EBV Cube NetCDF file generation:* The stacked layers are structured into an EBV Cube, organizing the data by scenario, metric, time step, and entity (the land-use class).

In the notebook, these steps span four sections: preparing the working environment (loading `terra`, `ebvcube` and setting paths), rasterizing and stacking the metrics, building the EBV Cube structure, and populating it.

### 1.1 PLUM scenarios and time steps

The workflow processes five scenarios across four years each. The scenario order follows the metadata JSON.

| Scenario                  | Abbreviation | Available Years        |
|---------------------------|--------------|------------------------|
| Nature for Nature         | NfN          | 2020, 2030, 2050, 2070 |
| Nature for Nature (local) | NfNl         | 2020, 2030, 2050, 2070 |
| Nature as Culture         | NaC          | 2020, 2030, 2050, 2070 |
| Nature for Society        | NfS          | 2020, 2030, 2050, 2070 |
| Business as usual         | BAU          | 2020, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   The PLUM scenario ZIP archives (tabular `.txt.gz` files), in the expected folder structure `SCENARIO/s1/YEAR/LandCover.txt.gz`
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   Enough temporary disk space for raster processing
-   A working `ebvcube` install, plus `terra`

```{r}
knitr::opts_chunk$set(cache = TRUE)
```

## 2. Load libraries

Here we load the packages used throughout the workflow: `terra` for building and writing the stacked GeoTIFFs, and `ebvcube` for creating and filling the EBV Cube NetCDF file.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")

library(terra)
library(ebvcube)
```

## 3. Configure temporary raster storage

Rasterization can write a lot of temporary data to disk, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp 
)

cat("Local temporary folder: ", local_temp)
```

## 4. Define input and output paths

Input: the folder of PLUM ZIP archives. Output: the folder for the stacked per-class GeoTIFFs, organized by metric and scenario.

```{r}
base_dir <- "../BES-SIM 2/PLUM_LU/0.5_degrees"
out_dir  <- "../BES-SIM 2/PLUM_LU/0.5_degrees/disgregated_rast"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

cat("Input root:", base_dir, "\n")
cat("Output root:", out_dir, "\n")
```

## 5. Define the main workflow settings

Set the scenarios, years, metrics, and the land-use classes to process.

```{r}
# Scenario identifiers (must match the ZIP internal paths AND the metadata JSON order)
scenarios <- c("NfN_NfN", "NfNl_NfNl", "NaC_NaC", "NfS_NfS", "BAU")

# Years and metrics (metric order must match the metadata JSON: 1 protected, 2 unprotected, 3 total)
target_years <- c("2020", "2030", "2050", "2070")
metrics      <- c("protected", "unprotected", "total")

# Folder level inside each ZIP, and the variable
level_4 <- "s1"
var_4   <- "LandCover"

# Land-use classes (entity order must match the metadata JSON)
landcover_cols <- c(
  "Cropland", "Pasture", "TimberForest", "CarbonForest", "UnmanagedForest",
  "OtherNatural", "Agrivoltaics", "Photovoltaics", "Barren", "Urban"
)

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Years:", paste(target_years, collapse = ", "), "\n")
cat("Metrics:", paste(metrics, collapse = ", "), "\n")
```

## 6. Build the 0.5° raster template

```{r}
zip_files <- list.files(base_dir, pattern = "\\.zip$", full.names = TRUE)
stopifnot(length(zip_files) > 0)

cat("Zip files found:\n", paste(zip_files, collapse = "\n"), "\n")

r_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = 0.5,
  crs = "EPSG:4326"
)

# Half-cell offset for converting lower-left coordinates to cell centres
half_res <- res(r_template) / 2
half_res
```

## 7. Rasterize and stack

For each scenario, read the yearly tables, compute the three percentage metrics per cell, and write one stacked GeoTIFF per metric and class (bands = years).

```{r}
start_time <- Sys.time()

ny   <- length(target_years)
ncls <- length(landcover_cols)

for (scen in scenarios) {

  scen_zip <- grep(scen, zip_files, value = TRUE)
  if (length(scen_zip) == 0L) {
    warning("No zip found for scenario: ", scen)
    next
  }
  scen_zip <- scen_zip[1]

  cat("\n=============================\n")
  cat("Scenario:", scen, "\n")

  # Per-year percentage matrices (ncell x class), one list per metric
  prot_years   <- vector("list", ny)
  unprot_years <- vector("list", ny)
  total_years  <- vector("list", ny)

  for (y in seq_along(target_years)) {

    yr <- target_years[y]
    rel_path <- sprintf("%s/%s/%s/%s.txt.gz", scen, level_4, yr, var_4)

    # Extract and read the gzipped table
    tmp_dir <- local_tmp
    unzip(scen_zip, files = rel_path, exdir = tmp_dir)

    txt_gz  <- file.path(tmp_dir, rel_path)
    con <- gzfile(txt_gz, "rt")
    dat <- read.table(con, header = TRUE, check.names = FALSE)
    # View(dat)
    close(con)
    unlink(txt_gz)

    cat("  year", yr, "\n")

    stopifnot(all(c("Lon", "Lat", "Protection") %in% names(dat)))
    stopifnot(all(landcover_cols %in% names(dat)))

    # Cell index from the cell centre; drop rows outside the grid
    dat$cell <- cellFromXY(r_template, cbind(dat$Lon + half_res[1],
                                             dat$Lat + half_res[2]))
    dat <- dat[!is.na(dat$cell), ]

    # Align protected and unprotected areas per cell
    # (one protected and one unprotected row per cell)
    cells_all <- sort(unique(dat$cell))
    P <- matrix(0, length(cells_all), ncls, dimnames = list(NULL, landcover_cols))
    U <- P

    pr <- dat[dat$Protection == "protected", ]
    un <- dat[dat$Protection == "unprotected", ]
    P[match(pr$cell, cells_all), ] <- as.matrix(pr[, landcover_cols])
    U[match(un$cell, cells_all), ] <- as.matrix(un[, landcover_cols])
    Tot <- P + U

    # Denominator: combined land area per cell (all classes, protected + unprotected).
    # To divide each metric by its own protection subset instead, change this line.
    denom <- rowSums(Tot)
    denom[denom == 0] <- NA

    P   <- 100 * P   / denom
    U   <- 100 * U   / denom
    Tot <- 100 * Tot / denom

    # Place the per-cell values into full-grid matrices (ncell x class)
    full_P <- matrix(NA_real_, ncell(r_template), ncls)
    full_U <- matrix(NA_real_, ncell(r_template), ncls)
    full_T <- matrix(NA_real_, ncell(r_template), ncls)
    full_P[cells_all, ] <- P
    full_U[cells_all, ] <- U
    full_T[cells_all, ] <- Tot

    prot_years[[y]]   <- full_P
    unprot_years[[y]] <- full_U
    total_years[[y]]  <- full_T

    rm(dat, pr, un, P, U, Tot, full_P, full_U, full_T)
    gc()
  }

  # Write one stacked GeoTIFF per metric and class (bands = years)
  for (metric in metrics) {

    metric_years <- switch(metric,
      protected   = prot_years,
      unprotected = unprot_years,
      total       = total_years
    )

    for (k in seq_along(landcover_cols)) {

      cls <- landcover_cols[k]

      # Assemble the year bands for this class
      vmat <- matrix(NA_real_, ncell(r_template), ny)
      for (y in seq_along(target_years)) vmat[, y] <- metric_years[[y]][, k]

      r_stack <- rast(r_template, nlyrs = ny)
      values(r_stack) <- vmat
      names(r_stack)  <- target_years

      out_file <- file.path(out_dir, metric, scen, level_4,
                            sprintf("PLUM_%s_%s_%s.tif", metric, scen, cls))
      dir.create(dirname(out_file), recursive = TRUE, showWarnings = FALSE)

      writeRaster(
        r_stack, out_file,
        datatype  = "FLT4S",
        overwrite = TRUE,
        gdal = c("COMPRESS=DEFLATE")
      )
    }
    cat("  wrote metric:", metric, "\n")
  }

  rm(prot_years, unprot_years, total_years)
  gc()
}

end_time <- Sys.time()
cat("\nRasterization completed in ",
    round(difftime(end_time, start_time, units = "mins"), 2), " minutes.\n")
```

## 8. Create EBV Cube NetCDF file

### 8.1 Check paths and available rasters

Define the metadata JSON and the target NetCDF file, and confirm the stacked GeoTIFFs from Section 7 are present in the expected folders. This check catches missing inputs before the cube is built.

```{r, eval=FALSE}
# Define date and check id
id <- 102
creation_date <- "20260626"
cat("Created date:", format(as.Date(creation_date, format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Define paths
root <- paste0("../ebvportal/upload/", id)
js   <- file.path(root, "metadata.json")

# Note:
# Please follow the naming convention
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0("../ebv_portal/", id, "/alexander_ecostr_id", id, "_", creation_date, ".nc")

cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Rasterized root folder:", out_dir, "\n")

# List all stacked GeoTIFFs
tif_files <- list.files(out_dir, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)

cat("Total stacked GeoTIFF files found:", length(tif_files), "\n")
```

### 8.2 Define EBV Cube structure

Define the cube's spatial, temporal, and thematic components: entities, extent and resolution, CRS, time steps, fill value, and data precision. Entities here are the land-use classes, and the values are continuous percentages (0–100).

```{r, }

# Define entities
entities <- landcover_cols
cat("Entity variable\n")
cat("=======================\n")
cat("Entity names: ", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")

# Spatio-temporal setting
r0     <- rast(tif_files[1])
extent <- as.vector(ext(r0))
res_xy <- res(r0)
epsg   <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = target_years, extent = extent, res = res_xy, epsg = epsg)

# Fill value and datatype
fillvalue <- -1
prec      <- "float"
```

### 8.3 Create empty NetCDF file

Create the empty NetCDF file from the properties defined above. It holds the cube structure but no raster data yet — that comes in Section 9.

```{r, eval=FALSE}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = extent,
  resolution = res_xy,
  timesteps  = target_years,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

cat("EBV cube created at:", nc, "\n")
```

## 9. Populate the EBV Cube

Insert the stacked rasters into the cube one by one. For each scenario, metric, and class, the workflow reads the stacked GeoTIFF and inserts each year band at its correct scenario / metric / time step / entity position.

```{r}
# Get the properties and entities
props      <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

cat("Entities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (m_no in seq_along(metrics)) {

    metric <- metrics[m_no]

    for (e_no in seq_len(n_entities)) {

      cls <- entities[e_no]

      in_file <- file.path(out_dir, metric, sc, level_4,
                           sprintf("PLUM_%s_%s_%s.tif", metric, sc, cls))

      if (!file.exists(in_file)) {
        stop("Missing entity file: ", in_file)
      }

      # One band per year in the stacked file
      for (t_no in seq_along(target_years)) {

        ebv_add_data(
          filepath_nc = nc,
          data        = in_file,
          band        = t_no,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          ignore_RAM  = FALSE,
          raw         = FALSE,
          verbose     = FALSE
        )
      }

      gc(verbose = FALSE)
      cat("  ✔ inserted", metric, "|", sc, "|", cls, "\n")
    }
  }
}

# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

cat("\n DONE. All layers inserted into:\n", nc, "\n")
```

## 10. Validation and visualization

A quick check that the cube is readable and that the metrics, entities, and time steps line up.

```{r}
ebv_analyse(nc, entity = 'UnmanagedForest', timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@ebv_cube$type
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 'UnmanagedForest',
        timestep = 2, verbose = FALSE)
```

## 11. Clean up temporary extraction files

With everything inserted, the temporary extraction folder can be removed to free disk space.

```{r}
cat("Removing temporary extraction folder:", local_tmp, "\n")
unlink(local_tmp, recursive = TRUE)

if (!dir.exists(local_tmp)) {
  cat("✔ Temporary extraction folder successfully removed.\n")
} else {
  warning("Temporary folder could not be removed: ", local_tmp)
}
```
