---
title: "PLUM Forest Management – Rasterization to 0.5° and EBV Cube Generation"
author: "Emmanuel Oceguera-Conchas, Andres Marmol-Guijarro, Magali Weissgerber"
date: 2026-02-05
output: html_document
---

NEEDS TO BE CHECK.

## 1. Introduction

This workflow processes the PLUM 0.5° forest-management data and converts it into an EBV Cube NetCDF file (dataset ID 100).

Each PLUM table reports three forest-management variables per 0.5° cell — rotation intensity, timber-forest area, and yield — split by protection status (`protected` / `unprotected`). The workflow rasterizes each variable separately for protected and unprotected forest, stacks the years into one raster per variable and protection, and stores the result as an EBV Cube. Protected and unprotected forest are the two EBV entities; the three variables are the metrics.

The workflow comprises two core steps:

-   *Rasterization:* Each forest-management variable is rasterized into the 0.5° grid separately for protected and unprotected forest, and stacked into one GeoTIFF per variable and protection (one band per year).
-   *EBV Cube NetCDF file generation:* The stacked layers are structured into an EBV Cube, organizing the data by scenario, metric (the three variables), time step, and entity (protected / unprotected forest).

In the notebook, these steps span four sections: preparing the working environment (loading `terra`, `ebvcube` and setting paths), rasterizing and stacking the variables, building the EBV Cube structure, and populating it.

### 1.1 PLUM scenarios and time steps

The workflow processes five scenarios across four years each. The scenario order follows the metadata JSON. Forest management is allowed in protected areas only under NaC and NfS; for NfN, NfNl, and BAU the protected entity contains no data.

| Scenario                  | Abbreviation | Available Years        |
|---------------------------|--------------|------------------------|
| Nature for Nature         | NfN          | 2020, 2030, 2050, 2070 |
| Nature for Nature (local) | NfNl         | 2020, 2030, 2050, 2070 |
| Nature as Culture         | NaC          | 2020, 2030, 2050, 2070 |
| Nature for Society        | NfS          | 2020, 2030, 2050, 2070 |
| Business as usual         | BAU          | 2020, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   The PLUM scenario ZIP archives (tabular `.txt.gz` files), in the expected folder structure `SCENARIO/s1/YEAR/Forestry.txt.gz`
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   Enough temporary disk space for raster processing
-   A working `ebvcube` install, plus `terra`

```{r}
knitr::opts_chunk$set(cache = TRUE)
```

## 2. Load libraries

Here we load the packages used throughout the workflow: `terra` for building and writing the stacked GeoTIFFs, and `ebvcube` for creating and filling the EBV Cube NetCDF file.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")

library(terra)
library(ebvcube)
```

## 3. Configure temporary raster storage

Rasterization can write a lot of temporary data to disk, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths

Input: the folder of PLUM ZIP archives. Output: the folder for the stacked per-variable GeoTIFFs, organized by metric and scenario.

```{r}
base_dir <- "../BES-SIM 2/PLUM_LU/0.5_degrees"
out_dir  <- "../BES-SIM 2/PLUM_LU/0.5_degrees/forestry_rast"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

cat("Input root:", base_dir, "\n")
cat("Output root:", out_dir, "\n")
```

## 5. Define the main workflow settings

Set the scenarios, years, metrics, and entities (protection status) to process.

```{r}
# Scenario identifiers (must match the ZIP internal paths AND the metadata JSON order)
scenarios <- c("NfN_NfN", "NfNl_NfNl", "NaC_NaC", "NfS_NfS", "BAU")

# Years (band order in each stack)
target_years <- c("2020", "2030", "2050", "2070")

# Metrics (order must match the metadata JSON: 1 RotationIntensity, 2 TimberForestArea, 3 Yield).
# These are also the value columns expected in the tables.
metrics <- c("RotationIntensity", "TimberForestArea", "Yield")

# Entities = protection status. entity_levels are the values in the table's Protection
# column; entities are the matching EBV entity names, in the same order.
entity_levels <- c("unprotected", "protected")
entities      <- c("Forestry unprotected", "Forestry protected")

# Folder level inside each ZIP, and the variable
level_4 <- "s1"
var_4   <- "Forestry"

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Years:", paste(target_years, collapse = ", "), "\n")
cat("Metrics:", paste(metrics, collapse = ", "), "\n")
cat("Entities:", paste(entities, collapse = ", "), "\n")
```

## 6. Build the 0.5° raster template

```{r}
zip_files <- list.files(base_dir, pattern = "\\.zip$", full.names = TRUE)
stopifnot(length(zip_files) > 0)

cat("Zip files found:\n", paste(zip_files, collapse = "\n"), "\n")

r_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = 0.5,
  crs = "EPSG:4326"
)

# Half-cell offset for converting lower-left coordinates to cell centres
half_res <- res(r_template) / 2
half_res
```

## 7. Rasterize and stack

For each scenario, read the yearly tables, split each into protected and unprotected forest, and write one stacked GeoTIFF per metric and protection (bands = years). Where a scenario has no protected forest, its protected stack is written as all-NA so the cube structure stays consistent.

```{r, eval=FALSE}
start_time <- Sys.time()

ny     <- length(target_years)
n_cell <- ncell(r_template)

for (scen in scenarios) {

  scen_zip <- grep(scen, zip_files, value = TRUE)
  if (length(scen_zip) == 0L) {
    warning("No zip found for scenario: ", scen)
    next
  }
  scen_zip <- scen_zip[1]

  cat("\n=============================\n")
  cat("Scenario:", scen, "\n")

  # Storage: per protection, per metric, a list of one full-grid vector per year
  vals <- list()
  for (prot in entity_levels) {
    vals[[prot]] <- list()
    for (m in metrics) vals[[prot]][[m]] <- vector("list", ny)
  }

  for (y in seq_along(target_years)) {

    yr <- target_years[y]
    rel_path <- sprintf("%s/%s/%s/%s.txt.gz", scen, level_4, yr, var_4)

    # Extract flat (strip the archive's folders), then read
    tmp_dir <- file.path(local_tmp, scen, level_4, yr)
    dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)
    unzip(scen_zip, files = rel_path, exdir = tmp_dir, junkpaths = TRUE)

    txt_gz <- file.path(tmp_dir, basename(rel_path))
    con <- gzfile(txt_gz, "rt")
    dat <- read.table(con, header = TRUE, check.names = FALSE)
    close(con)
    unlink(txt_gz)

    cat("  year", yr, "\n")

    stopifnot(all(c("Lon", "Lat", "Protection") %in% names(dat)))
    stopifnot(all(metrics %in% names(dat)))

    # Cell index from the cell centre; drop rows outside the grid
    dat$cell <- cellFromXY(r_template, cbind(dat$Lon + half_res[1],
                                             dat$Lat + half_res[2]))
    dat <- dat[!is.na(dat$cell), ]

    # Split by protection, then place each metric's values onto the full grid
    for (prot in entity_levels) {                  # "unprotected", then "protected"
      
      sub <- dat[dat$Protection == prot, ]         # Keep only rows of this proctection
      
      for (m in metrics) {                         # each of the 3 variables
        full <- rep(NA_real_, n_cell)              # a fresh all-NA vector, one slot per cell
        if (nrow(sub) > 0) full[sub$cell] <- sub[[m]]
        vals[[prot]][[m]][[y]] <- full             # park it in the right pigeonhole
      }
    }

    rm(dat)
    gc()
  }

  # Write one stacked GeoTIFF per metric and protection (bands = years)
  for (prot in entity_levels) {
    for (m in metrics) {

      vmat <- matrix(NA_real_, n_cell, ny)
      for (y in seq_along(target_years)) vmat[, y] <- vals[[prot]][[m]][[y]]

      r_stack <- rast(r_template, nlyrs = ny)
      values(r_stack) <- vmat
      names(r_stack)  <- target_years

      out_file <- file.path(out_dir, m, scen, level_4,
                            sprintf("Forestry_%s_%s_%s.tif", m, scen, prot))
      dir.create(dirname(out_file), recursive = TRUE, showWarnings = FALSE)

      writeRaster(
        r_stack, out_file,
        datatype  = "FLT4S",
        overwrite = TRUE,
        gdal = c("COMPRESS=DEFLATE")
      )
    }
    cat("  wrote protection:", prot, "\n")
  }

  rm(vals)
  gc()
}

end_time <- Sys.time()
cat("\nRasterization completed in ",
    round(difftime(end_time, start_time, units = "mins"), 2), " minutes.\n")
```

## 8. Create EBV Cube NetCDF file

### 8.1 Check paths and available rasters

Define the metadata JSON and the target NetCDF file, and confirm the stacked GeoTIFFs from Section 7 are present in the expected folders. This check catches missing inputs before the cube is built.

```{r, eval=FALSE}
# Define date and check id
id            <- 100
creation_date <- "20260626"
cat("Created date:", format(as.Date(creation_date, format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Define paths
root <- paste0("I:/ebvportal/upload/", id)
js   <- file.path(root, "metadata.json")

# Note:
# Please follow the naming convention
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0("S:/ebv_portal/", id, "/alexander_ecostr_id", id, "_", creation_date, ".nc")

cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Rasterized root folder:", out_dir, "\n")

# List all stacked GeoTIFFs
tif_files <- list.files(out_dir, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)

cat("Total stacked GeoTIFF files found:", length(tif_files), "\n")
```

### 8.2 Define EBV Cube structure

Define the cube's spatial, temporal, and thematic components: entities, extent and resolution, CRS, time steps, fill value, and data precision. Entities here are protected and unprotected forest, and the values are continuous forest-management variables.

```{r, eval=FALSE}
cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")

# Spatio-temporal setting
r0     <- rast(tif_files[1])
extent <- as.vector(ext(r0))
res_xy <- res(r0)
epsg   <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = target_years, extent = extent, res = res_xy, epsg = epsg)

# Fill value and datatype
fillvalue <- -1
prec      <- "float"
```

### 8.3 Create empty NetCDF file

Create the empty NetCDF file from the properties defined above. It holds the cube structure but no raster data yet — that comes in Section 9.

```{r, eval=FALSE}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = extent,
  resolution = res_xy,
  timesteps  = target_years,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

cat("EBV cube created at:", nc, "\n")
```

## 9. Populate the EBV Cube

Insert the stacked rasters into the cube one by one. For each scenario, metric, and entity (protection), the workflow reads the stacked GeoTIFF and inserts each year band at its correct scenario / metric / time step / entity position.

```{r, eval=FALSE}
# Get the properties and entities
props      <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

cat("Entities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (m_no in seq_along(metrics)) {

    metric <- metrics[m_no]

    for (e_no in seq_len(n_entities)) {

      prot <- entity_levels[e_no]

      in_file <- file.path(out_dir, metric, sc, level_4,
                           sprintf("Forestry_%s_%s_%s.tif", metric, sc, prot))

      if (!file.exists(in_file)) {
        warning("Missing file, skipped: ", in_file)
        next
      }

      # One band per year in the stacked file
      for (t_no in seq_along(target_years)) {

        ebv_add_data(
          filepath_nc = nc,
          data        = in_file,
          band        = t_no,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          ignore_RAM  = TRUE,
          raw         = TRUE,
          verbose     = FALSE
        )
      }

      gc(verbose = FALSE)
      cat("  ✔ inserted", metric, "|", sc, "|", entities[e_no], "\n")
    }
  }
}

# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

cat("\n🎉 DONE. All layers inserted into:\n", nc, "\n")
```

## 10. Validation and visualization

A quick check that the cube is readable and that the metrics, entities, and time steps line up.

```{r, eval=FALSE}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@ebv_cube$type
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[8, 1], entity = 2,
        timestep = 2, verbose = FALSE)
```

## 11. Clean up temporary extraction files

With everything inserted, the temporary extraction folder can be removed to free disk space.

```{r, eval=FALSE}
cat("Removing temporary extraction folder:", local_tmp, "\n")
unlink(local_tmp, recursive = TRUE)

if (!dir.exists(local_tmp)) {
  cat("✔ Temporary extraction folder successfully removed.\n")
} else {
  warning("Temporary folder could not be removed: ", local_tmp)
}
```
