---
title: "PLUM Land-Use (Crop Intensity) – Rasterization to 0.5° and EBV Cube Generation"
author: "Emmanuel Oceguera-Conchas, Andres Marmol Guijarro, Magali Weissgerber"
date: 2026-02-05
output: html_document
---

## 1. Introduction

This workflow processes the PLUM 0.5° crop land-use crop and intensity data and converts it into an EBV Cube NetCDF file (dataset ID 101).

Each PLUM table reports seven management variables per 0.5° cell — area, fertiliser intensity and quantity, irrigation intensity and quantity, other-management intensity, and yield — for each crop and farming type (`conventional` / `restricted`). Crop and farming type together form the EBV entity (24 in total); the seven variables are the metrics. The workflow rasterizes each variable per entity, stacks the years into one raster per variable and entity, and stores the result as an EBV Cube. Restricted farming occurs only under the NaC and NfS scenarios, so restricted entities are empty (NA) for the others

The workflow comprises two core steps:

-   *Rasterization:* Each management variable is rasterized into the 0.5° grid per crop and farming type, and stacked into one GeoTIFF per variable and entity (one band per year).
-   *EBV Cube NetCDF file generation:* The stacked layers are structured into an EBV Cube, organizing the data by scenario, metric (the seven variables), time step, and entity (crop × farming type).

In the notebook, these steps span four sections: preparing the working environment (loading `terra`, `ebvcube` and setting paths), rasterizing and stacking the variables, building the EBV Cube structure, and populating it.

### 1.1 PLUM scenarios and time steps

The workflow processes five scenarios across four years each. The scenario order follows the metadata JSON. Restricted farming exists only under NaC and NfS; the restricted entities are NA for BAU, NfN, and NfNl.

| Scenario                  | Abbreviation | Available Years        |
|---------------------------|--------------|------------------------|
| Nature for Nature         | NfN          | 2020, 2030, 2050, 2070 |
| Nature for Nature (local) | NfNl         | 2020, 2030, 2050, 2070 |
| Nature as Culture         | NaC          | 2020, 2030, 2050, 2070 |
| Nature for Society        | NfS          | 2020, 2030, 2050, 2070 |
| Business as usual         | BAU          | 2020, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   The PLUM scenario ZIP archives (tabular `.txt.gz` files), in the expected folder structure `SCENARIO/s1/YEAR/LandUse.txt.gz`
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   Enough temporary disk space for raster processing
-   A working `ebvcube` install, plus `terra`

```{r}
knitr::opts_chunk$set(cache = TRUE)   
```

## 2. Load libraries

Here we load the packages used throughout the workflow: `terra` for building and writing the stacked GeoTIFFs, and `ebvcube` for creating and filling the EBV Cube NetCDF file.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
#install.packages("ebvcube")

library(terra)
library(ebvcube)
```

## 3. Configure temporary raster storage

Rasterization can write a lot of temporary data to disk, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths

Input: the folder of PLUM ZIP archives. Output: the folder for the stacked per-variable GeoTIFFs, organized by metric and scenario.

```{r}
base_dir <- "../BES-SIM 2/PLUM_LU/0.5_degrees"
out_dir  <- "../BES-SIM 2/PLUM_LU/0.5_degrees/landuse_rast"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

cat("Input root:", base_dir, "\n")
cat("Output root:", out_dir, "\n")
```

## 5. Define the main workflow settings

Set the scenarios, years, metrics, and entities (crop × farming type) to process.

```{r}
# Scenario identifiers (must match the ZIP internal paths AND the metadata JSON order)
scenarios <- c("NfN_NfN", "NfNl_NfNl", "NaC_NaC", "NfS_NfS", "BAU")

# Years (band order in each stack)
target_years <- c("2020", "2030", "2050", "2070")

# Metrics (order must match the metadata JSON). These are the value columns in the tables:
# A area (Mha); FI fertiliser intensity (0-1); FQ fertiliser quantity (kgN/ha);
# II irrigation intensity (0-1); IQ irrigation quantity (L/m2);
# OI other-management intensity (0-1); Y yield (t/ha).
metrics <- c("A", "FI", "FQ", "II", "IQ", "OI", "Y")

# Entities = crop x farming type
entities <- c(
  "conventional-wheat",        "restricted-wheat",
  "conventional-sugar",        "restricted-sugar",
  "conventional-starchyRoots", "restricted-starchyRoots",
  "conventional-setaside",     "restricted-setaside",
  "conventional-rice",         "restricted-rice",
  "conventional-pulses",       "restricted-pulses",
  "conventional-pasture",      "restricted-pasture",
  "conventional-oilcropsOther","restricted-oilcropsOther",
  "conventional-oilcropsNFix", "restricted-oilcropsNFix",
  "conventional-maize",        "restricted-maize",
  "conventional-fruitveg",     "restricted-fruitveg",
  "conventional-energycrops",  "restricted-energycrops"
)

# Folder level inside each ZIP, and the variable
level_4 <- "s1"
var_4   <- "LandUse"

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Years:", paste(target_years, collapse = ", "), "\n")
cat("Metrics:", paste(metrics, collapse = ", "), "\n")
cat("Entities:", length(entities), "\n")
```

## 6. Build the 0.5° raster template

```{r}
zip_files <- list.files(base_dir, pattern = "\\.zip$", full.names = TRUE)
stopifnot(length(zip_files) > 0)

cat("Zip files found:\n", paste(zip_files, collapse = "\n"), "\n")

r_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = 0.5,
  crs = "EPSG:4326"
)

# Half-cell offset for converting lower-left coordinates to cell centres
half_res <- res(r_template) / 2
half_res
```

## 7. Rasterize and stack

For each scenario, read the yearly tables, split each into entities (crop × farming type), and write one stacked GeoTIFF per metric and entity (bands = years). Zeros are kept as data; an entity absent from a scenario (e.g. restricted under BAU) is written all-NA so the cube structure stays consistent.

```{r}
start_time <- Sys.time()

ny     <- length(target_years)
n_cell <- ncell(r_template)

for (scen in scenarios) {

  scen_zip <- grep(scen, zip_files, value = TRUE)
  if (length(scen_zip) == 0L) {
    warning("No zip found for scenario: ", scen)
    next
  }
  scen_zip <- scen_zip[1]

  cat("\n=============================\n")
  cat("Scenario:", scen, "\n")

  # Read each year and split into entity subsets (farming type + crop)
  year_splits <- vector("list", ny)

  for (y in seq_along(target_years)) {

    yr <- target_years[y]
    rel_path <- sprintf("%s/%s/%s/%s.txt.gz", scen, level_4, yr, var_4)

    # Extract flat (strip the archive's folders), then read
    tmp_dir <- file.path(local_tmp, scen, level_4, yr)
    dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)
    unzip(scen_zip, files = rel_path, exdir = tmp_dir, junkpaths = TRUE)

    txt_gz <- file.path(tmp_dir, basename(rel_path))
    con <- gzfile(txt_gz, "rt")
    dat <- read.table(con, header = TRUE, check.names = FALSE)
    close(con)
    unlink(txt_gz)

    cat("  year", yr, "\n")

    stopifnot(all(c("Lon", "Lat", "Type", "Crop") %in% names(dat)))
    stopifnot(all(metrics %in% names(dat)))

    # Cell index from the cell centre; drop rows outside the grid
    dat$cell <- cellFromXY(r_template, cbind(dat$Lon + half_res[1],
                                             dat$Lat + half_res[2]))
    dat <- dat[!is.na(dat$cell), ]

    # Entity key = farming type + crop, e.g. "conventional_wheat".
    # Keep only the columns we need, then split by entity.
    dat$entity <- paste(dat$Type, dat$Crop, sep = "-")
    dat <- dat[, c("entity", "cell", metrics)]
    year_splits[[y]] <- split(dat, dat$entity)

    rm(dat)
    gc()
  }

  # Write one stacked GeoTIFF per metric and entity (bands = years)
  for (m in metrics) {

    for (ent in entities) {

      vmat <- matrix(NA_real_, n_cell, ny)
      for (y in seq_along(target_years)) {
        sub <- year_splits[[y]][[ent]]   # NULL if this entity is absent that year
        if (!is.null(sub) && nrow(sub) > 0) vmat[sub$cell, y] <- sub[[m]]
      }
      # Note: zeros are real data and stay 0; only cells with no row remain NA.

      r_stack <- rast(r_template, nlyrs = ny)
      values(r_stack) <- vmat
      names(r_stack)  <- target_years

      out_file <- file.path(out_dir, m, scen, level_4,
                            sprintf("LandUse_%s_%s_%s.tif", m, scen, ent))
      
      dir.create(dirname(out_file), recursive = TRUE, showWarnings = FALSE)

      writeRaster(
        r_stack, out_file,
        datatype  = "FLT4S",
        overwrite = TRUE,
        gdal = c("COMPRESS=DEFLATE")
      )
    }
    cat("  wrote metric:", m, "\n")
  }

  rm(year_splits)
  gc()
}

end_time <- Sys.time()
cat("\nRasterization completed in ",
    round(difftime(end_time, start_time, units = "mins"), 2), " minutes.\n")
```

## 8. Create EBV Cube NetCDF file

### 8.1 Check paths and available rasters

Define the metadata JSON and the target NetCDF file, and confirm the stacked GeoTIFFs from Section 7 are present in the expected folders. This check catches missing inputs before the cube is built.

```{r}
# Define date and check id
id            <- 101
creation_date <- "20260629"
cat("Created date:", format(as.Date(creation_date, format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Define paths
root <- paste0("I:/ebvportal/upload/", id)
js   <- file.path(root, "metadata.json")

# Note:
# Please follow the naming convention
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- paste0("S:/ebv_portal/", id, "/alexander_ecostr_id", id, "_", creation_date, ".nc")

cat("EBV Cube NetCDF file name:", basename(nc), "\n")
cat("Rasterized root folder:", out_dir, "\n")

# List all stacked GeoTIFFs
tif_files <- list.files(out_dir, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)

cat("Total stacked GeoTIFF files found:", length(tif_files), "\n")
```

### 8.2 Define EBV Cube structure

Define the cube's spatial, temporal, and thematic components: entities, extent and resolution, CRS, time steps, fill value, and data precision. Entities here are crop × farming type, and the values are continuous management variables.

```{r, eval}
cat("Entity names:", paste(entities, collapse = ", "), "\n")
cat("Entities number:", length(entities), "\n")

# Spatio-temporal setting
r0     <- rast(tif_files[1])
extent <- as.vector(ext(r0))
res_xy <- res(r0)
epsg   <- 4326

cat("Spatio-temporal settings\n")
cat("=======================\n")
list(timesteps = target_years, extent = extent, res = res_xy, epsg = epsg)

# Fill value and datatype
fillvalue <- -1
prec      <- "float"
```

### 8.3 Create empty NetCDF file

Create the empty NetCDF file from the properties defined above. It holds the cube structure but no raster data yet — that comes in Section 9.

```{r}
cat("Creating EBV Cube NetCDF\n")
cat("=======================\n")

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = extent,
  resolution = res_xy,
  timesteps  = target_years,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

cat("EBV cube created at:", nc, "\n")
```

## 9. Populate the EBV Cube

Insert the stacked rasters into the cube one by one. For each scenario, metric, and entity, the workflow reads the stacked GeoTIFF and inserts each year band at its correct scenario / metric / time step / entity position.

```{r}
# Get the properties and entities
props      <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

cat("Entities in netCDF:", n_entities, "\n")

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  sc <- scenarios[s_no]

  for (m_no in seq_along(metrics)) {

    metric <- metrics[m_no]

    for (e_no in seq_len(n_entities)) {

      ent <- entities[e_no]

      in_file <- file.path(out_dir, metric, sc, level_4,
                           sprintf("LandUse_%s_%s_%s.tif", metric, sc, ent))

      if (!file.exists(in_file)) {
        warning("Missing file, skipped: ", in_file)
        next
      }

      # One band per year in the stacked file
      for (t_no in seq_along(target_years)) {

        ebv_add_data(
          filepath_nc = nc,
          data        = in_file,
          band        = t_no,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          ignore_RAM  = FALSE,
          raw         = FALSE,
          verbose     = FALSE
        )
      }

      gc(verbose = FALSE)
    }
    cat("  ✔ inserted", metric, "|", sc, "\n")
  }
}

# Check time
end.time <- Sys.time()
time.taken <- end.time - start.time
time.taken

cat("\n🎉 DONE. All layers inserted into:\n", nc, "\n")
```

## 10. Validation and visualization

A quick check that the cube is readable and that the metrics, entities, and time steps line up.

```{r}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@ebv_cube$type
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 1,
        timestep = 1, verbose = FALSE)
```

## 11. Clean up temporary extraction files

With everything inserted, the temporary extraction folder can be removed to free disk space.

```{r, eval=FALSE}
cat("Removing temporary extraction folder:", local_tmp, "\n")
unlink(local_tmp, recursive = TRUE)

if (!dir.exists(local_tmp)) {
  cat("✔ Temporary extraction folder successfully removed.\n")
} else {
  warning("Temporary folder could not be removed: ", local_tmp)
}
```
