---
title: "PLUM Land Use projections - EBV Cube"
author: "Emmanuel Oceguera Conchas"
date: "2026-02-03"
output: 
  html_document:
    df_print: paged
  pdf_document: default  
---

## Introduction

The aim of this workflow is to create an EBV Cube (netCDF) file for PLUM Land use (ID 99) 
that can be uploaded to the EBV Data Portal. We convert each land-use class into 
a binary presence/absence layer (1/0) and store each class as an EBV entity. 

In each file, there are multiple land use (e.g. urban, cropland, etc) as bands of the .tif.
Ocean and water are excluded from the entity list and are also masked to NA 
before binarizatio.


This datasates is diveded into scenarios:
- Each folder under `S:\\ebv_portal\\99\\data\\` represents a scenario
  - Business as usual (BAU)
  - Nature as Culture (NaC_NaC)
  - Nature for Society (NfS_NfS)
  - Nature for Nature (NfN_NfN)
  - Nature for Nature local (NfNl_NfNl)

The scenario order in the netCDF is enforced to match the
scenario order in the metadata JSON.

In each scenario there are multiple years
- Each .tff file is a multiband, where each band corresponds to a year
  - 2020
  - 2030
  - 2050
  - 2070
  - 2100

- Each `land_cover` class is transformed into a binary raster:
  - `1` = presence
  - `0` = absence
  - `NA` is preserved

- The datatype
  - INT1U as a byte (0/1)



```{r}
knitr::opts_chunk$set(cache = TRUE)
```


###  Clean environment and load libraries
Note: If any of the packages are not installed, uncomend lines.
```{r, eval=FALSE}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")
# install.packages("ebvcube")
# install.packages("future.apply")
library(ebvcube)
library(terra)
library(stringr)
library(dplyr)
library(future.apply)

```
### Terra temp settings and load libraries
The temp folder should be placed where there is nough memory space

```{r, eval=FALSE}
# Defile local temp folder
local_tmp <- "D:\\Emma\\temp"
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

# Terra temp settings
terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp
)

```


### Define paths
```{r, eval=FALSE}
# Define ID and date for the file name
id <- 99
date <- "20260212"

# Define root paths, json and output
root <- paste0("I:\\ebv_portal\\", id) # biocon\\
js   <- file.path(root, "metadata.json")
nc <- paste0("S:\\ebv_portal\\", id, "\\alexander_ecostr_id", id, "_", date, ".nc")


# Define .tif root and list .tif files
tif_root  <- file.path("S:\\ebv_portal", id, "data")
tif_files <- list.files(tif_root, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)
cat(paste0("Path: ", tif_files, "\n"), "\n")
```

### Specify parameters
```{r, eval=FALSE}
# Define scenarios and years
scenario_levels <- c("NfN", "NfNl", "NaC", "NfS", "BAU")
years <- c(2021, 2030, 2050, 2070, 2100)
timesteps <- as.character(years)

cat("Scenarios:", paste(scenario_levels, collapse=", "), "\n")
cat("Years:", paste(timesteps, collapse=", "), "\n")
```
### Define EBV cube structure and create empty NetCDF file
```{r, eval=FALSE}
# Get general spatial-temporal setting
r0 <- rast(tif_files[1])
datatype  <- datatype(r0[[1]])
extent <- as.vector(ext(r0))
res_xy  <- res(r0)
epsg    <- 4326

# Get levels/entitites from the raste 
class <- terra::levels(r0[[1]])
print(class[[1]])

# Get all entities
entitites_df <- as.data.frame(class[[1]])
class_id    <- as.integer(entitites_df$id)
class_names <- as.character(entitites_df$land_cover)

# Remove ocean, water and no data from the entities list
entities <- class_names[!(class_id %in% c(0, 77, 99))]
cat("Entities to be included in the EBV cube:\n", paste("-", entities, "\n"), "\n")


# Create empty EBV file
fillvalue <- -1
prec <- "byte"

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = extent,
  resolution = res_xy,
  timesteps  = timesteps,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

cat("EBV cube created at:", nc, "\n")

```

### Create binary rasters in parallel
Note: Before running this chunk verify the capacitites of your machine to adjust
  the pla() function in terms of workers.
  
```{r, eval=FALSE}

# Define values for masking
mask_ids <- c(0, 99, 77)

# Build a job table
jobs <- expand.grid(
  s_no = seq_along(scenario_levels),
  t_no = seq_along(timesteps),
  e_no = seq_along(class_id),
  KEEP.OUT.ATTRS = FALSE
)


# Use 3 multisessions to process in parallel
plan(multisession, workers = 3)

# Helper to make safe filenames
safe_name <- function(x) gsub("[^A-Za-z0-9_]+", "_", x)

# Create all binary tif files in parallel
out_paths <- future_sapply(seq_len(nrow(jobs)), function(i) {

  s_no <- jobs$s_no[i]
  t_no <- jobs$t_no[i]
  e_no <- jobs$e_no[i]

  sc <- scenario_levels[s_no]
  yr <- timesteps[t_no]
  code <- class_id[e_no]
  ent  <- entities[e_no]

  path_sc <- tif_files[grepl(paste0("_", sc, "_"), basename(tif_files))]
  if (length(path_sc) != 1) stop("Expected exactly 1 tif for scenario ", sc, " got ", length(path_sc))

  r <- rast(path_sc)
  r_t <- r[[t_no]]

  # mask water/ocean/nodata
  r_t_masked <- ifel(is.na(r_t), NA, ifel(r_t %in% mask_ids, NA, r_t))

  # TRUE presence, FALSE absence, NA for masked
  bin_r <- ifel(is.na(r_t_masked), NA, ifel(r_t_masked == code, 1, 0))

  out_tif <- file.path(local_tmp, "bin", sc, yr, paste0(safe_name(ent), ".tif"))
  dir.create(dirname(out_tif), showWarnings = FALSE, recursive = TRUE)

  writeRaster(
    bin_r, out_tif,
    overwrite = TRUE,
    datatype  = "INT1S",
    NAflag    = -1,
    wopt      = list(gdal = c("COMPRESS=DEFLATE"))
  )

  out_tif
})

plan(sequential)
```


### Populate EBV cube with data
In this process, entities are saved in temp folder,
```{r, eval=FALSE}

# Check propeties
props <- ebv_properties(nc)
n_entities <- length(props@general$entity_names)

# Define metric
metric_no <- 1

# loop through each scenario
for (s_no in seq_along(scenario_levels)) {
  sc <- scenario_levels[s_no]
  
  # Loop through each timestep
  for (t_no in seq_along(timesteps)) {
    yr <- timesteps[t_no]
    
    
    cat("\n=== INSERT | scenario:", sc, "| year:", yr, "===\n")
    
    
    e_start <- 1
    # Loop through each entity
    for (e_no in e_start:n_entities) {
      
      code <- entities[e_no]
      
      cat("Entity:", entities[e_no], " (code ", code, ")\n")
      
      # Create a temporary file on the local disk
      tmp_tif <- file.path(local_tmp, "bin", sc, yr, paste0(gsub("[ /]", "_", entities[e_no]), ".tif"))
      
      # Check if exists
      if (!file.exists(tmp_tif)) {
        stop("Temporary file not found: ", tmp_tif)
      }

      # Add data to the nc file, using one byte per pixel
      ebv_add_data(
        filepath_nc = nc,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        scenario    = s_no,
        metric      = 1,
        ignore_RAM  = FALSE,
        raw         = FALSE,
                verbose     = FALSE
      )
      # Print that entitiy is inserted
      cat("✔", entities[e_no], "- Entity added to EBV cube\n")
    }
    
    # Free up R Memory
    rm(r_t)
    gc()
  }
}

```

### Quick vallidation and visualization
```{r, eval=FALSE}    
ebv_analyse(nc, entity=1, timestep=1, scenario=1, metric=1)
props <- ebv_properties(nc, scenario=1, metric=1)
props@ebv_cube$type
props@general$entity_names
# Define data cube
datacubes <- ebv_datacubepaths(nc, verbose=FALSE)
# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1,1], entity = 1,
        timestep = 1, classes = 2, verbose = FALSE)
```
    