---
title: "LUH3 v1.1 — GeoTIFF Extraction"
subtitle: "Preprocessing step for all MAgPIE EBVcube workflows"
author: "Andres Marmol Guijarro, Emmanuel Oceguera-Conchas"
date: "2026-03-19"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

## 1. Introduction

This script is the shared preprocessing step for all MAgPIE EBVcube
workflows. It extracts selected time slices from four LUH3 v1.1 NetCDF
files prepared for CMIP7 and writes one GeoTIFF per variable × scenario
× year into a single flat output folder. The resulting GeoTIFFs are then
used as inputs by the individual EBVcube creation scripts (scripts
02–05).

The source data covers two climate scenarios, each driven by a different
Integrated Assessment Model (IAM):

| Scenario | Code | IAM | Description |
|------------------|------------------|------------------|---------------|
| Low-MAgPIE | `vl` | REMIND-MAgPIE | Very low emissions, land-use change driven by REMIND-MAgPIE |
| H-GCAM | `h` | GCAM | High emissions scenario, land-use change driven by GCAM |

Each scenario is provided in two NetCDF files:

-   *multiple-states*: land cover fractions and secondary vegetation
    properties (15 variables)
-   *multiple-management*: fertilization, irrigation, biofuels, forest
    harvest, protection, and added tree cover (34 variables)

Output GeoTIFFs follow the naming convention
`{file_type}_{scenario}_{variable}_{year}.tif`, e.g.
`states_vl_c3ann_2022.tif` or `management_h_rndwd_2050.tif`. This
encoding allows scenario, file type, variable, and year to be parsed
back from the filename without requiring subfolders.

> **Note:** Plantation variables (`pltns`, `prtct_pltns`, `pltns_bfuel`,
> `pltns_wdprd`), biofuel variables (`cpbf1_*`, `cpbf2_*`), and
> protection variables (`prtct_*`) are set to zero in LUH3 v1.1 and will
> produce all-zero GeoTIFFs. These are extracted for structural
> completeness and will be updated when future versions of the dataset
> provide non-zero values.

### 1.1 Overview of the workflow

This script follows three core processing stages:

1.  *Prepare the working environment.* Load the required R packages,
    define paths, and set the target years and variable lists.

2.  *Define the extraction function.* A reusable function reads one
    variable from a NetCDF file, selects the target time layers, masks
    fill values, and writes one LZW-compressed GeoTIFF per year.

3.  *Extract all variables.* Loop over both scenarios × both file types
    × all variables, calling the extraction function for each
    combination.

### 1.2 Before you start

To ensure the workflow runs smoothly, the following conditions should be
met:

-   The four raw LUH3 v1.1 NetCDF files are available in the `raw/`
    folder (see paths in section 4).
-   Sufficient disk space is available for the output GeoTIFFs. With two
    scenarios × 49 variables × 4 years, approximately 400 files will be
    created.
-   The `terra` package is installed and working correctly.
-   This script should be run before any of the EBVcube creation scripts
    (02–05), as they all depend on the GeoTIFFs produced here.

------------------------------------------------------------------------

## 2. Load libraries

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")

library(terra)
```

------------------------------------------------------------------------

## 3. Configure temporary raster storage

Raster operations can create large temporary files. It is recommended to
define a dedicated temporary folder on a drive with sufficient free
space. Adjust the path below to a local drive if needed.

```{r}
local_temp <- "../raw/temp_terra"
dir.create(local_temp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_temp
)

cat("Temporary folder:", local_temp, "\n")
```

------------------------------------------------------------------------

## 4. Define paths

Here we define the location of the raw LUH3 v1.1 NetCDF files and the
output folder where the extracted GeoTIFFs will be written.

All paths are defined relative to the location of this script, so the
workflow is portable and does not depend on absolute paths.

```{r}
# Set working directory to the location of this script (RStudio only)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Raw NetCDF input files
raw_dir <- "../raw"

# Output folder for extracted GeoTIFFs (shared with all EBVcube scripts)
out_dir <- "../raw/temp"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Low-MAgPIE scenario (vl) — REMIND-MAgPIE
states_nc_vl     <- file.path(raw_dir,
  "multiple-states_input4MIPs_landState_ScenarioMIP_UofMD-landState-vl-3-1_gn_2022-2100.nc")
management_nc_vl <- file.path(raw_dir,
  "multiple-management_input4MIPs_landState_ScenarioMIP_UofMD-landState-vl-3-1_gn_2022-2100.nc")

# H-GCAM scenario (h) — GCAM
states_nc_h      <- file.path(raw_dir,
  "multiple-states_input4MIPs_landState_ScenarioMIP_UofMD-landState-h-3-1_gn_2022-2100.nc")
management_nc_h  <- file.path(raw_dir,
  "multiple-management_input4MIPs_landState_ScenarioMIP_UofMD-landState-h-3-1_gn_2022-2100.nc")

# Verify input files exist
for (f in c(states_nc_vl, management_nc_vl, states_nc_h, management_nc_h)) {
  cat(basename(f), "— exists:", file.exists(f), "\n")
}
cat("Output folder:", out_dir, "\n")
```

------------------------------------------------------------------------

## 5. Define workflow settings

The LUH3 v1.1 files run annually from 2022 to 2100 (79 time steps). We
extract four specific years: 2022 (reference), 2030, 2050, and 2070. The
time index for a given year is computed as `year - 2022 + 1`.

```{r}
# Target years and their layer indices in the NetCDF files
target_years <- c(2022, 2030, 2050, 2070)
year_indices <- target_years - 2022 + 1

data.frame(Year = target_years, `Layer index` = year_indices,
           check.names = FALSE)
```

------------------------------------------------------------------------

## 6. Define variable lists

### 6.1 States variables

The *multiple-states* file contains 15 variables describing land cover
fractions and secondary vegetation properties.

```{r}
states_vars <- c(
  # Cropland
  "c3ann", "c3nfx", "c3per", "c4ann", "c4per",
  # Managed land
  "pastr", "range", "urban",
  # Natural vegetation
  "primf", "primn", "secdf", "secdn", "pltns",
  # Secondary vegetation properties
  "secma", "secmb"
)

states_meta <- data.frame(
  Variable = states_vars,
  Description = c(
    "C3 annual crops", "C3 nitrogen-fixing crops", "C3 perennial crops",
    "C4 annual crops", "C4 perennial crops",
    "Managed pasture", "Rangeland", "Urban land",
    "Primary forest", "Primary non-forest",
    "Secondary forest", "Secondary non-forest", "Plantation forests *",
    "Secondary mean age", "Secondary mean biomass carbon density"
  ),
  Units = c(rep("fraction", 13), "years", "kg m-2"),
  Category = c(
    rep("Cropland", 5), rep("Managed land", 3),
    rep("Natural vegetation", 5), rep("Secondary veg. properties", 2)
  )
)

cat("States variables:", length(states_vars), "\n")
knitr::kable(states_meta, align = "llll")
```

### 6.2 Management variables

The *multiple-management* file contains 34 variables describing land
management intensity across crop, forest, and protection domains.

```{r}
management_vars <- c(
  # Fertilization (kg ha-1 yr-1)
  "fertl_c3ann", "fertl_c3nfx", "fertl_c3per", "fertl_c4ann", "fertl_c4per",
  # Irrigation (fraction)
  "irrig_c3ann", "irrig_c3nfx", "irrig_c3per", "irrig_c4ann", "irrig_c4per",
  "flood",
  # 1st generation biofuel fraction (zero in LUH3 v1.1)
  "cpbf1_c3ann", "cpbf1_c3nfx", "cpbf1_c3per", "cpbf1_c4ann", "cpbf1_c4per",
  # 2nd generation biofuel fraction (zero in LUH3 v1.1)
  "cpbf2_c3ann", "cpbf2_c3nfx", "cpbf2_c3per", "cpbf2_c4ann", "cpbf2_c4per",
  # Harvested fraction of perennial crops
  "fharv_c3per", "fharv_c4per",
  # Forest harvest allocation (fraction)
  "rndwd", "fulwd", "combf", "pltns_bfuel", "pltns_wdprd",
  # Protected area fractions (zero in LUH3 v1.1)
  "prtct_primf", "prtct_primn", "prtct_secdf", "prtct_secdn", "prtct_pltns",
  # Added tree cover / afforestation (fraction)
  "addtc"
)

cat("Management variables:", length(management_vars), "\n")
```

------------------------------------------------------------------------

## 7. Define the extraction function

The function reads one variable from a NetCDF file, selects the target
time layers by index, masks fill values (any value ≥ 1×10¹⁹ is set to
`NA`), assigns WGS84 if the CRS is missing, and writes one
LZW-compressed GeoTIFF per target year.

The `scenario` and `file_type` arguments are embedded in the output
filename so that both can be parsed back from the filename alone.

```{r}
extract_variable <- function(nc_path, varname, years, indices, outdir,
                              scenario, file_type) {

  cat("  Processing:", varname, "\n")

  # Read the variable — skip gracefully if not found
  r <- tryCatch(
    rast(nc_path, subds = varname),
    error = function(e) {
      cat("  Skipping", varname, "— could not read:", e$message, "\n")
      return(NULL)
    }
  )
  if (is.null(r)) return(invisible(NULL))

  # Select target time layers
  nlyr_available <- nlyr(r)
  valid_idx      <- indices[indices <= nlyr_available]
  valid_yrs      <- years[indices <= nlyr_available]

  if (length(valid_idx) == 0) {
    cat("  Skipping", varname, "— no valid time indices\n")
    return(invisible(NULL))
  }

  r_sub <- r[[valid_idx]]

  # Assign WGS84 if CRS is missing
  if (is.na(crs(r_sub))) crs(r_sub) <- "EPSG:4326"

  # Mask fill values
  r_sub[r_sub >= 1e19] <- NA

  # Write one GeoTIFF per year
  written <- character()
  for (i in seq_along(valid_yrs)) {
    fname <- file.path(outdir,
                       paste0(file_type, "_", scenario, "_",
                              varname, "_", valid_yrs[i], ".tif"))
    writeRaster(r_sub[[i]], fname, overwrite = TRUE,
                gdal = c("COMPRESS=LZW", "TILED=YES"))
    written <- c(written, basename(fname))
  }

  return(written)
}
```

------------------------------------------------------------------------

## 8. Extract all variables

The extraction loops over both scenarios × both file types × all
variables. Progress is printed to the console for each combination.

```{r}
# Define scenario structure
scenarios <- list(
  vl = list(label = "Low-MAgPIE (VL)", states = states_nc_vl,
             management = management_nc_vl),
  h  = list(label = "H-GCAM (H)",      states = states_nc_h,
             management = management_nc_h)
)

# Run extraction
all_log <- list()
start_time <- Sys.time()

for (sc in names(scenarios)) {

  cat("\n==============================\n")
  cat("Scenario:", scenarios[[sc]]$label, "\n")
  cat("==============================\n")

  # States variables
  cat("\n--- States ---\n")
  for (v in states_vars) {
    all_log[[paste(sc, "states", v, sep = "_")]] <- extract_variable(
      nc_path   = scenarios[[sc]]$states,
      varname   = v,
      years     = target_years,
      indices   = year_indices,
      outdir    = out_dir,
      scenario  = sc,
      file_type = "states"
    )
  }

  # Management variables
  cat("\n--- Management ---\n")
  for (v in management_vars) {
    all_log[[paste(sc, "management", v, sep = "_")]] <- extract_variable(
      nc_path   = scenarios[[sc]]$management,
      varname   = v,
      years     = target_years,
      indices   = year_indices,
      outdir    = out_dir,
      scenario  = sc,
      file_type = "management"
    )
  }
}

elapsed <- Sys.time() - start_time
cat("\nExtraction complete. Time elapsed:", round(elapsed, 1),
    units(elapsed), "\n")
```

------------------------------------------------------------------------

## 9. Summary and verification

After extraction, we count the output GeoTIFFs and verify the expected
number of files per scenario and file type.

```{r}
tif_files <- list.files(out_dir, pattern = "\\.tif$", full.names = FALSE)

# Parse filename components
library(stringr)

parsed <- data.frame(
  filename  = tif_files,
  file_type = str_extract(tif_files, "^(states|management)"),
  scenario  = str_extract(tif_files, "(?<=^(states|management)_)(vl|h)"),
  year      = as.integer(str_extract(tif_files, "\\d{4}(?=\\.tif)"))
)

# Summary counts
summary_tbl <- as.data.frame(table(parsed$file_type, parsed$scenario))
names(summary_tbl) <- c("File type", "Scenario", "Count")

cat("Output folder:", out_dir, "\n")
cat("Total GeoTIFFs found:", length(tif_files), "\n\n")
cat("Expected per scenario:\n")
cat("  States     :", length(states_vars) * length(target_years), "files\n")
cat("  Management :", length(management_vars) * length(target_years), "files\n\n")
print(summary_tbl)

cat("\nThese GeoTIFFs are the inputs for scripts 02-05.\n")
```

------------------------------------------------------------------------

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP ·
Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
