---
title: "LUH3 v1.1 — GeoTIFF Extraction"
subtitle: "Preprocessing step for all MAgPIE EBVcube workflows"
author: "Emmanuel Oceguera-Conchas, Andres Marmol-Guijarro, Magali Weissgerber"
date: 2026-03-19
output: html_document
---

## 1. Introduction

This script is the shared preprocessing step for all MAgPIE EBVcube workflows. It extracts selected time slices from four LUH3 v1.1 NetCDF files prepared for CMIP7 and writes one GeoTIFF per variable × scenario × year into a single flat output folder. The resulting GeoTIFFs are then used as inputs by the individual EBVcube creation scripts (scripts 02–05).

The source data covers two climate scenarios, each driven by a different Integrated Assessment Model (IAM), and each provided in two netCDF files: a *state* file (105 variables: land cover fractions and secondary vegetation properties) and a *mangament* file (34 variables: fertilization, irrigation, biofuels, forest haverst, proection, and added tree cover.)

The workflow comprises two cores stepts:

-   *Selection:* Each variable is read from its NetCDF file, the four target year layers are selected, and fill values are masked to `NA`.
-   *Extraction:* One LZW-compressed GeoTIFF is written per variable, scenario, and year, named `{file_type}_{scenario}_{variable}_{year}.tif` (e.g. `states_vl_c3ann_2022.tif`) so all four can be parsed back from the filename without subfolders.

In the notebook, these steps spann three sections: preparing the working environment (loading `terra` and setting paths), defining the variable lists, and extracting and writing the GeoTIFFs.

> **Note:** Plantation variables (`pltns`, `prtct_pltns`, `pltns_bfuel`, `pltns_wdprd`), biofuel variables (`cpbf1_*`, `cpbf2_*`), and protection variables (`prtct_*`) are zero in LUH3 v1.1 and will produce all-zero GeoTIFFs. They are extracted for structural completeness and will be updated when future versions of the dataset provide non-zero values.

### 1.1 Scenarios and time steps

The workflow processes two scenarios across four years each (2022, 2030, 2050, 2070), extracted from files that run annually from 2022 to 2100.

| Scenario | Code | IAM | Description |
|------------------|------------------|------------------|------------------|
| Low-MAgPIE | `vl` | REMIND-MAgPIE | Very low emissions, land-use change driven by REMIND-MAgPIE |
| H-GCAM | `h` | GCAM | High emissions scenario, land-use change driven by GCAM |

### 1.2 Before you start

Make sure the following are in place:

-   The four raw LUH3 v1.1 NetCDF files, available in the `raw/` folder (see paths in [Section 4](#sec-4.-define-input-and-output-paths))
-   Enough disk space for the output GeoTIFFs (two scenarios × 49 variables × four years ≈ 400 files)
-   A working `terra` installation
-   This script is run before any of the EBVcube creation scripts (02–05), which all depend on the GeoTIFFs produced here

```{r}
knitr::opts_chunk$set(cache = TRUE)
```

## 2. Load libraries

Here we load the only package used in this workflow: `terra`, for reading the LUH3 NetCDF files and writing the extracted GeoTIFFs.

### 2.1 Clean environment and load libraries

If required, install the missing package before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages("terra")

library(terra)
```

## 3. Configure temporary raster storage

Extracting and writing the GeoTIFFs can create large temporary files, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths {#sec-4.-define-input-and-output-paths}

Input: the four raw LUH3 v1.1 NetCDF files.\
Output: the flat folder of extracted GeoTIFFs shared by scripts 02–05. Paths are relative to this script's location..

```{r}
# Set working directory to the location of this script (RStudio only) - no needed
#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Raw NetCDF input files
raw_dir <- "../raw"

# Output folder for extracted GeoTIFFs (shared with all EBVcube scripts)
out_dir <- "../raw/temp"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Low-MAgPIE scenario (vl) — REMIND-MAgPIE
states_nc_vl     <- file.path(raw_dir,
  "multiple-states_input4MIPs_landState_ScenarioMIP_UofMD-landState-vl-3-1_gn_2022-2100.nc")
management_nc_vl <- file.path(raw_dir,
  "multiple-management_input4MIPs_landState_ScenarioMIP_UofMD-landState-vl-3-1_gn_2022-2100.nc")

# H-GCAM scenario (h) — GCAM
states_nc_h      <- file.path(raw_dir,
  "multiple-states_input4MIPs_landState_ScenarioMIP_UofMD-landState-h-3-1_gn_2022-2100.nc")
management_nc_h  <- file.path(raw_dir,
  "multiple-management_input4MIPs_landState_ScenarioMIP_UofMD-landState-h-3-1_gn_2022-2100.nc")

# Lookup of NetCDF path per scenario and file type (used in Section 7)
nc_paths <- list(
  vl = c(states = states_nc_vl, management = management_nc_vl),
  h  = c(states = states_nc_h,  management = management_nc_h)
)

# Verify input files exist
for (f in c(states_nc_vl, management_nc_vl, states_nc_h, management_nc_h)) {
  cat(basename(f), "— exists:", file.exists(f), "\n")
}
cat("Output folder:", out_dir, "\n")
```

## 5. Define the main workflow settings

Set the scenarios, file types, and target years. The files run annually from 2022 to 2100, so the layer index for a given year is `year - 2022 + 1`.

```{r}
scenarios  <- c("vl", "h")
file_types <- c("states", "management")

# Target years and their layer indices in the NetCDF files
target_years <- c(2022, 2030, 2050, 2070)
year_indices <- target_years - 2022 + 1

data.frame(Year = target_years, `Layer index` = year_indices,
           check.names = FALSE)
```

## 6. Define variable lists

### 6.1 States variables

The *states* file contains 15 variables describing land-cover fractions and secondary-vegetation properties.

```{r}
states_vars <- c(
  # Cropland
  "c3ann", "c3nfx", "c3per", "c4ann", "c4per",
  # Managed land
  "pastr", "range", "urban",
  # Natural vegetation
  "primf", "primn", "secdf", "secdn", "pltns",
  # Secondary vegetation properties
  "secma", "secmb"
)

states_meta <- data.frame(
  Variable = states_vars,
  Description = c(
    "C3 annual crops", "C3 nitrogen-fixing crops", "C3 perennial crops",
    "C4 annual crops", "C4 perennial crops",
    "Managed pasture", "Rangeland", "Urban land",
    "Primary forest", "Primary non-forest",
    "Secondary forest", "Secondary non-forest", "Plantation forests *",
    "Secondary mean age", "Secondary mean biomass carbon density"
  ),
  Units = c(rep("fraction", 13), "years", "kg m-2"),
  Category = c(
    rep("Cropland", 5), rep("Managed land", 3),
    rep("Natural vegetation", 5), rep("Secondary veg. properties", 2)
  )
)

cat("States variables:", length(states_vars), "\n")
knitr::kable(states_meta, align = "llll")
```

### 6.2 Management variables

The *management* file contains 34 variables describing land-management intensity across crop, forest, and protection domains.

```{r}
management_vars <- c(
  # Fertilization (kg ha-1 yr-1)
  "fertl_c3ann", "fertl_c3nfx", "fertl_c3per", "fertl_c4ann", "fertl_c4per",
  # Irrigation (fraction)
  "irrig_c3ann", "irrig_c3nfx", "irrig_c3per", "irrig_c4ann", "irrig_c4per",
  "flood",
  # 1st generation biofuel fraction (zero in LUH3 v1.1)
  "cpbf1_c3ann", "cpbf1_c3nfx", "cpbf1_c3per", "cpbf1_c4ann", "cpbf1_c4per",
  # 2nd generation biofuel fraction (zero in LUH3 v1.1)
  "cpbf2_c3ann", "cpbf2_c3nfx", "cpbf2_c3per", "cpbf2_c4ann", "cpbf2_c4per",
  # Harvested fraction of perennial crops
  "fharv_c3per", "fharv_c4per",
  # Forest harvest allocation (fraction)
  "rndwd", "fulwd", "combf", "pltns_bfuel", "pltns_wdprd",
  # Protected area fractions (zero in LUH3 v1.1)
  "prtct_primf", "prtct_primn", "prtct_secdf", "prtct_secdn", "prtct_pltns",
  # Added tree cover / afforestation (fraction)
  "addtc"
)

cat("Management variables:", length(management_vars), "\n")
```

## 7. Extract all variables {#sec-7.-extract-all-variables}

For each scenario, file type, and variable, read the variable, select the four target year layers, mask fill values (any value ≥ 1×10¹⁹ → `NA`), assign WGS84 if the CRS is missing, and write one LZW-compressed GeoTIFF per year.

```{r}
start_time <- Sys.time()

for (sc in scenarios) {

  for (ft in file_types) {

    nc_path <- nc_paths[[sc]][[ft]]
    vars    <- if (ft == "states") states_vars else management_vars

    cat("\n==============================\n")
    cat("Scenario:", sc, "| File type:", ft, "\n")
    cat("==============================\n")

    for (v in vars) {

      cat("  Processing:", v, "\n")

      # Read the variable from the NetCDF
      r <- rast(nc_path, subds = v)

      # Select the target year layers that exist in the file
      valid     <- year_indices[year_indices <= nlyr(r)]
      valid_yrs <- target_years[year_indices <= nlyr(r)]
      if (length(valid) == 0) {
        warning("No valid time indices for ", v); next
      }
      r_sub <- r[[valid]]

      # Assign WGS84 if the CRS is missing
      if (!nzchar(crs(r_sub))) crs(r_sub) <- "EPSG:4326"

      # Mask fill values
      r_sub[r_sub >= 1e19] <- NA

      # Write one GeoTIFF per year
      for (i in seq_along(valid_yrs)) {
        out_file <- file.path(out_dir,
                              sprintf("%s_%s_%s_%s.tif", ft, sc, v, valid_yrs[i]))
        writeRaster(r_sub[[i]], out_file, overwrite = TRUE,
                    gdal = c("COMPRESS=LZW", "TILED=YES"))
      }
    }
  }
}

elapsed <- Sys.time() - start_time
cat("\nExtraction complete. Time elapsed:", round(elapsed, 1),
    units(elapsed), "\n")
```

## 8. Summary and verification

Count the output GeoTIFFs and check the totals per scenario and file type against the expected numbers.

```{r}
tif_files <- list.files(out_dir, pattern = "\\.tif$", full.names = FALSE)

# Parse filename components: {file_type}_{scenario}_{variable}_{year}.tif
parsed <- data.frame(
  filename  = tif_files,
  file_type = str_extract(tif_files, "^(states|management)"),
  scenario  = str_extract(tif_files, "(?<=^(states|management)_)(vl|h)"),
  year      = as.integer(str_extract(tif_files, "\\d{4}(?=\\.tif)"))
)

# Summary counts
summary_tbl <- as.data.frame(table(parsed$file_type, parsed$scenario))
names(summary_tbl) <- c("File type", "Scenario", "Count")

cat("Output folder:", out_dir, "\n")
cat("Total GeoTIFFs found:", length(tif_files), "\n\n")
cat("Expected per scenario:\n")
cat("  States     :", length(states_vars) * length(target_years), "files\n")
cat("  Management :", length(management_vars) * length(target_years), "files\n\n")
print(summary_tbl)

cat("\nThese GeoTIFFs are the inputs for scripts 02-05.\n")
```

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
