---
title: "Land Cover States — EBVcube creation"
subtitle: "LUH3 v1.1 | id105 (VL), id106 (H), id105-combined"
author: "Andres Marmol Guijarro"
date: "2026-03-25"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

## 1. Introduction

This script creates three EBVcube NetCDF files for the LUH3 v1.1 land
cover states dataset, covering land cover fractions for two climate
scenarios and their combination.

| Cube | Dataset ID | JSON | Scenario | IAM |
|---|---|---|---|---|
| `Popp_ecostr_id105_20260325.nc` | 105 | `105.json` | VL only | REMIND-MAgPIE |
| `Chini_ecostr_id106_20260325.nc` | 106 | `106.json` | H only | GCAM |
| `Popp&Chini_ecostr_id105_20260325.nc` | 105 | `105-2.json` | VL + H | REMIND-MAgPIE & GCAM |

Each cube contains:

- *1 metric*: land cover type (fraction 0–1)
- *13 entities*: one per land cover class (see section 5)
- *4 timesteps*: 2022, 2030, 2050, 2070

> **Note:** Secondary vegetation properties (`secma`, `secmb`) are
> extracted in script 01 but are not included in this cube, as they use
> different units (years and kg m⁻²) and are treated separately.

### 1.1 Overview of the workflow

This script follows four core processing stages:

1. *Prepare the working environment.*
   Load required packages, define paths, and set shared parameters.

2. *Define the spatial template.*
   Read one reference GeoTIFF to extract the spatial extent, resolution,
   and CRS used for all three cubes.

3. *Create and populate the three EBVcubes.*
   For each of the three cubes (VL, H, combined), create an empty
   NetCDF file from its JSON metadata, then insert all entity GeoTIFFs
   for every timestep.

4. *Summarise results.*
   Print a completion report for each cube.

### 1.2 Before you start

To ensure the workflow runs smoothly, the following conditions should
be met:

- Script `01_extract_tiffs_LUH3v1.1.Rmd` has been run and the
  GeoTIFFs are available in `raw/temp/`.
- The three JSON metadata files (`105.json`, `106.json`, `105-2.json`)
  are available in the `raw/` folder.
- The `ebvcube`, `terra`, and `rhdf5` packages are installed.
- The output folder `cubes/land_cover/` exists or can be created.

---

## 2. Load libraries

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages(c("ebvcube", "terra", "rhdf5", "stringr", "sf", "jsonlite"))

library(ebvcube)
library(terra)
library(jsonlite)
library(stringr)
library(sf)
library(rhdf5)
```

---

## 3. Define paths

All paths are defined relative to the script location so the workflow
is portable across machines.

```{r}
# Set working directory to the location of this script (RStudio only)
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

# Directories
raw_dir <- "../raw"          # JSON metadata files and source NetCDF files
pathin  <- "../raw/temp"     # Pre-extracted GeoTIFFs (output of script 01)
cube    <- "../cubes/land_cover"  # Output folder for EBVcube NetCDF files

dir.create(cube, recursive = TRUE, showWarnings = FALSE)

# JSON metadata files
json_vl       <- file.path(raw_dir, "105.json")
json_h        <- file.path(raw_dir, "106.json")
json_combined <- file.path(raw_dir, "105-2.json")

# Output EBVcube NetCDF files
cube_vl       <- file.path(cube, "Popp_ecostr_id105_20260325.nc")
cube_h        <- file.path(cube, "Chini_ecostr_id106_20260325.nc")
cube_combined <- file.path(cube, "Popp&Chini_ecostr_id105_20260325.nc")

# Verify inputs
cat("Input folder (GeoTIFFs) exists  :", dir.exists(pathin),        "\n")
cat("Output folder exists             :", dir.exists(cube),          "\n")
cat("json_vl       exists             :", file.exists(json_vl),      "\n")
cat("json_h        exists             :", file.exists(json_h),       "\n")
cat("json_combined exists             :", file.exists(json_combined), "\n")
```

---

## 4. Workflow settings

```{r}
# Target years and timestep labels
years     <- c("2022", "2030", "2050", "2070")
timesteps <- paste0(years, "-01-01")

# EBVcube technical parameters
epsg <- 4326       # WGS84
prec <- "float"    # Data precision
fv   <- -3.4e+38   # Fill value for missing data

cat("Years    :", paste(years, collapse = ", "), "\n")
cat("EPSG     :", epsg, "\n")
cat("Fill val :", fv, "\n")
```

---

## 5. Entities and spatial template

The 13 entities correspond to the harmonised land cover classes from
the LUH3 v1.1 states dataset. Secondary vegetation properties
(`secma`, `secmb`) are excluded here as they use different units.

```{r}
entities <- c(
  # Cropland
  "c3ann", "c3nfx", "c3per", "c4ann", "c4per",
  # Managed land
  "pastr", "range", "urban",
  # Natural vegetation
  "primf", "primn", "secdf", "secdn", "pltns"
)

cat("Entities (", length(entities), "):", paste(entities, collapse = ", "), "\n")
```

We read one reference GeoTIFF to extract the spatial extent, resolution,
and CRS that will be shared across all three cubes.

```{r}
template    <- rast(file.path(pathin, "states_vl_c3ann_2022.tif"))
cube_extent <- as.vector(ext(template))
cube_res    <- terra::res(template)

cat("Extent     :", cube_extent, "\n")
cat("Resolution :", cube_res, "\n")
cat("CRS        :", crs(template, describe = TRUE)$code, "\n")
```

---

## 6. REMIND-MAgPIE VL scenario — Popp_ecostr_id105_20260325.nc

### 6.1 Create empty NetCDF

An empty EBVcube NetCDF file is created from the JSON metadata. This
establishes the dataset structure (spatial extent, CRS, entities,
timesteps) without yet containing any data.

```{r}
ebv_create(
  jsonpath   = json_vl,
  outputpath = cube_vl,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (VL) ===\n")
print(ebv_datacubepaths(cube_vl, verbose = FALSE))
```

### 6.2 Populate the EBVcube — VL

Each entity GeoTIFF is inserted into the correct position in the cube,
looping over all timesteps × entities.

```{r}
created_vl <- character()
errors_vl  <- character()

s_no <- 1   # Single scenario in this cube
m_no <- 1   # Single metric

for (t_no in seq_along(years)) {
  yr <- years[t_no]
  cat("\n=== VL | Year:", yr, "===\n")

  for (e_no in seq_along(entities)) {
    ent     <- entities[e_no]
    tmp_tif <- file.path(pathin, paste0("states_vl_", ent, "_", yr, ".tif"))

    if (!file.exists(tmp_tif)) {
      msg <- paste("File not found:", basename(tmp_tif))
      errors_vl <- c(errors_vl, msg)
      warning(msg)
      next
    }

    res <- tryCatch({
      ebv_add_data(
        filepath_nc = cube_vl,
        scenario    = s_no,
        metric      = m_no,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = FALSE
      )
      basename(tmp_tif)
    }, error = function(e) e)

    if (inherits(res, "error")) {
      errors_vl <- c(errors_vl, paste("FAILED:", ent, yr, "|", conditionMessage(res)))
    } else {
      created_vl <- c(created_vl, res)
      cat("  ok", ent, "\n")
    }
  }
}

cat("\nVL — Inserted:", length(created_vl),
    "/ expected:", length(entities) * length(years),
    "| Failed:", length(errors_vl), "\n")
if (length(errors_vl) > 0) print(errors_vl)
```

---

## 7. GCAM H scenario — Chini_ecostr_id106_20260325.nc

### 7.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_h,
  outputpath = cube_h,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (H) ===\n")
print(ebv_datacubepaths(cube_h, verbose = FALSE))
```

### 7.2 Populate the EBVcube — H

```{r}
created_h <- character()
errors_h  <- character()

s_no <- 1
m_no <- 1

for (t_no in seq_along(years)) {
  yr <- years[t_no]
  cat("\n=== H | Year:", yr, "===\n")

  for (e_no in seq_along(entities)) {
    ent     <- entities[e_no]
    tmp_tif <- file.path(pathin, paste0("states_h_", ent, "_", yr, ".tif"))

    if (!file.exists(tmp_tif)) {
      msg <- paste("File not found:", basename(tmp_tif))
      errors_h <- c(errors_h, msg)
      warning(msg)
      next
    }

    res <- tryCatch({
      ebv_add_data(
        filepath_nc = cube_h,
        scenario    = s_no,
        metric      = m_no,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = FALSE
      )
      basename(tmp_tif)
    }, error = function(e) e)

    if (inherits(res, "error")) {
      errors_h <- c(errors_h, paste("FAILED:", ent, yr, "|", conditionMessage(res)))
    } else {
      created_h <- c(created_h, res)
      cat("  ok", ent, "\n")
    }
  }
}

cat("\nH — Inserted:", length(created_h),
    "/ expected:", length(entities) * length(years),
    "| Failed:", length(errors_h), "\n")
if (length(errors_h) > 0) print(errors_h)
```

---

## 8. Combined VL + H — Popp&Chini_ecostr_id105_20260325.nc

### 8.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_combined,
  outputpath = cube_combined,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (combined) ===\n")
print(ebv_datacubepaths(cube_combined, verbose = FALSE))
```

### 8.2 Populate the EBVcube — combined

The combined cube contains both scenarios as separate scenario slots
(VL = scenario 1, H = scenario 2).

```{r}
created_combined <- character()
errors_combined  <- character()

m_no <- 1

scenarios_combined <- list(
  vl = list(code = "vl", s_no = 1, label = "VL"),
  h  = list(code = "h",  s_no = 2, label = "H")
)

for (sc in scenarios_combined) {
  cat("\n==============================\n")
  cat("Scenario:", sc$label, "\n")
  cat("==============================\n")

  for (t_no in seq_along(years)) {
    yr <- years[t_no]
    cat("\n=== Combined |", sc$label, "| Year:", yr, "===\n")

    for (e_no in seq_along(entities)) {
      ent     <- entities[e_no]
      tmp_tif <- file.path(pathin,
                           paste0("states_", sc$code, "_", ent, "_", yr, ".tif"))

      if (!file.exists(tmp_tif)) {
        msg <- paste("File not found:", basename(tmp_tif))
        errors_combined <- c(errors_combined, msg)
        warning(msg)
        next
      }

      res <- tryCatch({
        ebv_add_data(
          filepath_nc = cube_combined,
          scenario    = sc$s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          data        = tmp_tif,
          band        = 1,
          ignore_RAM  = FALSE,
          raw         = FALSE,
          verbose     = FALSE
        )
        basename(tmp_tif)
      }, error = function(e) e)

      if (inherits(res, "error")) {
        errors_combined <- c(errors_combined,
                             paste("FAILED:", sc$label, ent, yr, "|",
                                   conditionMessage(res)))
      } else {
        created_combined <- c(created_combined, res)
        cat("  ok", ent, "\n")
      }
    }
  }
  gc()
}

cat("\nCombined — Inserted:", length(created_combined),
    "/ expected:", length(entities) * length(years) * 2,
    "| Failed:", length(errors_combined), "\n")
if (length(errors_combined) > 0) print(errors_combined)
```

---

## 9. Summary

```{r}
expected_single   <- length(entities) * length(years)
expected_combined <- expected_single * 2

cat("==============================\n")
cat("VL cube      :", cube_vl, "\n")
cat("  Inserted:", length(created_vl), "/ expected:", expected_single,
    "| failed:", length(errors_vl), "\n\n")

cat("H cube       :", cube_h, "\n")
cat("  Inserted:", length(created_h), "/ expected:", expected_single,
    "| failed:", length(errors_h), "\n\n")

cat("Combined cube:", cube_combined, "\n")
cat("  Inserted:", length(created_combined), "/ expected:", expected_combined,
    "| failed:", length(errors_combined), "\n")
cat("==============================\n")
```

---

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
