---
title: "ScenMIP7-LUH3 v1.1 Land-Cover States — EBVcube creation"
subtitle: "id105 — VL and H scenarios"
author: "Emmanuel Oceguera-Conchas, Andres Marmol-Guijarro, Magali Weissgerber"
date: 2026-03-25
---

## 1. Introduction

This workflow processes the LUH3 v1.1 land-cover states dataset, holding both climate scenarios VL (Very low emissions) and H (High emissions) and converts it into an EBV Cube NetCDF file (dataset ID 105)

The land-cover data are the per-class area fractions extracted in script 01. The cube holds a single metric (land-cover fraction, 0–1) across 13 entities (one per land-cover class), four time steps (2022, 2030, 2050, 2070), and two scenarios; VL and H.

The workflow comprises two core steps:

-   *Cube creation:* An empty NetCDF file is built from the JSON metadata, establishing the spatial extent, CRS, entities, time steps, and scenarios.
-   *Population:* Each entity's GeoTIFF is inserted at its scenario / metric / time step / entity position, for both scenarios.

In the notebook, these steps span three sections: preparing the working environment (loading `ebvcube` and `terra`, setting paths and parameters), defining the spatial template, and creating and populating the cube.

> **Note:** Secondary-vegetation properties (`secma`, `secmb`) are extracted in script 01 but are not included in this cube, as they use different units (years and kg m⁻²) and are treated separately.

### 1.1 Scenarios and time steps

The cube holds two scenarios across four years each.

| Scenario | Code | IAM | Description | Available Years |
|---------------|---------------|---------------|---------------|---------------|
| Low-MAgPIE | `vl` | REMIND-MAgPIE | Very low emissions, land-use change from REMIND-MAgPIE | 2022, 2030, 2050, 2070 |
| H-GCAM | `h` | GCAM | High emissions, land-use change from GCAM | 2022, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   Script 01 has been run and the GeoTIFFs are available in `raw/temp/`
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   The `ebvcube` and `terra` packages are installed

## 2. Load libraries

Here we load the packages used throughout the workflow: `ebvcube` for creating and filling the cube, and `terra` for reading the reference raster.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages(c("ebvcube", "terra"))

library(ebvcube)
library(terra)
```

## 3. Configure temporary raster storage

Reading the input rasters can create temporary files, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths

Input: the pre-extracted GeoTIFFs from script 01 and the JSON metadata file. Output: the EBV Cube NetCDF file. Paths are relative to this script's location.

```{r}
# Directories
raw_dir <- "../raw"               # JSON metadata file and source NetCDF files
pathin  <- "../raw/temp"          # Pre-extracted GeoTIFFs (output of script 01)
cube    <- "../BES-SIM 2/ScenMIP7-LUH3/cube"              # Output folder for the EBV Cube NetCDF file
dir.create(cube, recursive = TRUE, showWarnings = FALSE)


cat("Input folder (GeoTIFFs):", pathin, "\n")
cat("Output folder          :", cube,   "\n")
```

## 5. Define the main workflow settings

Set the scenarios, time steps, the cube's technical parameters, and the 13 entities. The entities are the harmonised land-cover classes from the LUH3 v1.1 states dataset, excluding the secondary-vegetation properties.

```{r}
# Scenario codes (order must match the metadata JSON: 1 VL, 2 H)
scenarios <- c("vl", "h")

# Target years (time steps)
years <- c("2022", "2030", "2050", "2070")

# EBV Cube technical parameters
epsg <- 4326       # WGS84
prec <- "float"    # Data precision
fv   <- -3.4e+38   # Fill value for missing data

# Entities = land-cover classes (order must match the metadata JSON)
entities <- c(
  # Cropland
  "c3ann", "c3nfx", "c3per", "c4ann", "c4per",
  # Managed land
  "pastr", "range", "urban",
  # Natural vegetation
  "primf", "primn", "secdf", "secdn", "pltns"
)

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Years    :", paste(years, collapse = ", "), "\n")
cat("EPSG     :", epsg, "\n")
cat("Fill val :", fv, "\n")
cat("Entities (", length(entities), "):", paste(entities, collapse = ", "), "\n")
```

## 6. Define the spatial template

Read one reference GeoTIFF to extract the spatial extent, resolution, and CRS used for the cube.

```{r}
template    <- rast(file.path(pathin, "states_vl_c3ann_2022.tif"))
cube_extent <- as.vector(ext(template))
cube_res    <- terra::res(template)

cat("Extent     :", cube_extent, "\n")
cat("Resolution :", cube_res, "\n")
cat("CRS        :", crs(template, describe = TRUE)$code, "\n")
```

## 7. Create EBV Cube NetCDF file

### 7.1 Check paths and available rasters

Define the metadata JSON and the target NetCDF file, and confirm the GeoTIFFs from script 01 are present in the expected folder. This check catches missing inputs before the cube is built.

```{r}
# Define date and check id
id            <- 105
creation_date <- "20260325"
cat("Created date:", format(as.Date(creation_date, format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Metadata JSON
js <- file.path(raw_dir, "metadata.json")

# Note:
# Please follow the naming convention
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- file.path(cube, paste0("Popp&Chini_ecostr_id", id, "_", creation_date, ".nc"))

cat("JSON metadata exists      :", file.exists(js), "\n")
cat("EBV Cube NetCDF file name :", basename(nc),    "\n")
cat("Input GeoTIFF folder      :", pathin,          "\n")

# List all input GeoTIFFs
tif_files <- list.files(pathin, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)

cat("Total input GeoTIFF files found:", length(tif_files), "\n")
```

### 7.2 Create empty NetCDF file

Create the empty NetCDF file from the JSON metadata. It holds the cube structure (spatial extent, CRS, entities, time steps, scenarios) but no raster data yet — that comes in [Section 8](#sec-8.-populate-the-ebv-cube).

```{r}
ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = years,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure ===\n")
print(ebv_datacubepaths(nc, verbose = FALSE))
```

## 8. Populate the EBV Cube {#sec-8.-populate-the-ebv-cube}

Insert every entity GeoTIFF at its correct scenario / metric / time step / entity position, looping over both scenarios.

```{r}
m_no     <- 1   # single metric (land-cover fraction)
inserted <- 0L

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  code <- scenarios[s_no]

  for (t_no in seq_along(years)) {

    yr <- years[t_no]

    for (e_no in seq_along(entities)) {

      ent     <- entities[e_no]
      in_file <- file.path(pathin, sprintf("states_%s_%s_%s.tif", code, ent, yr))

      if (!file.exists(in_file)) {
        warning("Missing file, skipped: ", in_file)
        next
      }

      ebv_add_data(
        filepath_nc = nc,
        data        = in_file,
        band        = 1,
        scenario    = s_no,
        metric      = m_no,
        entity      = e_no,
        timestep    = t_no,
        ignore_RAM  = FALSE,
        raw         = FALSE,
        verbose     = FALSE
      )
      inserted <- inserted + 1L
    }
  }
  cat("  ✔ inserted scenario", code, "\n")
}

end.time <- Sys.time()
cat("\nDone. Time:", round(difftime(end.time, start.time, units = "mins"), 2), "min\n")
```

## 9. Validation and visualization

A quick check that the cube is readable and that the scenarios, entities, and time steps line up.

```{r}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 1,
        timestep = 1, verbose = FALSE)
```

## 10. Summary

```{r}
expected <- length(scenarios) * length(years) * length(entities)

cat("==============================\n")
cat("Cube     :", basename(nc), "\n")
cat("Inserted :", inserted, "/ expected:", expected, "\n")
cat("==============================\n")
```

------------------------------------------------------------------------

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
