---
title: "Forestry Management — EBVcube creation"
subtitle: "LUH3 v1.1 | id107 (VL, H, combined)"
author: "Andres Marmol Guijarro"
date: "2026-03-25"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

## 1. Introduction

This script creates three EBVcube NetCDF files for the LUH3 v1.1
forestry management dataset. The cubes describe how harvested wood
biomass carbon is allocated across different forest use types for two
climate scenarios and their combination.

| Cube | JSON | Scenario | IAM |
|---|---|---|---|
| `Popp_ecoser_id107vl_20260325.nc` | `id107_vl.json` | VL only | REMIND-MAgPIE |
| `Chini_ecoser_id107h_20260325.nc` | `id107_h.json` | H only | GCAM |
| `Popp&Chini_ecoser_id107_20260325.nc` | `id107_combined.json` | VL + H | REMIND-MAgPIE & GCAM |

Each cube contains:

- *1 metric*: allocation of harvested wood biomass carbon (fraction 0–1)
- *5 entities*: `rndwd`, `fulwd`, `combf`, `pltns_bfuel`, `pltns_wdprd`
- *4 timesteps*: 2022, 2030, 2050, 2070

> **Note:** Plantation variables (`pltns_bfuel`, `pltns_wdprd`) are
> included as entities but are spatially empty in LUH3 v1.1. They are
> inserted using fill values. The variable `addtc` (added tree cover,
> afforestation) is excluded from this cube because it uses a different
> unit (fraction of grid cell) and is zero in LUH3 v1.1 for both
> scenarios.

### 1.1 Overview of the workflow

1. *Prepare the working environment.*
   Load required packages, define paths, and set shared parameters.

2. *Define the spatial template and empty-raster helper.*
   Read a reference GeoTIFF and define a helper function for inserting
   fill-value rasters in place of missing entities.

3. *Create and populate the three EBVcubes.*
   For each cube (VL, H, combined), create an empty NetCDF file from
   its JSON metadata and insert all entity GeoTIFFs.

4. *Summarise results.*

### 1.2 Before you start

- Script `01_extract_tiffs_LUH3v1.1.Rmd` has been run and the
  GeoTIFFs are available in `raw/temp/`.
- The three JSON metadata files (`id107_vl.json`, `id107_h.json`,
  `id107_combined.json`) are available in the `raw/` folder.
- The `ebvcube`, `terra`, and `rhdf5` packages are installed.

---

## 2. Load libraries

```{r}
rm(list = ls())
gc()

library(ebvcube)
library(terra)
library(jsonlite)
library(stringr)
library(sf)
library(rhdf5)
```

---

## 3. Define paths

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

raw_dir <- "../raw"
pathin  <- "../raw/temp"
cube    <- "../cubes/forest_management"

dir.create(cube, recursive = TRUE, showWarnings = FALSE)

json_vl       <- file.path(raw_dir, "id107_vl.json")
json_h        <- file.path(raw_dir, "id107_h.json")
json_combined <- file.path(raw_dir, "id107_combined.json")

cube_vl       <- file.path(cube, "Popp_ecoser_id107vl_20260325.nc")
cube_h        <- file.path(cube, "Chini_ecoser_id107h_20260325.nc")
cube_combined <- file.path(cube, "Popp&Chini_ecoser_id107_20260325.nc")

cat("Input folder exists   :", dir.exists(pathin),        "\n")
cat("Output folder exists  :", dir.exists(cube),          "\n")
cat("json_vl       exists  :", file.exists(json_vl),      "\n")
cat("json_h        exists  :", file.exists(json_h),       "\n")
cat("json_combined exists  :", file.exists(json_combined), "\n")
```

---

## 4. Workflow settings

```{r}
# 5 entities — wood harvest allocation types
# pltns_bfuel and pltns_wdprd are spatially empty in LUH3 v1.1
entities <- c("rndwd", "fulwd", "combf", "pltns_bfuel", "pltns_wdprd")

years     <- c("2022", "2030", "2050", "2070")
timesteps <- paste0(years, "-01-01")

epsg <- 4326
prec <- "float"
fv   <- -3.4e+38

# Entities that have no GeoTIFF data in LUH3 v1.1 (plantation variables)
empty_entities <- c("pltns_bfuel", "pltns_wdprd")

cat("Entities (", length(entities), "):", paste(entities, collapse = ", "), "\n")
cat("Empty entities         :", paste(empty_entities, collapse = ", "), "\n")
```

---

## 5. Spatial template and empty-raster helper

We read one reference GeoTIFF to extract the shared spatial properties,
and define a helper function for creating fill-value rasters when no
source GeoTIFF is available for an entity.

```{r}
template    <- rast(file.path(pathin, "management_vl_rndwd_2022.tif"))
cube_extent <- as.vector(ext(template))
cube_res    <- terra::res(template)

cat("Extent    :", cube_extent, "\n")
cat("Resolution:", cube_res, "\n")
cat("CRS       :", crs(template, describe = TRUE)$code, "\n")
```

```{r}
# Helper: create a fill-value raster for entities with no source data.
# Writes a temporary file and returns the path for use with ebv_add_data
# (raw = TRUE bypasses scale/offset transformation).
make_empty <- function(template, sc_code, entity, yr, fv) {
  empty         <- template
  values(empty) <- fv
  out_path      <- file.path(tempdir(),
                             paste0("empty_", sc_code, "_", entity, "_", yr, ".tif"))
  writeRaster(empty, out_path, overwrite = TRUE,
              gdal = c("COMPRESS=LZW", "TILED=YES"))
  out_path
}
```

---

## 6. REMIND-MAgPIE VL scenario — Popp_ecoser_id107vl_20260325.nc

### 6.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_vl,
  outputpath = cube_vl,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (VL) ===\n")
print(ebv_datacubepaths(cube_vl, verbose = FALSE))
```

### 6.2 Populate the EBVcube — VL

Plantation entities (`pltns_bfuel`, `pltns_wdprd`) have no source data
in LUH3 v1.1 and are filled with the cube fill value.

```{r}
created_vl <- character()
errors_vl  <- character()

s_no <- 1
m_no <- 1

for (t_no in seq_along(years)) {
  yr <- years[t_no]
  cat("\n=== VL | Year:", yr, "===\n")

  for (e_no in seq_along(entities)) {
    ent <- entities[e_no]

    if (ent %in% empty_entities) {
      # Plantation variables — insert fill-value raster
      tmp_tif <- make_empty(template, "vl", ent, yr, fv)
      use_raw <- TRUE
    } else {
      tmp_tif <- file.path(pathin,
                           paste0("management_vl_", ent, "_", yr, ".tif"))
      use_raw <- FALSE
    }

    if (!use_raw && !file.exists(tmp_tif)) {
      msg <- paste("File not found:", basename(tmp_tif))
      errors_vl <- c(errors_vl, msg)
      warning(msg)
      next
    }

    res <- tryCatch({
      ebv_add_data(
        filepath_nc = cube_vl,
        scenario    = s_no,
        metric      = m_no,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        ignore_RAM  = FALSE,
        raw         = use_raw,
        verbose     = FALSE
      )
      basename(tmp_tif)
    }, error = function(e) e)

    if (inherits(res, "error")) {
      errors_vl <- c(errors_vl, paste("FAILED:", ent, yr, "|", conditionMessage(res)))
    } else {
      created_vl <- c(created_vl, res)
      cat("  ok", ent, "\n")
    }
  }
}

cat("\nVL — Inserted:", length(created_vl),
    "/ expected:", length(entities) * length(years),
    "| Failed:", length(errors_vl), "\n")
if (length(errors_vl) > 0) print(errors_vl)
```

---

## 7. GCAM H scenario — Chini_ecoser_id107h_20260325.nc

### 7.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_h,
  outputpath = cube_h,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (H) ===\n")
print(ebv_datacubepaths(cube_h, verbose = FALSE))
```

### 7.2 Populate the EBVcube — H

```{r}
created_h <- character()
errors_h  <- character()

s_no <- 1
m_no <- 1

for (t_no in seq_along(years)) {
  yr <- years[t_no]
  cat("\n=== H | Year:", yr, "===\n")

  for (e_no in seq_along(entities)) {
    ent <- entities[e_no]

    if (ent %in% empty_entities) {
      tmp_tif <- make_empty(template, "h", ent, yr, fv)
      use_raw <- TRUE
    } else {
      tmp_tif <- file.path(pathin,
                           paste0("management_h_", ent, "_", yr, ".tif"))
      use_raw <- FALSE
    }

    if (!use_raw && !file.exists(tmp_tif)) {
      msg <- paste("File not found:", basename(tmp_tif))
      errors_h <- c(errors_h, msg)
      warning(msg)
      next
    }

    res <- tryCatch({
      ebv_add_data(
        filepath_nc = cube_h,
        scenario    = s_no,
        metric      = m_no,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        ignore_RAM  = FALSE,
        raw         = use_raw,
        verbose     = FALSE
      )
      basename(tmp_tif)
    }, error = function(e) e)

    if (inherits(res, "error")) {
      errors_h <- c(errors_h, paste("FAILED:", ent, yr, "|", conditionMessage(res)))
    } else {
      created_h <- c(created_h, res)
      cat("  ok", ent, "\n")
    }
  }
}

cat("\nH — Inserted:", length(created_h),
    "/ expected:", length(entities) * length(years),
    "| Failed:", length(errors_h), "\n")
if (length(errors_h) > 0) print(errors_h)
```

---

## 8. Combined VL + H — Popp&Chini_ecoser_id107_20260325.nc

### 8.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_combined,
  outputpath = cube_combined,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (combined) ===\n")
print(ebv_datacubepaths(cube_combined, verbose = FALSE))
```

### 8.2 Populate the EBVcube — combined

```{r}
created_combined <- character()
errors_combined  <- character()

m_no <- 1

scenarios_combined <- list(
  vl = list(code = "vl", s_no = 1, label = "VL"),
  h  = list(code = "h",  s_no = 2, label = "H")
)

for (sc in scenarios_combined) {
  cat("\n==============================\n")
  cat("Scenario:", sc$label, "\n")
  cat("==============================\n")

  for (t_no in seq_along(years)) {
    yr <- years[t_no]
    cat("\n=== Combined |", sc$label, "| Year:", yr, "===\n")

    for (e_no in seq_along(entities)) {
      ent <- entities[e_no]

      if (ent %in% empty_entities) {
        tmp_tif <- make_empty(template, sc$code, ent, yr, fv)
        use_raw <- TRUE
      } else {
        tmp_tif <- file.path(pathin,
                             paste0("management_", sc$code, "_", ent, "_", yr, ".tif"))
        use_raw <- FALSE
      }

      if (!use_raw && !file.exists(tmp_tif)) {
        msg <- paste("File not found:", basename(tmp_tif))
        errors_combined <- c(errors_combined, msg)
        warning(msg)
        next
      }

      res <- tryCatch({
        ebv_add_data(
          filepath_nc = cube_combined,
          scenario    = sc$s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          data        = tmp_tif,
          band        = 1,
          ignore_RAM  = FALSE,
          raw         = use_raw,
          verbose     = FALSE
        )
        basename(tmp_tif)
      }, error = function(e) e)

      if (inherits(res, "error")) {
        errors_combined <- c(errors_combined,
                             paste("FAILED:", sc$label, ent, yr, "|",
                                   conditionMessage(res)))
      } else {
        created_combined <- c(created_combined, res)
        cat("  ok", ent, "\n")
      }
    }
  }
  gc()
}

cat("\nCombined — Inserted:", length(created_combined),
    "/ expected:", length(entities) * length(years) * 2,
    "| Failed:", length(errors_combined), "\n")
if (length(errors_combined) > 0) print(errors_combined)
```

---

## 9. Summary

```{r}
expected_single   <- length(entities) * length(years)
expected_combined <- expected_single * 2

cat("==============================\n")
cat("VL cube      :", cube_vl, "\n")
cat("  Inserted:", length(created_vl), "/ expected:", expected_single,
    "| failed:", length(errors_vl), "\n\n")

cat("H cube       :", cube_h, "\n")
cat("  Inserted:", length(created_h), "/ expected:", expected_single,
    "| failed:", length(errors_h), "\n\n")

cat("Combined cube:", cube_combined, "\n")
cat("  Inserted:", length(created_combined), "/ expected:", expected_combined,
    "| failed:", length(errors_combined), "\n")
cat("==============================\n")
```

---

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
