---
title: "Crop Management Intensity — EBVcube creation"
subtitle: "LUH3 v1.1 | id108 (VL, H, combined)"
author: "Andres Marmol Guijarro"
date: "2026-03-30"
output: html_document
editor_options:
  markdown:
    wrap: 72
---

## 1. Introduction

This script creates three EBVcube NetCDF files for the LUH3 v1.1 crop
management intensity dataset. The cubes describe the intensity of
agricultural management across different crop functional types for two
climate scenarios and their combination.

| Cube | JSON | Scenario | IAM |
|---|---|---|---|
| `popp_ecoser_id108vl_20260330.nc` | `id108_vl.json` | VL only | REMIND-MAgPIE |
| `chini_ecoser_id108h_20260410.nc` | `id108_h.json` | H only | GCAM |
| `popp&chini_ecoser_id108_20260410.nc` | `id108_combined.json` | VL + H | REMIND-MAgPIE & GCAM |

Each cube contains:

- *4 metrics*: fertilization (`fertl`), irrigation (`irrig`),
  1st-generation biofuel fraction (`cpbf1`), 2nd-generation biofuel
  fraction (`cpbf2`)
- *6 entities*: `c3ann`, `c3nfx`, `c3per`, `c4ann`, `c4per`, `flood`
- *4 timesteps*: 2022, 2030, 2050, 2070

> **Note on the `flood` entity:** The `flood` variable describes the
> fraction of C3 annual crops under permanent flooding (paddy rice). It
> is only meaningful for the irrigation metric (`irrig`). For
> fertilization (`fertl`) and biofuel metrics (`cpbf1`, `cpbf2`), the
> `flood` entity is inserted as a fill-value raster.
>
> **Note on the `flood` GeoTIFF path:** The flood variable was extracted
> from the raw NetCDF under its own name `flood`, so its GeoTIFF is
> `management_{sc}_flood_{year}.tif` — not prefixed with `irrig_`.
>
> **Note on biofuel variables:** `cpbf1` and `cpbf2` are zero across the
> entire globe for both scenarios in LUH3 v1.1. The cubes are built for
> structural completeness and will be updated in future dataset versions.

### 1.1 Overview of the workflow

1. *Prepare the working environment.*
2. *Define the spatial template and empty-raster helper.*
3. *Create and populate three EBVcubes* (VL, H, combined), looping
   over metrics × timesteps × entities for each.
4. *Summarise results.*

### 1.2 Before you start

- Script `01_extract_tiffs_LUH3v1.1.Rmd` has been run successfully.
- JSON files (`id108_vl.json`, `id108_h.json`, `id108_combined.json`)
  are available in the `raw/` folder.
- The `ebvcube`, `terra`, and `rhdf5` packages are installed.

---

## 2. Load libraries

```{r}
rm(list = ls())
gc()

library(ebvcube)
library(terra)
library(jsonlite)
library(stringr)
library(sf)
library(rhdf5)
```

---

## 3. Define paths

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

raw_dir <- "../raw"
pathin  <- "../raw/temp"
cube    <- "../cubes/crop_managment_intensity"

dir.create(cube, recursive = TRUE, showWarnings = FALSE)

json_vl       <- file.path(raw_dir, "id108_vl.json")
json_h        <- file.path(raw_dir, "id108_h.json")
json_combined <- file.path(raw_dir, "id108_combined.json")

cube_vl       <- file.path(cube, "popp_ecoser_id108vl_20260330.nc")
cube_h        <- file.path(cube, "chini_ecoser_id108h_20260410.nc")
cube_combined <- file.path(cube, "popp&chini_ecoser_id108_20260410.nc")

cat("Input folder exists   :", dir.exists(pathin),        "\n")
cat("Output folder exists  :", dir.exists(cube),          "\n")
cat("json_vl       exists  :", file.exists(json_vl),      "\n")
cat("json_h        exists  :", file.exists(json_h),       "\n")
cat("json_combined exists  :", file.exists(json_combined), "\n")
```

---

## 4. Workflow settings

```{r}
# Entities — 6 crop functional types
entities <- c("c3ann", "c3nfx", "c3per", "c4ann", "c4per", "flood")

# Metrics — 4 management intensity types
# The metric prefix is used to construct GeoTIFF filenames
metrics <- list(
  list(metric = "fertl", label = "Fertilization"),
  list(metric = "irrig", label = "Irrigation"),
  list(metric = "cpbf1", label = "Biofuel 1st generation"),
  list(metric = "cpbf2", label = "Biofuel 2nd generation")
)

years     <- c("2022", "2030", "2050", "2070")
timesteps <- paste0(years, "-01-01")

epsg <- 4326
prec <- "float"
fv   <- -3.4e+38

cat("Entities (", length(entities), "):", paste(entities, collapse = ", "), "\n")
cat("Metrics  (", length(metrics),  "):",
    paste(sapply(metrics, `[[`, "label"), collapse = ", "), "\n")
```

---

## 5. Spatial template and empty-raster helper

```{r}
template    <- rast(file.path(pathin, "management_vl_irrig_c3ann_2022.tif"))
cube_extent <- as.vector(ext(template))
cube_res    <- terra::res(template)

cat("Extent    :", cube_extent, "\n")
cat("Resolution:", cube_res, "\n")
cat("CRS       :", crs(template, describe = TRUE)$code, "\n")
```

```{r}
# Helper: creates a fill-value raster for entity/metric combinations
# that have no source GeoTIFF (e.g. flood entity for fertl, cpbf1, cpbf2).
# Uses raw = TRUE when inserted, bypassing scale/offset transformation.
make_empty <- function(template, sc_code, metric, entity, yr, fv) {
  empty         <- template
  values(empty) <- fv
  out_path      <- file.path(tempdir(),
                             paste0("empty_", sc_code, "_", metric,
                                    "_", entity, "_", yr, ".tif"))
  writeRaster(empty, out_path, overwrite = TRUE,
              gdal = c("COMPRESS=LZW", "TILED=YES"))
  out_path
}
```

---

## 6. REMIND-MAgPIE VL scenario — popp_ecoser_id108vl_20260330.nc

### 6.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_vl,
  outputpath = cube_vl,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (VL) ===\n")
print(ebv_datacubepaths(cube_vl, verbose = FALSE))
```

### 6.2 Populate the EBVcube — VL

The loop runs over all 4 metrics × 4 years × 6 entities. The `flood`
entity is filled with the fill value for all metrics except irrigation.

```{r}
created_vl <- character()
errors_vl  <- character()

s_no <- 1

for (m_no in seq_along(metrics)) {
  m <- metrics[[m_no]]
  cat("\n==============================\n")
  cat("Metric", m_no, "—", m$label, "\n")
  cat("==============================\n")

  for (t_no in seq_along(years)) {
    yr <- years[t_no]
    cat("\n=== VL |", m$label, "| Year:", yr, "===\n")

    for (e_no in seq_along(entities)) {
      ent <- entities[e_no]

      # flood entity: only meaningful for irrigation
      # — insert fill value for all other metrics
      if (ent == "flood" && m$metric != "irrig") {
        tmp_tif <- make_empty(template, "vl", m$metric, ent, yr, fv)
        use_raw <- TRUE
      } else if (ent == "flood") {
        # flood GeoTIFF has no metric prefix in its filename
        tmp_tif <- file.path(pathin, paste0("management_vl_flood_", yr, ".tif"))
        use_raw <- FALSE
      } else {
        tmp_tif <- file.path(pathin,
                             paste0("management_vl_", m$metric, "_", ent, "_", yr, ".tif"))
        use_raw <- FALSE
      }

      if (!use_raw && !file.exists(tmp_tif)) {
        msg <- paste("File not found:", basename(tmp_tif))
        errors_vl <- c(errors_vl, msg)
        warning(msg)
        next
      }

      res <- tryCatch({
        ebv_add_data(
          filepath_nc = cube_vl,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          data        = tmp_tif,
          band        = 1,
          ignore_RAM  = FALSE,
          raw         = use_raw,
          verbose     = FALSE
        )
        basename(tmp_tif)
      }, error = function(e) e)

      if (inherits(res, "error")) {
        errors_vl <- c(errors_vl,
                       paste("FAILED:", m$label, ent, yr, "|", conditionMessage(res)))
      } else {
        created_vl <- c(created_vl, res)
        cat("  ok", ent, "\n")
      }
    }
  }
  gc()
}

cat("\nVL — Inserted:", length(created_vl),
    "/ expected:", length(metrics) * length(years) * length(entities),
    "| Failed:", length(errors_vl), "\n")
if (length(errors_vl) > 0) print(errors_vl)
```

---

## 7. GCAM H scenario — chini_ecoser_id108h_20260410.nc

### 7.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_h,
  outputpath = cube_h,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (H) ===\n")
print(ebv_datacubepaths(cube_h, verbose = FALSE))
```

### 7.2 Populate the EBVcube — H

```{r}
created_h <- character()
errors_h  <- character()

s_no <- 1

for (m_no in seq_along(metrics)) {
  m <- metrics[[m_no]]
  cat("\n==============================\n")
  cat("Metric", m_no, "—", m$label, "\n")
  cat("==============================\n")

  for (t_no in seq_along(years)) {
    yr <- years[t_no]
    cat("\n=== H |", m$label, "| Year:", yr, "===\n")

    for (e_no in seq_along(entities)) {
      ent <- entities[e_no]

      if (ent == "flood" && m$metric != "irrig") {
        tmp_tif <- make_empty(template, "h", m$metric, ent, yr, fv)
        use_raw <- TRUE
      } else if (ent == "flood") {
        tmp_tif <- file.path(pathin, paste0("management_h_flood_", yr, ".tif"))
        use_raw <- FALSE
      } else {
        tmp_tif <- file.path(pathin,
                             paste0("management_h_", m$metric, "_", ent, "_", yr, ".tif"))
        use_raw <- FALSE
      }

      if (!use_raw && !file.exists(tmp_tif)) {
        msg <- paste("File not found:", basename(tmp_tif))
        errors_h <- c(errors_h, msg)
        warning(msg)
        next
      }

      res <- tryCatch({
        ebv_add_data(
          filepath_nc = cube_h,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          data        = tmp_tif,
          band        = 1,
          ignore_RAM  = FALSE,
          raw         = use_raw,
          verbose     = FALSE
        )
        basename(tmp_tif)
      }, error = function(e) e)

      if (inherits(res, "error")) {
        errors_h <- c(errors_h,
                      paste("FAILED:", m$label, ent, yr, "|", conditionMessage(res)))
      } else {
        created_h <- c(created_h, res)
        cat("  ok", ent, "\n")
      }
    }
  }
  gc()
}

cat("\nH — Inserted:", length(created_h),
    "/ expected:", length(metrics) * length(years) * length(entities),
    "| Failed:", length(errors_h), "\n")
if (length(errors_h) > 0) print(errors_h)
```

---

## 8. Combined VL + H — popp&chini_ecoser_id108_20260410.nc

### 8.1 Create empty NetCDF

```{r}
ebv_create(
  jsonpath   = json_combined,
  outputpath = cube_combined,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = timesteps,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure (combined) ===\n")
print(ebv_datacubepaths(cube_combined, verbose = FALSE))
```

### 8.2 Populate the EBVcube — combined

```{r}
created_combined <- character()
errors_combined  <- character()

scenarios_combined <- list(
  vl = list(code = "vl", s_no = 1, label = "VL"),
  h  = list(code = "h",  s_no = 2, label = "H")
)

for (sc in scenarios_combined) {
  cat("\n==============================\n")
  cat("Scenario:", sc$label, "\n")
  cat("==============================\n")

  for (m_no in seq_along(metrics)) {
    m <- metrics[[m_no]]
    cat("\n--- Metric", m_no, "—", m$label, "---\n")

    for (t_no in seq_along(years)) {
      yr <- years[t_no]
      cat("\n=== Combined |", sc$label, "|", m$label, "| Year:", yr, "===\n")

      for (e_no in seq_along(entities)) {
        ent <- entities[e_no]

        if (ent == "flood" && m$metric != "irrig") {
          tmp_tif <- make_empty(template, sc$code, m$metric, ent, yr, fv)
          use_raw <- TRUE
        } else if (ent == "flood") {
          tmp_tif <- file.path(pathin,
                               paste0("management_", sc$code, "_flood_", yr, ".tif"))
          use_raw <- FALSE
        } else {
          tmp_tif <- file.path(pathin,
                               paste0("management_", sc$code, "_",
                                      m$metric, "_", ent, "_", yr, ".tif"))
          use_raw <- FALSE
        }

        if (!use_raw && !file.exists(tmp_tif)) {
          msg <- paste("File not found:", basename(tmp_tif))
          errors_combined <- c(errors_combined, msg)
          warning(msg)
          next
        }

        res <- tryCatch({
          ebv_add_data(
            filepath_nc = cube_combined,
            scenario    = sc$s_no,
            metric      = m_no,
            entity      = e_no,
            timestep    = t_no,
            data        = tmp_tif,
            band        = 1,
            ignore_RAM  = FALSE,
            raw         = use_raw,
            verbose     = FALSE
          )
          basename(tmp_tif)
        }, error = function(e) e)

        if (inherits(res, "error")) {
          errors_combined <- c(errors_combined,
                               paste("FAILED:", sc$label, m$label, ent, yr, "|",
                                     conditionMessage(res)))
        } else {
          created_combined <- c(created_combined, res)
          cat("  ok", ent, "\n")
        }
      }
    }
  }
  gc()
}

cat("\nCombined — Inserted:", length(created_combined),
    "/ expected:", length(metrics) * length(years) * length(entities) * 2,
    "| Failed:", length(errors_combined), "\n")
if (length(errors_combined) > 0) print(errors_combined)
```

---

## 9. Summary

```{r}
expected_single   <- length(metrics) * length(years) * length(entities)
expected_combined <- expected_single * 2

cat("==============================\n")
cat("VL cube      :", cube_vl, "\n")
cat("  Inserted:", length(created_vl), "/ expected:", expected_single,
    "| failed:", length(errors_vl), "\n\n")

cat("H cube       :", cube_h, "\n")
cat("  Inserted:", length(created_h), "/ expected:", expected_single,
    "| failed:", length(errors_h), "\n\n")

cat("Combined cube:", cube_combined, "\n")
cat("  Inserted:", length(created_combined), "/ expected:", expected_combined,
    "| failed:", length(errors_combined), "\n")
cat("==============================\n")
```

---

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
