---
title: "ScenMIP7-LUH3 v1.1 Crop Management Intensity — EBVcube creation"
subtitle: "id107 — VL and H scenarios"
author: "Emmanuel Oceguera-Conchas, Andres Marmol-Guijarro, Magali Weissgerber"
date: 2026-03-30
---

## 1. Introduction

This workflow processes the LUH3 v1.1 crop management intensity dataset, holding both climate scenarios VL (Very low emissions) and H (High emissions) and converts it into an EBV Cube NetCDF file (dataset ID 107)

The crop-management data describe the intensity of agricultural management across crop functional types, extracted in script 01. The cube holds four metrics (fertilization, irrigation, 1st-generation biofuel fraction, and 2nd-generation biofuel fraction) across 6 entities (one per crop functional type), four time steps (2022, 2030, 2050, 2070), and two scenarios; VL and H.

The workflow comprises two core steps:

-   *Cube creation:* An empty NetCDF file is built from the JSON metadata, establishing the spatial extent, CRS, entities, metrics, time steps, and scenarios.
-   *Population:* Each entity's GeoTIFF is inserted at its scenario / metric / time step / entity position, for both scenarios.

In the notebook, these steps span three sections: preparing the working environment (loading `ebvcube` and `terra`, setting paths and parameters), defining the spatial template, and creating and populating the cube.

> **Note on the `flood` entity:** The `flood` variable describes the fraction of C3 annual crops under permanent flooding (paddy rice). It is only meaningful for the irrigation metric (`irrig`); for the fertilization and biofuel metrics it is left at the cube's fill value. Its GeoTIFF is named `management_{sc}_flood_{year}.tif` (no metric prefix), unlike the other entities.

> **Note on biofuel metrics:** `cpbf1` and `cpbf2` are zero across the entire globe for both scenarios in LUH3 v1.1. They are built for structural completeness and will be updated in future dataset versions.

### 1.1 Scenarios and time steps

The cube holds two scenarios across four years each.

| Scenario | Code | IAM | Description | Available Years |
|-------------|-------------|----------------|----------------------|---------------------|
| Low-MAgPIE | `vl` | REMIND-MAgPIE | Very low emissions, land-use change from REMIND-MAgPIE | 2022, 2030, 2050, 2070 |
| H-GCAM | `h` | GCAM | High emissions, land-use change from GCAM | 2022, 2030, 2050, 2070 |

### 1.2 Before you start

Make sure the following are in place:

-   Script 01 has been run and the GeoTIFFs are available in `raw/temp/`
-   The dataset's `metadata.json` file (Obtain from the [EBV Data Portal](https://portal.geobon.org/datasets))
-   The `ebvcube` and `terra` packages are installed

## 2. Load libraries

Here we load the packages used throughout the workflow: `ebvcube` for creating and filling the cube, and `terra` for reading the reference raster.

### 2.1 Clean environment and load libraries

If required, install the missing packages before running the workflow.

```{r}
# Clean environment
rm(list = ls())
gc()

# install.packages(c("ebvcube", "terra"))

library(ebvcube)
library(terra)
```

## 3. Configure temporary raster storage

Reading the input rasters can create temporary files, so point `terra` at a drive with enough free space.

```{r}
local_tmp <- "D:/temp" # Define local disk
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  memfrac = 0.3,
  tempdir = local_tmp 
)

cat("Local temporary folder: ", local_tmp)
```

## 4. Define input and output paths

Input: the pre-extracted GeoTIFFs from script 01 and the JSON metadata file. Output: the EBV Cube NetCDF file. Paths are relative to this script's location.

```{r}
# Directories
raw_dir <- "../raw"               # JSON metadata file and source NetCDF files
pathin  <- "../raw/temp"          # Pre-extracted GeoTIFFs (output of script 01)
cube    <- "../BES-SIM 2/ScenMIP7-LUH3/cube"              # Output folder for the EBV Cube NetCDF file
dir.create(cube, recursive = TRUE, showWarnings = FALSE)


cat("Input folder (GeoTIFFs):", pathin, "\n")
cat("Output folder          :", cube,   "\n")
```

## 5. Define the main workflow settings

Set the scenarios, time steps, metrics, the cube's technical parameters, and the 6 entities. The entities are the crop functional types from the LUH3 v1.1 management dataset.

```{r}
# Scenario codes (order must match the metadata JSON: 1 VL, 2 H)
scenarios <- c("vl", "h")

# Target years (time steps)
years <- c("2022", "2030", "2050", "2070")

# Metrics (order must match the metadata JSON). The metric code is used to build
# the GeoTIFF filenames: fertl (fertilization), irrig (irrigation),
# cpbf1 / cpbf2 (1st / 2nd generation biofuel fraction).
metrics <- c("fertl", "irrig", "cpbf1", "cpbf2")

# EBV Cube technical parameters
epsg <- 4326       # WGS84
prec <- "float"    # Data precision
fv   <- -3.4e+38   # Fill value for missing data

# Entities = crop functional types (order must match the metadata JSON).
# flood applies only to the irrigation metric (see Section 8).
entities <- c("c3ann", "c3nfx", "c3per", "c4ann", "c4per", "flood")

cat("Scenarios:", paste(scenarios, collapse = ", "), "\n")
cat("Years    :", paste(years, collapse = ", "), "\n")
cat("Metrics  :", paste(metrics, collapse = ", "), "\n")
cat("EPSG     :", epsg, "\n")
cat("Fill val :", fv, "\n")
cat("Entities (", length(entities), "):", paste(entities, collapse = ", "), "\n")
```

## 6. Define the spatial template

Read one reference GeoTIFF to extract the spatial extent, resolution, and CRS used for the cube.

```{r}
template    <- rast(file.path(pathin, "management_vl_irrig_c3ann_2022.tif"))
cube_extent <- as.vector(ext(template))
cube_res    <- terra::res(template)

cat("Extent     :", cube_extent, "\n")
cat("Resolution :", cube_res, "\n")
cat("CRS        :", crs(template, describe = TRUE)$code, "\n")
```

## 7. Create EBV Cube NetCDF file

### 7.1 Check paths and available rasters

Define the metadata JSON and the target NetCDF file, and confirm the GeoTIFFs from script 01 are present in the expected folder. This check catches missing inputs before the cube is built.

```{r}
# Define date and check id
id            <- 107
creation_date <- "20260330"
cat("Created date:", format(as.Date(creation_date, format = "%Y%m%d")), "\n")
cat("ID:", id, "\n")

# Metadata JSON
js <- file.path(raw_dir, "metadata.json")

# Note:
# Please follow the naming convention
# "<first_author>_<ebv_class>_id<ID>_<YYYYMMDD>.nc"
nc <- file.path(cube, paste0("Popp&Chini_ecoser_id", id, "_", creation_date, ".nc"))

cat("JSON metadata exists      :", file.exists(js), "\n")
cat("EBV Cube NetCDF file name :", basename(nc),    "\n")
cat("Input GeoTIFF folder      :", pathin,          "\n")

# List all input GeoTIFFs
tif_files <- list.files(pathin, pattern = "\\.tif$", recursive = TRUE, full.names = TRUE)

cat("Total input GeoTIFF files found:", length(tif_files), "\n")
```

### 7.2 Create empty NetCDF file

Create the empty NetCDF file from the JSON metadata. It holds the cube structure (spatial extent, CRS, entities, metrics, time steps, scenarios) but no raster data yet — that comes in Section 8.

```{r}
ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = cube_extent,
  resolution = cube_res,
  timesteps  = years,
  fillvalue  = fv,
  prec       = prec,
  overwrite  = TRUE,
  verbose    = FALSE
)

cat("=== Cube structure ===\n")
print(ebv_datacubepaths(nc, verbose = FALSE))
```

## 8. Populate the EBV Cube

Insert every entity GeoTIFF at its correct scenario / metric / time step / entity position, looping over scenarios, metrics, years, and entities. The `flood` entity applies only to the irrigation metric and uses its own filename; for the other metrics it is skipped and remains at the cube's fill value.

```{r}
inserted <- 0L

start.time <- Sys.time()
for (s_no in seq_along(scenarios)) {

  code <- scenarios[s_no]

  for (m_no in seq_along(metrics)) {

    metric <- metrics[m_no]

    for (t_no in seq_along(years)) {

      yr <- years[t_no]

      for (e_no in seq_along(entities)) {

        ent <- entities[e_no]

        # flood: only meaningful for irrigation, and named without a metric prefix
        if (ent == "flood") {
          if (metric != "irrig") next   # leave at the cube's fill value
          in_file <- file.path(pathin, sprintf("management_%s_flood_%s.tif", code, yr))
        } else {
          in_file <- file.path(pathin,
                               sprintf("management_%s_%s_%s_%s.tif", code, metric, ent, yr))
        }

        if (!file.exists(in_file)) {
          warning("Missing file, skipped: ", in_file)
          next
        }

        ebv_add_data(
          filepath_nc = nc,
          data        = in_file,
          band        = 1,
          scenario    = s_no,
          metric      = m_no,
          entity      = e_no,
          timestep    = t_no,
          ignore_RAM  = FALSE,
          raw         = FALSE,
          verbose     = FALSE
        )
        inserted <- inserted + 1L
      }
    }
    cat("  ✔ inserted", code, "| metric", metric, "\n")
  }
}

end.time <- Sys.time()
cat("\nDone. Time:", round(difftime(end.time, start.time, units = "mins"), 2), "min\n")
```

## 9. Validation and visualization

A quick check that the cube is readable and that the scenarios, metrics, entities, and time steps line up.

```{r}
ebv_analyse(nc, entity = 1, timestep = 1, scenario = 1, metric = 1)

props <- ebv_properties(nc, scenario = 1, metric = 1)
props@general$entity_names

# Define data cube paths
datacubes <- ebv_datacubepaths(nc, verbose = FALSE)

# Simple plot
ebv_map(filepath = nc, datacubepath = datacubes[1, 1], entity = 1,
        timestep = 1, verbose = FALSE)
```

## 10. Summary

```{r}
# flood contributes only via the irrigation metric; the other entities apply to all metrics
n_nonflood <- length(entities) - 1L
expected   <- length(scenarios) * length(years) * (n_nonflood * length(metrics) + 1L)

cat("==============================\n")
cat("Cube     :", basename(nc), "\n")
cat("Inserted :", inserted, "/ expected:", expected, "\n")
cat("==============================\n")
```

*Dataset: LUH3 V1.1 — UofMD · Hurtt et al. 2020 · CMIP7 ScenarioMIP · Low-MAgPIE (VL) and H-GCAM (H) · 2022–2100*
