---
title: 'PLUM (ID 99) — EBV Cube: Presence/Absence as byte binary'
<<<<<<< HEAD
output: html_document
---

## Overview
=======
output:
  html_document: default
---

```{r }
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE)
```

## Objective 
>>>>>>> fcecd523726a6e9f9eb560227f4470acd1e7078b

this notebook creates an **EBV Cube** (netCDF) for the **PLUM (ID 99)** dataset from multiband GeoTIFFs, where:

 
- Each folder under `S:\\ebv_portal\\99\\data\\` represents a **scenario**.
- Each GeoTIFF is a **multiband**, where each band corresponds to a year.
- Each `land_cover` class is transformed into a binary raster:
  - `1` = presence
  - `0` = absence
  - `NA` is preserved

**Note:** In netCDF/CF, the binary is represented as a byte (0/1).

## Load libraries and define paths
Load the necessary libraries and define the paths. The JSON file with the metadata should have been created beforehand. Name the file according to the naming convention. Save the output file
in the S: drive. 

```{r}
rm(list = ls())
gc()

library(ebvcube)
library(terra)
library(stringr)
library(dplyr)

id <- 99
date <- "20260113"

root <- paste0("I:\\biocon\\ebv_portal\\", id)
js   <- file.path(root, "metadata.json")

nc <- paste0("S:\\ebv_portal\\", id, "\\alexander_ecostr_id", id, "_", date, ".nc")

tif_root  <- file.path("S:\\ebv_portal", id, "data")
tif_files <- list.files(tif_root, pattern="\\.tif$", recursive=TRUE, full.names=TRUE)
stopifnot(length(tif_files) > 0)

cat(paste0("Path: ", tif_files, "\n"), "\n")

```


## Scenarios and years 
Define scenarios and years available in the data.This will be used to map the 
data correctly into the netCDF. 

```{r}
p_norm <- gsub("\\\\", "/", tif_files)
scs <- basename(dirname(p_norm))
scs <- gsub("_.*", "", scs)
scenario_levels <- c("NfN", "NfNl", "NaC", "NfS", "BAU")

# Match scenarios found in the data with the expected ones
stopifnot(setequal(scenario_levels, unique(scs)))


band_years_available <- c(2021, 2030, 2050, 2070, 2100)
timesteps_int <- band_years_available
timesteps     <- as.character(timesteps_int)

band_idx <- match(timesteps_int, band_years_available)
stopifnot(!any(is.na(band_idx)))
cat("Scenarios:", paste(scenario_levels, collapse=", "), "\n")
cat("Years:", paste(timesteps, collapse=", "), "\n")

```

## Template and classes/entitites
Read the first GeoTIFF to extract spatial properties and land cover classes.
We define class_id and class_names, and not neded classes (ocean, no data) will 
be removed in the next step.


```{r}
r0 <- rast(tif_files[1])

extent <- as.vector(ext(r0))
res_xy  <- res(r0)
epsg    <- 4326

lev <- terra::levels(r0[[1]])
stopifnot(length(lev) >= 1)

entities_values_df <- as.data.frame(lev[[1]])
class_id    <- as.integer(entities_values_df$id)
class_names <- as.character(entities_values_df$land_cover)

entities_full <- class_names
length(class_id)
cat( paste0(class_id, ": ", class_names, collapse="\n"), "\n")
```

## Remove and define the final entities 
Delete entities that are not necessary. In this case, the following should be removed:
- 0: Ocean
- 99: No data

```{r}
exclude_ids   <- c(0, 99)
exclude_names <- c("Ocean", "No data")

keep <- !(entities_values_df$id %in% exclude_ids |
          entities_values_df$land_cover %in% exclude_names)

entities_filtered_df <- entities_values_df[keep, ]

```

Define final entities

```{r}
class_id    <- as.integer(entities_filtered_df$id)
class_names <- as.character(entities_filtered_df$land_cover)

entities <- class_names
cat( paste0(class_id, ": ", class_names, collapse="\n"), "\n")
```

## Create an empty EBV netCDF file
The fill values and the precision (data type) for the NetCDF file should be defined.
In this case, use bytes, as this is the most efficient way to save real binary cubes. 
One byte is used per pixel, as opposed to double, which uses eight bytes per pixel, or integer, which uses four bytes.

```{r}
fillvalue <- -1
prec <- "byte"
metric_no <- 1

ebv_create(
  jsonpath   = js,
  outputpath = nc,
  entities   = entities,
  epsg       = epsg,
  extent     = extent,
  resolution = res_xy,
  timesteps  = timesteps,
  fillvalue  = fillvalue,
  prec       = prec,
  force_4D   = TRUE,
  overwrite  = TRUE,
  verbose    = TRUE
)

cat("EBV cube created at:", nc, "\n")

```

<<<<<<< HEAD
=======
---
>>>>>>> fcecd523726a6e9f9eb560227f4470acd1e7078b
## Insert data into the netCDF (precense/absence)
Create an temporary folder in locla disk to optimized time and write temporary files.
The inserction data to the file, should be by order - loop through scenarios, timesteps, entities.

```{r, eval=FALSE}
local_tmp <- "C:/temp/ebvcube_tmp"
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  tempdir = local_tmp,
  progress = 1
)


# loop through each scenario
for (s_no in seq_along(scenario_levels)) {
  
  sc <- scenario_levels[s_no]
  idx_sc <- which(scs == sc)

  if (length(idx_sc) != 1) {
    warning("Scenario ", sc, ": expected exactly 1 file, got ", length(idx_sc), " -> skipping")
    next
  }

  r <- rast(tif_files[idx_sc])
  
  # Loop through each timestep
  for (t_no in seq_along(timesteps_int)) {

    b  <- band_idx[t_no]
    yr <- timesteps_int[t_no]
    cat("Scenario ", sc, " - Year ", yr, " (band ", b, ")\n")

    r_t <- r[[b]]
    
    # Loop through each entity
    for (e_no in seq_along(class_id)) {
      
      code <- class_id[e_no]
      
      cat("Entity:", entities[e_no], " (code ", code, ")\n")
      
      # Convert to binary to store it as one byte per pixel
      bin_r <- ifel(is.na(r_t), NA, ifel(r_t == code, 1, 0))
      
      # Create a temporary file on the local disk
      tmp_tif <- tempfile(tmpdir = local_tmp, fileext = ".tif")
      
      # Write raster in local disk
      writeRaster(
        bin_r, tmp_tif,
        overwrite = TRUE,
        datatype  = "INT1S", # signed byte
        NAflag    = -1,          
        wopt      = list(gdal = c("COMPRESS=DEFLATE")))

      # Add data to the nc file, using one byte per pixel
      ebv_add_data(
        filepath_nc = nc,
        entity      = e_no,
        timestep    = t_no,
        data        = tmp_tif,
        band        = 1,
        scenario    = s_no,
        metric      = 1,
        ignore_RAM  = TRUE,
        raw         = TRUE,
        verbose     = FALSE
      )
  
      # Remove the temporary file
      rm(bin_r)
    }
    
    # Free up R Memory
    rm(r_t)
    gc()
  }
}

```

## Quick vallidation and visualization

<<<<<<< HEAD
```{r, eval=FALSE}
=======
```{r validate}
>>>>>>> fcecd523726a6e9f9eb560227f4470acd1e7078b
ebv_analyse(nc, entity=1, timestep=1, scenario=1, metric=1)
props <- ebv_properties(nc, scenario=1, metric=1)
props@ebv_cube$type
# H5T_STD_I8LE 	8-bit little-endian signed integer (2's complement)
props@general$entity_names

datacubes <- ebv_datacubepaths(nc, verbose=FALSE)

ebv_map(filepath = nc, datacubepath = datacubes[1,1], entity = 1,
        timestep = 1, classes = 2, verbose = FALSE)

```
