---
title: 'PLUM LU rasterization — R Notebook '
output:
  html_document:
    df_print: paged
  html_notebook: default
---

```{r setup}
knitr::opts_chunk$set(
  echo = TRUE,
  message = TRUE,
  warning = TRUE
)
```

## Context and goal

This notebook converts **PLUM LU** input data distributed as **.zip** archives (containing compressed `.txt.gz` tables) into **global GeoTIFF rasters** at **0.5° resolution** in **WGS84**.

The workflow produces one raster per combination of:

- scenario (top-level folder inside each zip)
- level (e.g. `s1`)
- year (e.g. `2020`, `2030`, ...)
- variable (e.g. `LandUse`, `Forestry`)
- and, when present, per `Type` and `Crop`

Each output GeoTIFF may contain multiple bands corresponding to value columns such as  
`A, FI, FQ, II, IQ, OI, Y` (only if present in the input table).

---

## Notes on the original script (bugs and pitfalls)

- Output filenames were hard-coded with `LandUse_...` even when processing other variables (e.g. `Forestry`).
- A second raster template was created but never used.
- A fixed `+0.25` offset was used to convert coordinates; this assumes a 0.5° grid and should instead be derived from the raster resolution.
- The script implicitly assumes the presence of `Type`, `Crop`, and value columns; this notebook checks for them explicitly and fails early if assumptions are violated.

---

## Load libraries

```{r libraries}
library(terra)
```

---

## Define input paths and processing parameters
The scenarios, years and variables to process can be adjusted as needed.
Here we focus on the `Forestry` variable for all scenarios and years.

```{r parameters}

# Base directory containing scenario zip files
base_dir <- "S:/Emmanuel_OcegueraConchas/BES-SIM 2/PLUM_LU/0.5_degrees"

# Output directory for GeoTIFFs
out_dir  <- "S:/Emmanuel_OcegueraConchas/BES-SIM 2/PLUM_LU/0.5_degrees_rasters_final"

# Scenario identifiers (must match zip internal paths)
scenarios <- c("BAU", "NaC_NaC", "NfN_NfN", "NfS_NfS", "NfNl_NfNl")

# Years to process
target_years <- c("2020", "2030", "2050", "2070", "2100")

# Scenario level inside each zip
level_4 <- "s1"

# Variables to process
vars_4 <- c("LandUse")

# Expected value columns (only those present will be used)
default_value_cols <- c("A", "FI", "FQ", "II", "IQ", "OI", "Y")
```

---

## Locate scenario zip files

```{r find-zips}
zip_files <- list.files(base_dir, pattern = "\\.zip$", full.names = TRUE)
stopifnot(length(zip_files) > 0)

cat("Zip files found:\n", paste(zip_files, collapse = "\n"), "\n")
```

---

## Create a global raster template

```{r template}
r_template <- rast(
  xmin = -180, xmax = 180,
  ymin = -90,  ymax = 90,
  resolution = 0.5,
  crs = "EPSG:4326"
)

# Half-cell offset for converting lower-left coordinates to cell centers
half_res <- res(r_template) / 2
half_res
```

---

## Main processing workflow
The logic here is the following:
- For each scenario:
  - Locate the corresponding zip file
  - List files inside the zip
  - Identify relevant files based on year and variable patterns
  - For each relevant file:
    - Extract year and variable from the file path
    - Unzip and read the compressed text table
    - Check for presence of `Type`, `Crop`, and value columns
    - Rasterize data, splitting by `Type` and `Crop` if present
    - Write output GeoTIFFs with appropriate filenames

```{r process, warning=TRUE}

local_tmp <- "C:/temp/plum_rasterization"
dir.create(local_tmp, showWarnings = FALSE, recursive = TRUE)

terraOptions(
  todisk  = TRUE,
  tempdir = local_tmp,
  progress = 1
)

start_time <- Sys.time()

# Loop over scenarios
for (scen in scenarios) {

  scen_zip <- grep(scen, zip_files, value = TRUE)

  if (length(scen_zip) == 0L) {
    warning("No zip file found for scenario: ", scen)
    next
  }
  scen_zip <- scen_zip[1]

  cat("\n=============================\n")
  cat("Scenario:", scen, "\nZip:", scen_zip, "\n")

  # List files inside the zip archive
  file_list <- unzip(scen_zip, list = TRUE)$Name

  year_pattern <- paste(target_years, collapse = "|")
  var_pattern  <- paste(vars_4, collapse = "|")

  # construct regex pattern to match relevant files
  pattern <- sprintf(
    "^%s/%s/(%s)/(%s)\\.txt\\.gz$",
    scen,
    level_4,
    year_pattern,
    var_pattern
  )

  # Find matching files
  candidate_files <- grep(pattern, file_list, value = TRUE)
  
  if (length(candidate_files) == 0L) {
    warning("No matching files found for scenario: ", scen)
    next
  }
  
  cat("Candidate files:\n", paste(candidate_files, collapse = "\n"), "\n")

  # Process each candidate file 
  for (rel_path in candidate_files) {

    # Extract year and variable from the file path
    year <- sub(sprintf("^%s/%s/(\\d{4})/.*$", scen, level_4), "\\1", rel_path)
    var  <- sub("^.*/(.*)\\.txt\\.gz$", "\\1", rel_path)
    
    tmp_dir <- file.path(tempdir(), scen, level_4, year)
    dir.create(tmp_dir, recursive = TRUE, showWarnings = FALSE)
    unzip(scen_zip, files = rel_path, exdir = tmp_dir)
    
    txt_gz <- file.path(tmp_dir, rel_path)

    con <- gzfile(txt_gz, "rt")
    dat <- read.table(con, header = TRUE)
    close(con)


    if (!all(c("Lon", "Lat") %in% names(dat))) {
      stop("Missing Lon/Lat columns in file: ", rel_path)
    }
    
    value_cols <- intersect(default_value_cols, names(dat))
    if (length(value_cols) == 0L) {
      stop("No expected value columns found in file: ", rel_path)
    }
    
    has_type_crop <- all(c("Type", "Crop") %in% names(dat))

    # Rasterization logic depends on presence of Type and Crop
    if (has_type_crop) {

      # Split by Type 
      dat_by_type <- split(dat, dat$Type)

      # Iterate over Types
      for (type in names(dat_by_type)) {
        
        # Split by Crop within each Type
        dat_by_crop <- split(dat_by_type[[type]], dat_by_type[[type]]$Crop)

        # Iterate over Crops
        for (crop in names(dat_by_crop)) {
          
          # Process each crop subset
          subdat <- dat_by_crop[[crop]]

          # Compute cell centers
          xy_centers <- cbind(
            x = subdat$Lon + half_res[1],
            y = subdat$Lat + half_res[2]
          )

          # Get cell indices for these centers
          cells <- cellFromXY(r_template, xy_centers)

          # Initialize raster stack
          r_stack <- NULL

          # Create raster bands for each value column
          for (vcol in value_cols) {
            r_band <- r_template
            values(r_band) <- NA
            r_band[cells] <- subdat[[vcol]]
            names(r_band) <- vcol
            r_stack <- if (is.null(r_stack)) r_band else c(r_stack, r_band)
          }

          # Construct output file path
          out_file <- file.path(
            out_dir, scen, level_4,
            sprintf("%s_%s_%s_%s_%s.tif",
                    var, scen, year,
                    gsub("[^A-Za-z0-9]+", "_", type),
                    gsub("[^A-Za-z0-9]+", "_", crop))
          )

          # Ensure output directory exists
          dir.create(dirname(out_file), recursive = TRUE, showWarnings = FALSE)
          writeRaster(r_stack, out_file, overwrite = TRUE)
        }
      }

    } else {

      # No Type/Crop splitting; rasterize entire dataset
      xy_centers <- cbind(
        x = dat$Lon + half_res[1],
        y = dat$Lat + half_res[2]
      )

      cells <- cellFromXY(r_template, xy_centers)

      r_stack <- NULL

      # Create raster bands for each value column
      for (vcol in value_cols) {
        r_band <- r_template
        values(r_band) <- NA
        r_band[cells] <- dat[[vcol]]
        names(r_band) <- vcol
        r_stack <- if (is.null(r_stack)) r_band else c(r_stack, r_band)
      }

      out_file <- file.path(
        out_dir, scen, level_4,
        sprintf("%s_%s_%s.tif", var, scen, year)
      )

      dir.create(dirname(out_file), recursive = TRUE, showWarnings = FALSE)
      writeRaster(r_stack, out_file, overwrite = TRUE)
    }

    unlink(txt_gz)
  }
}

end_time <- Sys.time()
print(end_time - start_time)
```
